### 2017-11-22 MS-cclifeexpec.Rmd
maintext

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: C:/Users/Matt/Documents/Matt - Work Stuff/r code/delete/svm-latex-ms.tex
title: "Climate change projected to reduce European life expectancy"
thanks: "All data and code that supports these conclusions are available upon request at hauer@uga.edu."
author:
- name: Mathew E. Hauer
  affiliation: University of Georgia
- name: Alexis R. Santos
  affiliation: Pennsylvania State University

abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change. However, life expectancy at birth is more easily communicated to the public, understood by the public, and controls for different geographic, economic, and demographic groups. No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy. Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting cost of climate change on mortality. In this paper, we convert excess mortality estimates into potential reductions of life expectancy at birth and report the cost of climate change on human longevity. We find climate change could become the third largest life expectancy reducer behind heart disease and cancer in some European countries. Thus, the cost of inaction on climate change could approach one year of life in some countries."

keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography: C:/Users/Matt/Documents/Untitled/ccmort.bib
biblio-style: nature
fig_caption: true
# use apsr or nature
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(png)
rm(list=ls())
library(tidyverse)
library(RColorBrewer)
library(scales)
library(tmap)
library(tmaptools)

lt_lancet <- read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/lt_afterr_09032017.csv")
lt_lifetables <- read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/lancet_lifetables_09032017.csv")

lt_lancet$'country name' <- lt_lancet$COUNTRY
lt_lancet<- lt_lancet[order(lt_lancet$DIF_MID), ]
lt_lancet$'country name' <- factor(lt_lancet$'country name', levels = lt_lancet$'country name')
```

# Introduction

Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimatin], estimates of displacement [@nicholls2011sea;@hauer2016millions;@hauer2017migration], or human conflict [@barnett2007climate]  but have the potential to lead to the loss of human life [@forzieri2017increasing;@pachauri2014climate]. As impact quantification studies move further from the physical sciences of climate change into the social sciences, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].

Scholars have long tried to estimate the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer]. Although such estimates are useful, the excess mortality estimates are rather sterile - one death is a tragedy, a million deaths a statistic - and difficult to relate to on an individual level. Life expectancy at birth, on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone. To date, no studies exist that comprehensively examine the potential reductions of life expectancy due to climate change. Without properly situating the potential loss of life within the contexts of metrics such as life expectancy, we risk underestimating the impact of climate change on human mortality.

Using published data on excess mortality, we connect climate change excess mortality to life expectancy in a mortality model. By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in twenty-eight European countries. This approach allows us to quantify the impact of climate change on human longevity and answer the following question: How much could climate change mortality reduce life expectancy at birth? Our results can situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.

# Methods
We estimate changes in life expectancy by combining three primary datasets: excess mortality data from Forzieri et al. [@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD]. Forzieri et al. [@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario. They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases. Their results represent business-as-usual climate change and human development without incorporating potential adaptation. They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes. These data provide the anticipated fatalities but not the effect these deaths might have on life expectancy.

To convert excess mortality to life expectancy, we use data gathered from the HMD [@HMD] for corresponding European countries in five-year life tables in conjunction with cause-specific mortality data from the GBD [@GBD;@wang2012age]. The HMD comprises only complete, official vital event statistics and represents the most accurate and complete compilation of mortality data available. We allocate excess multi-hazard mortality for each country from supplementary table S8 [@forzieri2017increasing] based on the observed mortality schedule for environmental heat and cold exposure deaths from the GBD for each country in the previous decade.

We then recalculated age-specific mortality rates in five-year life tables by adding the climate change mortality onto the underlying mortality data and create four subsequent life tables (baseline, low, mid, high) using standard life table techniques [@wunsch2013life] based on the published confidence intervals from [@forzieri2017increasing]. We used a variant of the cause-deleted life tables approach [@brand2005approximations;@beltran2008integrated] to measure the impact of excess mortality on life expectancy at birth or $e_0$. Changes in life expectancy are reported based on differentials from $e_0$ between the 2080s and the most recent complete life table in the HMD. Complete life tables for all countries are available in the Supplementary Dataset and all equations are located in the Supplementary Information.

# Results
We find that climate change could alter life expectancy by -0.22 years (-2.5 months) in the average European country (**Fig. 1**). This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States. Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.

```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change. We report changes in life expectancies due to climate change for twenty-eight European countries. The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}

ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=DIF_MID)) +
  geom_pointrange(stat="identity") +
  geom_point(stat="identity", fill="black", size=8) +
   geom_hline(yintercept=0) +
  geom_text(color="white", size=2) +
  #scale_x_discrete(limits = rev(levels(the_factor)))
  scale_x_reverse(breaks = lt_lancet$RANK, labels=lt_lancet$COUNTRY) +
  coord_flip() +
  theme_bw() +
  #ylim(-2, 2) +
  theme(panel.grid.minor = element_blank()) +

   labs(y=expression(Delta* e[0]),
        x= "Country")
```

Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies. Whereas many Northern European countries could experience negligible impacts on life expectancy, five European countries could see life expectancy changes in excess of -0.5 years (Spain, Luxembourg, France, Italy, and Ireland). This group of countries could see life expectancy changes of more than -1.0 years if climate change hazards are more intense than anticipated. In some countries, climate change could thus become a bigger killer than trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or all accident related mortality (-0.84 years) [@arias2013united] by the end of the century.

```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}

data(Europe)
Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)

bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)

  tm_shape(Europe2) +
  tm_fill("DIF_MID", title = expression(Delta* e[0 ~italic(MID)]), palette = "YlOrBr", style = "jenks") +
  tm_borders(alpha = 0.5) +  
  tm_text("Test" , size="AREA", root=5) +
  tm_format_Europe() 
```

While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig. 2**). The most westerly European countries tend to have the greatest reductions. Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change. Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.

# Discussion
In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries. Previous studies on climate change and excess mortality potentially misrepresent the impact climate change could have on human mortality. Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy]. Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer]. Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes. Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].

Without adaptation measures, our results suggest climate change could emerge as a significant new mortality vector and could pose a major public health threat for some European countries by the end of the century, echoing previous findings [@forzieri2017increasing;@patz2005impact]. Life expectancy has steadily risen across the world for the last century [@oeppen2002broken] and our results suggest that climate change alone could spur a sharp reversal in these trends. We expect one European countries to see life expectancy reductions more than one year under the middle scenario (Luxembourg), but if climate change has a greater impact on mortality than anticipated, five European countries (Luxembourg, Spain, Italy, France, Ireland) could see life expectancy reductions more than one year with Luxembourg experiencing a reduction of more than two.

Reductions such as those should not be taken lightly. Many of the children born today are likely to still be alive by the end of the century and will be in the age groups most threatened by extreme heat - the biggest mortality risk associated with climate change [@keatinge2000heat]. If climate change unfolds as a more aggressive mortality vector, only major cardiovascular diseases and malignant neoplasms would contribute more to life expectancy reductions in numerous European countries [@arias2013united]. This would make climate change one of the most aggressive new mortality vectors to emerge over the last quarter-century, representing a major threat to public health in many parts of Europe.

Prospective studies on the emerging threat from climate change rely on linking contemporary mortality with future mortality. However, climate change could reshape future mortality through other causes of death.  Climate change affects health behaviors that in turn increase mortality risk through increased alcohol and substance abuse, violent behavior, insecurity, increase in post-traumatic stress due to weather-related trauma, increase in stress due to climate change and schizophrenia, increase in the use of medications that reduce the ability to perspire and sweat, etc. [@patz2005impact] The International Classifications of Diseases and Related Health Problems (ICD) does not contain ``climate change" as an official cause of death, so we can only speculate that the impact of climate change could be larger than reported here. Although we do not model these potential impacts, our results could thus be considered conservative.

We also share the concerns of Lee et al. [@lee2017comprehensive] concerning the business-as-usual climatic assumptions. It is likely that many countries and communities will deploy a wide variety of adaptation measures [@haines2006climate;@kovats2003methods;@ebi2006approach]. These adaptation measures rely on accurate information about the potential mortality vectors. Our models and those produced by Forzieri et al. [@forzieri2017increasing] present plausible scenarios on the potential impact of climate change on human mortality and provide crucial information to public health officials, national governments, and international organizations. The time frames associated with climate change allow ample time for this potential health crisis to be averted.

Finally, we would like to point out that future research should not only transform excess mortality into life expectancy decrements. Given the influence of climate change in diseases and causes of death, it is imperative to quantify the extent to which climate change will derive in increasing costs for health care systems in these countries. The health care structures are being taxed by population aging [@rechel2009can] and rising health care costs, yet it remains unclear how climate change will exacerbate these pressures.

\newpage

# Supplemenatary Information
## Data
To create life expectancy differentials, we used three sources of data. The first is the baseline life table data. Baseline life tables comes from the Human Mortality Database (HMD) [@HMD]. These datasets comprise only official vital event registration and are considered the most complete and accurate source of mortality data available. We downloaded the most recent 1x5 (by age and year) life table data for twenty-eight European countries. We used the following data elements from the five-year life table data in the HMD:

\begin{enumerate}
\item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
\item $_na_x$ or the average length of survival between ages $x$ and $x+n$.
\item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
\end{enumerate}

The second dataset comes from published projections of climate-related mortality in Europe [@forzieri2017increasing]. Forzieri et al. [@forzieri2017increasing] combined data from the International Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool (NatCatSERVICE) to create exposure and fatality statistics related to six weather-related hazards -- heatwaves, cold waves, droughts, wildfires, river and coastal floods, and windstorms. They downscaled these exposures to  1km grid scales and integrated them into small-area demographic projections for Europe out to 2100. Heatwaves and coldwaves account for the vast majority of projected fatalities. The number of future anticipated deaths are available in their Supplementary Table S8 [@forzieri2017increasing] and provide the magnitude of deaths for our study.

The third dataset comes from the GBD [@GBD;@wang2012age]. The GBD is the most comprehensive worldwide dataset of epidemiological data produced and provides cause-specific mortality by age/sex/geography/year for 249 causes of death in 195 countries and territories. We gathered data on age-specific deaths and mortality rates ($m_x$) for mortality from environmental heat and cold exposure (cause C.2.9) for the period 2006-2015 for each country in the study. These data provide a mortality schedule to fit the projected climate change excess mortality from Fozieri et al [@forzieri2017increasing].

## Methods
We derived a number of additional variables to accomplish our analysis. First, we abridged the life table data from the HMD from 100+ to 80+ to conform to the GBD cause-specific mortality schedules.

From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}$).

We then derived additional $m_x$ rates for each age group $x$ in each country $i$ for each scenario $s$ ($BASE$, $LOW$, $MID$, $HIGH$) from Forzieri et al [@forzieri2017increasing].

$$ _nm_{x,i,BASE} = _nD_{x,i}^{HMD} / _nP_{x,i}^{HMD} $$
$$ _n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + \hat{D_i} \cdot _nt_{x,i} ) / _nP_{x,i}^{HMD} $$

where $D_{x,i}$ is the number of deaths in age group $x$ in country $i$ from the HMD, $P_{x,i}$ is the population in age group $x$ from the HMD, $\hat{D_x}$ is the number of deaths from Forzieri et al. [@forzieri2017increasing], and $t_{x,i}$ is the proportion of mortality experienced in each age group $x$ from the GBD. Thus, the anticipated additional mortality due to climate change is added to each age group based on the underlying mortality cause-specific mortality schedule observed between 2006-2015 in the GBD.

We then calculated $q_x$ values, or the probability of dying, for each scenario $s$ for each country.

$$_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}$$
$$_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}$$

We calculated each additional life table value identically for each scenario $s$ using the standard life table equations:

$$_nd_{x,i,s} = _nl_{x,i,s} \cdot _nq_{x,i,s}$$
$$_nl_{x,i,s} = _nl_{x-1,i,s} - _nd_{x-1,i,s}$$
$$_nL_{x,i,s} = _na_{x,i,s} \cdot _nl_{x,i,s} + ((n-_na_{x,i,s}) \cdot _nl_{x,i,s})$$
$$_nT_{x,i,s} = _nL_{x,i,s} + _nT_{x+1,i,s}$$
$$_ne_{x,i,s} = _nT_{x,i,s} / _nl_{x,i,s}$$

To determine the differences between life expectancy compared to the baseline, we simply subtract $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].

### 2017-11-22 MS-cclifeexpec.tex
maintext

\documentclass[12pt,]{article}
\usepackage[left=1in,top=1in,right=1in,bottom=1in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage{lmodern}

  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}

\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center

\renewenvironment{abstract}
 {{%
    \setlength{\leftmargin}{0mm}
    \setlength{\rightmargin}{\leftmargin}%
  }%
  \relax}
 {\endlist}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother

\setcounter{secnumdepth}{0}

\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}

\title{Climate change projected to reduce European life expectancy \thanks{All data and code that supports these conclusions are available upon
request at \href{mailto:hauer@uga.edu}{\nolinkurl{hauer@uga.edu}}.}  }

\author{\Large Mathew E. Hauer\vspace{0.05in} \newline\normalsize\emph{University of Georgia}   \and \Large Alexis R. Santos\vspace{0.05in} \newline\normalsize\emph{Pennsylvania State University}  }

\date{}

\usepackage{titlesec}

\titleformat*{\section}{\normalsize\bfseries}
\titleformat*{\subsection}{\normalsize\itshape}
\titleformat*{\subsubsection}{\normalsize\itshape}
\titleformat*{\paragraph}{\normalsize\itshape}
\titleformat*{\subparagraph}{\normalsize\itshape}

\usepackage{natbib}
\bibliographystyle{nature}
\usepackage[strings]{underscore} % protect underscores in most circumstances

\newtheorem{hypothesis}{Hypothesis}
\usepackage{setspace}

\makeatletter
\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \PassOptionsToPackage{hyphens}{url}\usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \PassOptionsToPackage{hyphens}{url}\usepackage[unicode=true]{hyperref}
\fi
}

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Mathew E. Hauer (University of Georgia) and Alexis R. Santos (Pennsylvania State University)},
             pdfkeywords = {Climate change, Life expectancy, Mortality, Demography, Excess
mortality, Europe, Public health},  
            pdftitle={Climate change projected to reduce European life expectancy},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\usepackage[all]{nowidow}
\usepackage{rotating}

% add tightlist ----------
\providecommand{\tightlist}{%
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

% \pagenumbering{arabic}% resets `page` counter to 1 
%
% \maketitle

{% \usefont{T1}{pnc}{m}{n}
\setlength{\parindent}{0pt}
\thispagestyle{plain}
{\fontsize{18}{20}\selectfont\raggedright 
\maketitle  % title \par

}

{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
\textbf{\authorfont Mathew E. Hauer} \hskip 15pt \emph{\small University of Georgia}   \par \textbf{\authorfont Alexis R. Santos} \hskip 15pt \emph{\small Pennsylvania State University}

}

}

\begin{abstract}

    \hbox{\vrule height .2pt width 39.14pc}

    \vskip 8.5pt % \small

\noindent Climate change related estimates of excess mortality clearly demonstrate
the dramatic impact on public health and human mortality from climate
change. However, life expectancy at birth is more easily communicated to
the public, understood by the public, and controls for different
geographic, economic, and demographic groups. No studies have
comprehensively connected excess mortality due to climate change to
anticipated reductions in life expectancy. Without properly situating
the potential loss of life within the contexts of life expectancy, we
risk misrepresenting cost of climate change on mortality. In this paper,
we convert excess mortality estimates into potential reductions of life
expectancy at birth and report the cost of climate change on human
longevity. We find climate change could become the third largest life
expectancy reducer behind heart disease and cancer in some European
countries. Thus, the cost of inaction on climate change could approach
one year of life in some countries.

\vskip 8.5pt \noindent \emph{Keywords}: Climate change, Life expectancy, Mortality, Demography, Excess
mortality, Europe, Public health \par

    \hbox{\vrule height .2pt width 39.14pc}

\end{abstract}

\vskip 6.5pt

\noindent \doublespacing \section{Introduction}\label{introduction}

Climate change's implications on humanity go far beyond estimates of
economic damages \citep{hsiang2017estimatin}, estimates of displacement
\citep{nicholls2011sea, hauer2016millions, hauer2017migration}, or human
conflict \citep{barnett2007climate} but have the potential to lead to
the loss of human life
\citep{forzieri2017increasing, pachauri2014climate}. As impact
quantification studies move further from the physical sciences of
climate change into the social sciences, properly quantifying and
conveying the impact of climate change on public health is of increasing
importance \citep{melillo2014climate, cloyd2016engagement}.

Scholars have long tried to estimate the mortality risks associated with
climate change and typically use excess or extra mortality
\citep{forzieri2017increasing, wilson2017climate, mcmichael2006climate, zanobetti2012summer}.
Although such estimates are useful, the excess mortality estimates are
rather sterile - one death is a tragedy, a million deaths a statistic -
and difficult to relate to on an individual level. Life expectancy at
birth, on the other hand, provides potent comparisons of mortality
vectors and converts excess mortality into an intuitively understood
metric, relatable to everyone. To date, no studies exist that
comprehensively examine the potential reductions of life expectancy due
to climate change. Without properly situating the potential loss of life
within the contexts of metrics such as life expectancy, we risk
underestimating the impact of climate change on human mortality.

Using published data on excess mortality, we connect climate change
excess mortality to life expectancy in a mortality model. By estimating
the increase in age-specific mortality rates associated with previously
published excess death estimates, we can assess how much climate change
could reduce the anticipated longevity of the average person in
twenty-eight European countries. This approach allows us to quantify the
impact of climate change on human longevity and answer the following
question: How much could climate change mortality reduce life expectancy
at birth? Our results can situate climate change mortality within the
broader context of human mortality and can be used to inform public
health interventions to prevent such futures.

\section{Methods}\label{methods}

We estimate changes in life expectancy by combining three primary
datasets: excess mortality data from Forzieri et al.
\citep{forzieri2017increasing}, cause-of-death distributions from the
Global Burden of Disease study (GBD) \citep{GBD, wang2012age}, and life
tables from the Human Mortality Database (HMD) \citep{HMD}. Forzieri et
al. \citep{forzieri2017increasing} focused on the projected excess
mortality of heatwaves, coldwaves, wildfires, droughts, river and
coastal floods, and windstorms under the business-as-usual greenhouse
gas emission scenario. They modelled long-term, small-area demographic
changes in European countries that correspond to a middle-of-the-road
socioeconomic scenario and based their projected mortality on
extrapolations of an exhaustive examination of contemporary disaster
databases. Their results represent business-as-usual climate change and
human development without incorporating potential adaptation. They found
a potential 150,000+ climate change related fatalities per year by the
mid 2080s with climate change contributing to 90\% of mortality rise as
opposed to population changes. These data provide the anticipated
fatalities but not the effect these deaths might have on life
expectancy.

To convert excess mortality to life expectancy, we use data gathered
from the HMD \citep{HMD} for corresponding European countries in
five-year life tables in conjunction with cause-specific mortality data
from the GBD \citep{GBD, wang2012age}. The HMD comprises only complete,
official vital event statistics and represents the most accurate and
complete compilation of mortality data available. We allocate excess
multi-hazard mortality for each country from supplementary table S8
\citep{forzieri2017increasing} based on the observed mortality schedule
for environmental heat and cold exposure deaths from the GBD for each
country in the previous decade.

We then recalculated age-specific mortality rates in five-year life
tables by adding the climate change mortality onto the underlying
mortality data and create four subsequent life tables (baseline, low,
mid, high) using standard life table techniques \citep{wunsch2013life}
based on the published confidence intervals from
\citep{forzieri2017increasing}. We used a variant of the cause-deleted
life tables approach
\citep{brand2005approximations, beltran2008integrated} to measure the
impact of excess mortality on life expectancy at birth or \(e_0\).
Changes in life expectancy are reported based on differentials from
\(e_0\) between the 2080s and the most recent complete life table in the
HMD. Complete life tables for all countries are available in the
Supplementary Dataset and all equations are located in the Supplementary
Information.

\section{Results}\label{results}

We find that climate change could alter life expectancy by -0.22 years
(-2.5 months) in the average European country (\textbf{Fig. 1}). This
reduction is comparable to mortality due to influenza and pneumonia
\citep{arias2013united} in the United States. Although the average
European country could see a change of -0.22 years, several countries
are likely to experience considerably greater reductions in life
expectancy. Luxembourg could see a change of up to -2.2 years and the
medium-variant of climate hazards could cost the average Spaniard 0.98
years of life.

\begin{figure}
\centering
\includegraphics{MS-cclifeexpec_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{Change in life expectancy at birth (\(e_0\)) due to
business-as-usual climate change. We report changes in life expectancies
due to climate change for twenty-eight European countries. The central
values represent the ensemble median while the stems represent the upper
and lower bounds of the inter-model climate variability.}
\end{figure}

Our results also suggest climate change mortality differentials are
likely to unfold along highly uneven geographies. Whereas many Northern
European countries could experience negligible impacts on life
expectancy, five European countries could see life expectancy changes in
excess of -0.5 years (Spain, Luxembourg, France, Italy, and Ireland).
This group of countries could see life expectancy changes of more than
-1.0 years if climate change hazards are more intense than anticipated.
In some countries, climate change could thus become a bigger killer than
trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute
myocardial infarction (-0.87 years), or all accident related mortality
(-0.84 years) \citep{arias2013united} by the end of the century.

\begin{figure}
\centering
\includegraphics{MS-cclifeexpec_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{Estimated reduction in life expectancy at birth (\(e_0\)) by
the 2080s under the \(MID\) scenario.}
\end{figure}

While Forzieri et al's \citep{forzieri2017increasing} results suggest
that the greatest climate change related mortality will unfold along a
north-south gradient, we find the greatest reduction in life expectancy
unfolds along an east-west gradient (\textbf{Fig. 2}). The most westerly
European countries tend to have the greatest reductions. Neighboring
pairs of countries at similar latitudes could experience vastly
different mortality regimes, but neighboring pairs of countries at
similar longitudes seem to experience more similar mortality regimes
from climate change. Additionally, these results point to the importance
of converting excess mortality into life expectancy to properly quantify
the effects of climate change mortality.

\section{Discussion}\label{discussion}

In this article, we demonstrate the impact climate change could have on
life expectancy at birth in twenty-eight European countries. Previous
studies on climate change and excess mortality potentially misrepresent
the impact climate change could have on human mortality. Contrary to
excess mortality estimates, life expectancy is routinely used as a
primary metric for communicating overall health outcomes and enjoys
widespread use by major international organizations
\citep{world2015world, marmot2012building, salomon2012healthy}. Life
expectancy and its derivatives are the recommended metrics for
population health \citep{parrish2010peer}. Additionally, it connects
mortality estimates into intuitively understandable metrics, translating
global estimates of mortality into individual outcomes. Our work reveals
the extent to which climate change could reduce the average person's
longevity by the end of the century, the first study of its kind,
significantly expanding our understanding of climate change and public
health; thus linking two of the major areas in current developmental and
sustainability discussion at both national and international levels
\citep{abel2016meeting}.

Without adaptation measures, our results suggest climate change could
emerge as a significant new mortality vector and could pose a major
public health threat for some European countries by the end of the
century, echoing previous findings
\citep{forzieri2017increasing, patz2005impact}. Life expectancy has
steadily risen across the world for the last century
\citep{oeppen2002broken} and our results suggest that climate change
alone could spur a sharp reversal in these trends. We expect one
European countries to see life expectancy reductions more than one year
under the middle scenario (Luxembourg), but if climate change has a
greater impact on mortality than anticipated, five European countries
(Luxembourg, Spain, Italy, France, Ireland) could see life expectancy
reductions more than one year with Luxembourg experiencing a reduction
of more than two.

Reductions such as those should not be taken lightly. Many of the
children born today are likely to still be alive by the end of the
century and will be in the age groups most threatened by extreme heat -
the biggest mortality risk associated with climate change
\citep{keatinge2000heat}. If climate change unfolds as a more aggressive
mortality vector, only major cardiovascular diseases and malignant
neoplasms would contribute more to life expectancy reductions in
numerous European countries \citep{arias2013united}. This would make
climate change one of the most aggressive new mortality vectors to
emerge over the last quarter-century, representing a major threat to
public health in many parts of Europe.

Prospective studies on the emerging threat from climate change rely on
linking contemporary mortality with future mortality. However, climate
change could reshape future mortality through other causes of death.
Climate change affects health behaviors that in turn increase mortality
risk through increased alcohol and substance abuse, violent behavior,
insecurity, increase in post-traumatic stress due to weather-related
trauma, increase in stress due to climate change and schizophrenia,
increase in the use of medications that reduce the ability to perspire
and sweat, etc. \citep{patz2005impact} The International Classifications
of Diseases and Related Health Problems (ICD) does not contain ``climate
change" as an official cause of death, so we can only speculate that the
impact of climate change could be larger than reported here. Although we
do not model these potential impacts, our results could thus be
considered conservative.

We also share the concerns of Lee et al. \citep{lee2017comprehensive}
concerning the business-as-usual climatic assumptions. It is likely that
many countries and communities will deploy a wide variety of adaptation
measures \citep{haines2006climate, kovats2003methods, ebi2006approach}.
These adaptation measures rely on accurate information about the
potential mortality vectors. Our models and those produced by Forzieri
et al. \citep{forzieri2017increasing} present plausible scenarios on the
potential impact of climate change on human mortality and provide
crucial information to public health officials, national governments,
and international organizations. The time frames associated with climate
change allow ample time for this potential health crisis to be averted.

Finally, we would like to point out that future research should not only
transform excess mortality into life expectancy decrements. Given the
influence of climate change in diseases and causes of death, it is
imperative to quantify the extent to which climate change will derive in
increasing costs for health care systems in these countries. The health
care structures are being taxed by population aging
\citep{rechel2009can} and rising health care costs, yet it remains
unclear how climate change will exacerbate these pressures.

\newpage

\section{Supplemenatary Information}\label{supplemenatary-information}

\subsection{Data}\label{data}

To create life expectancy differentials, we used three sources of data.
The first is the baseline life table data. Baseline life tables comes
from the Human Mortality Database (HMD) \citep{HMD}. These datasets
comprise only official vital event registration and are considered the
most complete and accurate source of mortality data available. We
downloaded the most recent 1x5 (by age and year) life table data for
twenty-eight European countries. We used the following data elements
from the five-year life table data in the HMD:

\begin{enumerate}
\item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
\item $_na_x$ or the average length of survival between ages $x$ and $x+n$.
\item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
\end{enumerate}

The second dataset comes from published projections of climate-related
mortality in Europe \citep{forzieri2017increasing}. Forzieri et al.
\citep{forzieri2017increasing} combined data from the International
Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool
(NatCatSERVICE) to create exposure and fatality statistics related to
six weather-related hazards -- heatwaves, cold waves, droughts,
wildfires, river and coastal floods, and windstorms. They downscaled
these exposures to 1km grid scales and integrated them into small-area
demographic projections for Europe out to 2100. Heatwaves and coldwaves
account for the vast majority of projected fatalities. The number of
future anticipated deaths are available in their Supplementary Table S8
\citep{forzieri2017increasing} and provide the magnitude of deaths for
our study.

The third dataset comes from the GBD \citep{GBD, wang2012age}. The GBD
is the most comprehensive worldwide dataset of epidemiological data
produced and provides cause-specific mortality by age/sex/geography/year
for 249 causes of death in 195 countries and territories. We gathered
data on age-specific deaths and mortality rates (\(m_x\)) for mortality
from environmental heat and cold exposure (cause C.2.9) for the period
2006-2015 for each country in the study. These data provide a mortality
schedule to fit the projected climate change excess mortality from
Fozieri et al \citep{forzieri2017increasing}.

\subsection{Methods}\label{methods-1}

We derived a number of additional variables to accomplish our analysis.
First, we abridged the life table data from the HMD from 100+ to 80+ to
conform to the GBD cause-specific mortality schedules.

From the GBD data, we derived variable \(t_{x,i}\) as the proportion of
deaths, \(D\), from each age group \(x\) in each country \(i\)
(\(D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}\)).

We then derived additional \(m_x\) rates for each age group \(x\) in
each country \(i\) for each scenario \(s\) (\(BASE\), \(LOW\), \(MID\),
\(HIGH\)) from Forzieri et al \citep{forzieri2017increasing}.

\[ _nm_{x,i,BASE} = _nD_{x,i}^{HMD} / _nP_{x,i}^{HMD} \]
\[ _n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + \hat{D_i} \cdot _nt_{x,i} ) / _nP_{x,i}^{HMD} \]

where \(D_{x,i}\) is the number of deaths in age group \(x\) in country
\(i\) from the HMD, \(P_{x,i}\) is the population in age group \(x\)
from the HMD, \(\hat{D_x}\) is the number of deaths from Forzieri et al.
\citep{forzieri2017increasing}, and \(t_{x,i}\) is the proportion of
mortality experienced in each age group \(x\) from the GBD. Thus, the
anticipated additional mortality due to climate change is added to each
age group based on the underlying mortality cause-specific mortality
schedule observed between 2006-2015 in the GBD.

We then calculated \(q_x\) values, or the probability of dying, for each
scenario \(s\) for each country.

\[_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}\]
\[_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}\]

We calculated each additional life table value identically for each
scenario \(s\) using the standard life table equations:

\[_nd_{x,i,s} = _nl_{x,i,s} \cdot _nq_{x,i,s}\]
\[_nl_{x,i,s} = _nl_{x-1,i,s} - _nd_{x-1,i,s}\]
\[_nL_{x,i,s} = _na_{x,i,s} \cdot _nl_{x,i,s} + ((n-_na_{x,i,s}) \cdot _nl_{x,i,s})\]
\[_nT_{x,i,s} = _nL_{x,i,s} + _nT_{x+1,i,s}\]
\[_ne_{x,i,s} = _nT_{x,i,s} / _nl_{x,i,s}\]

To determine the differences between life expectancy compared to the
baseline, we simply subtract \(e_{x,i,s}\) (for scenario \(LOW\),
\(MID\), and \(HIGH\)) from \(e_{x,i,BASE}\) as described by
\citep{beltran2008integrated}.
\newpage
\singlespacing 
\bibliography{C:/Users/Matt/Documents/Untitled/ccmort}
\end{document}

### 2017-11-22 lancet_cc_09032017.R
uploads

@@ -1,0 +1,184 @@
+rm()
+library(tidyverse)
+library(HMDHFDplus)
+library(LifeTables)
+library(xlsx)
+library(tmap)
+library(tmaptools)
+library(RColorBrewer)
+
+
+myusername <- userInput()
+mypassword <- userInput()
+
+lancetdata <- read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/LANcET_CCMORTDATA.csv",colClasses=c("CNTRY"="character"))
+
+Countries <- getHMDcountries()
+
+deaths <- do.call(rbind, lapply(Countries, function(CNTRY){
+  Dat <- readHMDweb(CNTRY = CNTRY, item = "Deaths_5x1", fixup=TRUE, username = myusername, password = mypassword)
+Dat$CNTRY <- CNTRY
+Dat}))
+
+pops <- do.call(rbind, lapply(Countries, function(CNTRY){
+  Dat <- readHMDweb(CNTRY = CNTRY, item = "Population", fixup=TRUE, username = myusername, password = mypassword)
+  Dat$CNTRY <- CNTRY
+  Dat}))
+
+lt <- do.call(rbind, lapply(Countries, function(CNTRY){
+  Dat <- readHMDweb(CNTRY = CNTRY, item = "bltper_5x1", fixup=TRUE, username = myusername, password = mypassword)
+  Dat$CNTRY <- CNTRY
+  Dat}))
+
+
+lt2 <- lt%>%
+  mutate(Age = ifelse(Age >= 80, 80, Age)) %>%
+  group_by(CNTRY, Year, Age)%>%
+  summarise(ax = sum(ax))
+
+deaths2 <- deaths %>%
+  mutate(Age = ifelse(Age >=80, 80, Age)) %>%
+  group_by(CNTRY, Year, Age) %>%
+  summarise(deaths = sum(Total) )
+
+
+pops2 <- pops %>%
+  mutate(Age = ifelse(Age >= 80, 80, 
+               ifelse(Age >= 75, 75,
+               ifelse(Age >= 65, 65,
+               ifelse(Age >= 60, 60,
+               ifelse(Age >= 55, 55,
+               ifelse(Age >= 50, 50,
+               ifelse(Age >= 45, 45,
+               ifelse(Age >= 40, 40,
+               ifelse(Age >= 35, 35,
+               ifelse(Age >= 30, 30,
+               ifelse(Age >= 25, 25,
[…]

### 2018-02-18 lancet_cc_09032017.R
projfile

@@ -6,26 +6,61 @@ library(xlsx)
 library(tmap)
 library(tmaptools)
 library(RColorBrewer)
+library(data.table)
 
 
 myusername <- userInput()
 mypassword <- userInput()
 
-lancetdata <- read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/LANcET_CCMORTDATA.csv",colClasses=c("CNTRY"="character"))
+lancetdata <- tribble(
+  ~CNTRY,~MID,~LOW,~HIGH,
+  "AUT",467,121,797,
+  "BEL",2841,91,5007,
+  "BGR",44,36,63,
+  "CHE",1981,497,3247,
+  "CYP",143,62,279,
+  "CZE",375,48,628,
+  "DEUTNP",7319,620,12132,
+  "DNK",24,4,56,
+  "EST",12,0,22,
+  "ESP",41326,33495,53802,
+  "FIN",23,3,44,
+  "FRATNP",38728,14235,72947,
+  "GRC",2838,1520,5606,
+  "HRV",1145,624,1811,
+  "HUN",559,191,985,
+  "IRL",1566,476,4195,
+  "ISL",3,1,5,
+  "ITA",41965,23694,57401,
+  "LTU",15,1,25,
+  "LUX",392,12,820,
+  "LVA",17,1,30,
+  "MLT",0,0,0,
+  "NLD",1984,126,4114,
+  "NOR",15,4,34,
+  "POL",313,52,561,
+  "PRT",4573,3808,6284,
+  "ROU",177,97,301,
+  "SWE",52,10,113,
+  "SVN",561,251,847,
+  "SVK",63,17,118,
+  "GBR_NP",2515,388,7484
+)
+  read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/LANcET_CCMORTDATA.csv",colClasses=c("CNTRY"="character"))
 
 Countries <- getHMDcountries()
 
-deaths <- do.call(rbind, lapply(Countries, function(CNTRY){
+deaths <- rbindlist(lapply(Countries, function(CNTRY){
   Dat <- readHMDweb(CNTRY = CNTRY, item = "Deaths_5x1", fixup=TRUE, username = myusername, password = mypassword)
 Dat$CNTRY <- CNTRY
 Dat}))
 
-pops <- do.call(rbind, lapply(Countries, function(CNTRY){
+pops <- rbindlist(lapply(Countries, function(CNTRY){
[…]

### 2018-02-19 lancet_cc_09032017.R
Added comments

@@ -8,10 +8,11 @@ library(tmaptools)
 library(RColorBrewer)
 library(data.table)
 
-
+###   Inputting user info for HMD
 myusername <- userInput()
 mypassword <- userInput()
 
+###   Importing the data from Forizeri et al.
 lancetdata <- tribble(
   ~CNTRY,~MID,~LOW,~HIGH,
   "AUT",467,121,797,
@@ -46,37 +47,42 @@ lancetdata <- tribble(
   "SVK",63,17,118,
   "GBR_NP",2515,388,7484
 )
-  read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/LANcET_CCMORTDATA.csv",colClasses=c("CNTRY"="character"))
-
+#  read.csv("data/LANcET_CCMORTDATA.csv",colClasses=c("CNTRY"="character"))
+ 
+###   Getting a country list from the HMD
 Countries <- getHMDcountries()
 
+###   Downloading the DEATHS data from the HMD in 5-year age groups by single-year.
 deaths <- rbindlist(lapply(Countries, function(CNTRY){
   Dat <- readHMDweb(CNTRY = CNTRY, item = "Deaths_5x1", fixup=TRUE, username = myusername, password = mypassword)
 Dat$CNTRY <- CNTRY
 Dat}))
 
+###   Downloading the POPULATION data from the HMD.
 pops <- rbindlist(lapply(Countries, function(CNTRY){
   Dat <- readHMDweb(CNTRY = CNTRY, item = "Population", fixup=TRUE, username = myusername, password = mypassword)
   Dat$CNTRY <- CNTRY
   Dat}))
 
+###   Downloading the LIFE TABLE data from the HMD in 5-year age groups by single-year.
 lt <- rbindlist(lapply(Countries, function(CNTRY){
   Dat <- readHMDweb(CNTRY = CNTRY, item = "bltper_5x1", fixup=TRUE, username = myusername, password = mypassword)
   Dat$CNTRY <- CNTRY
   Dat}))
 
-
+###   Topcoding the age groups to 80+ of the LIFE TABLE data, and summing the ax value.
 lt2 <- lt%>%
   mutate(Age = ifelse(Age >= 80, 80, Age)) %>%
   group_by(CNTRY, Year, Age)%>%
   summarise(ax = sum(ax))
 
+###   Topcoding the age groups of the DEATHS data and summing.
 deaths2 <- deaths %>%
[…]

### 2018-02-19 MS-cclifeexpec.Rmd
fixed some file locations.

@@
 
-lt_lancet <- read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/lt_afterr_09032017.csv")
-lt_lifetables <- read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/lancet_lifetables_09032017.csv")
+lt_lancet <- read.csv("data/lt_afterr_09032017.csv")
+lt_lifetables <- read.csv("data/lancet_lifetables_09032017.csv")
 

### 2018-02-19 lancet_cc_09032017.R

@@ -149,10 +149,10 @@ a2 <- left_join(a, GBD_data) %>%
          qx_high = ifelse(Age == 80, 1, mx_high/(1+((width-ax)*mx_high))))
 
 
-write.table(a2, "C:/Users/Matt/Documents/Matt - Work Stuff/r code/r_output2.txt", sep="\t")
+write.table(a2, "data/r_output2.txt", sep="\t")
 
-lt_lancet <- read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/lt_afterr_09032017.csv")
-lt_lifetables <- read.csv("C:/Users/Matt/Documents/Matt - Work Stuff/r code/lancet_lifetables_09032017.csv")
+lt_lancet <- read.csv("data/lt_afterr_09032017.csv")
+lt_lifetables <- read.csv("data/lancet_lifetables_09032017.csv")
 
 lt_lancet$'country name' <- lt_lancet$COUNTRY
 lt_lancet<- lt_lancet[order(lt_lancet$DIF_MID), ]

### 2018-02-19 MS-cclifeexpec.Rmd
put the latex template in

@@
     latex_engine: pdflatex
-    template: C:/Users/Matt/Documents/Matt - Work Stuff/r code/delete/svm-latex-ms.tex
+    template: LATEX/svm-latex-ms.tex
 title: "Climate change projected to reduce European life expectancy"
@@
 spacing: double
-bibliography: C:/Users/Matt/Documents/Untitled/ccmort.bib
+bibliography: LATEX/ccmort.bib
 biblio-style: nature

### 2018-03-21 MS-cclifeexpec.Rmd
minor edits.

@@
 title: "Climate change projected to reduce European life expectancy"
-thanks: "All data and code that supports these conclusions are available upon request at hauer@uga.edu."
+thanks: "All data and code that supports these conclusions are available as supplementary materials."
 author:
-- name: Mathew E. Hauer
+- name: Mathew E. Hauer^1^*
   affiliation: University of Georgia
-- name: Alexis R. Santos
+- name: Alexis R. Santos^2^
   affiliation: Pennsylvania State University
@@
 abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change.
-However, life expectancy at birth is more easily communicated to the public, understood by the public, and controls for different geographic, economic, and demographic groups.
+However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups.
 No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy.
@@
 - \usepackage{rotating}
+- \usepackage{fancyhdr}
+- \pagestyle{fancy}
+- \fancyhead{}
+- \fancyhead[CE,CO]{Climate change life expectancy}
 ---
+
+* Corresponding author. hauer@uga.edu. p: 706-542-9369.
+
+^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.
+
+^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.
+
+\newpage
 
@@
 Forzieri et al.
-[@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario.
-They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases.
+[@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario (SRES A1B).
+They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario (Shared-Socioeconomic Pathway 2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario) and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases within spatially rectified, high-resolution gridded datasets.
+While the disaster data are not standardized across country, Forzeri et al.
+imputed data in incomplete time periods and countries.
+This imputation could mask spatial variability at the sub-national level, but here we use their country-aggregated results.
 Their results represent business-as-usual climate change and human development without incorporating potential adaptation.
@@
 # Results
-We find that climate change could alter life expectancy by -0.22 years (-2.5 months) in the average European country (**Fig.
+We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig.
 1**).
@@
 Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change.
+Emergent research in Atmospheric Rivers suggest Western Europe could be more susceptible to these types of events [@ramos2015daily].
 Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.
@@
 Reductions such as those should not be taken lightly.
-Many of the children born today are likely to still be alive by the end of the century and will be in the age groups most threatened by extreme heat - the biggest mortality risk associated with climate change [@keatinge2000heat].
+Many of the children born today are likely to still be alive by the end of the century and will be in the age groups most threatened by extreme heat (65+) - the biggest mortality risk associated with climate change [@keatinge2000heat].
 If climate change unfolds as a more aggressive mortality vector, only major cardiovascular diseases and malignant neoplasms would contribute more to life expectancy reductions in numerous European countries [@arias2013united].
@@
 These datasets comprise only official vital event registration and are considered the most complete and accurate source of mortality data available.
-We downloaded the most recent 1x5 (by age and year) life table data for twenty-eight European countries.
-We used the following data elements from the five-year life table data in the HMD:
+We downloaded the most recent 5x1 (by age and year) life table data for twenty-eight European countries.
+We used the following data elements from the five-year age group life table data in the HMD:
 
@@
 
+```{r suppfig, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("**Mortality distribution of Heat-related mortality in 28 European countries**.
+The thick black line is the average\\label{suppfig}")}
+GBD_data <- read.csv("data/GBD_data.csv") %>%
+  #mutate(perdist = log(val)) %>%
+  group_by(countrycode, Age) %>%
+  summarise(meanval = mean(val),
+            upperval = mean(upper),
+            lowerval = mean(lower)) %>%
+  group_by(countrycode) %>%
+  mutate(countmeanval = sum(meanval),
+         countupperval = sum(upperval),
+         countlowerval = sum(lowerval),
+         perdist = (meanval / countmeanval)) %>%
+  rename(CNTRY = countrycode)
+
+ggplot(data=GBD_data, aes(x=Age, y=perdist)) +
+  geom_line(aes(color=CNTRY), alpha=1) +
+  geom_smooth(span=0.3, se=F, col="black", lwd =2) +
+  theme_bw() +
+  theme(legend.position = c(0.8,0.25),
+        legend.title=element_blank(),
+        legend.text=element_text(size=4),
+        legend.key.size = unit(1,"line")) +
+  guides(col = guide_legend(ncol = 4)) +
+  scale_y_log10()+
+  labs(y="log(Mortality Percentage Distribution)",
+       #title = "Mortality distribution of Heat-related mortality in 28 European countries",
+       caption = "Source: Global Burden of Disease")
+
+```
+
 ## Methods
-We derived a number of additional variables to accomplish our analysis. First, we abridged the life table data from the HMD from 100+ to 80+ to conform to the GBD cause-specific mortality schedules.
+We derived a number of additional variables to accomplish our analysis.
+First, we abridged the life table data from the HMD from ages 100+ to 80+ to conform to the GBD cause-specific mortality schedules.
 
-From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}$).
+From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}$).
 
@@
 
-$$ _nm_{x,i,BASE} = _nD_{x,i}^{HMD} / _nP_{x,i}^{HMD} $$
-$$ _n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + \hat{D_i} \cdot _nt_{x,i} ) / _nP_{x,i}^{HMD} $$
+\begin{equation}
+_nm_{x,i,BASE} = _nD_{x,i}^{HMD} \,/\, _nP_{x,i}^{HMD}
+\end{equation}
+
+\begin{equation}
+_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_i} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
+\end{equation}
 
@@
 
-$$_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}$$
-$$_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}$$
+\begin{equation}
+_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}
+\end{equation}
 
-We calculated each additional life table value identically for each scenario $s$ using the standard life table equations:
+\begin{equation}
+_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}
+\end{equation}
+
+We calculated each additional life table value identically for each scenario $s$ using standard life table equations:
 

### 2018-03-23 MS-cclifeexpec.Rmd
fixed some minor errors. Ready for age-standardized mortality rates.

--- 2018-02-19 MS-cclifeexpec.Rmd
@@
 library(tidyverse)
-library(RColorBrewer)
-library(scales)
+library(HMDHFDplus)
+library(LifeTables)
+library(xlsx)
 library(tmap)
 library(tmaptools)
+library(RColorBrewer)
+library(data.table)
 

### 2018-03-23 lancet_cc_09032017.R

@@ -47,6 +47,34 @@ lancetdata <- tribble(
   "SVK",63,17,118,
   "GBR_NP",2515,388,7484
 )
+
+ec_agestandard <- tribble(
+  ~Age, ~StnrdPop,
+  0, 1600,
+  1, 6400,
+  5, 7000,
+  10,7000,
+  15,7000,
+  20,7000,
+  25,7000,
+  30,7000,
+  35,7000,
+  40,7000,
+  45,7000,
+  50,7000,
+  55,6000,
+  60,5000,
+  65,4000,
+  70,3000,
+  75,2000,
+  80,2000
+)
+
+countrycodes <- read.csv("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv") %>%
+  rename(CNTRY = alpha.3,
+         COUNTRY = name) %>%
+  mutate(ISO3 = CNTRY) %>%
+  select(COUNTRY, CNTRY, ISO3)
 #  read.csv("data/LANcET_CCMORTDATA.csv",colClasses=c("CNTRY"="character"))
  
 ###   Getting a country list from the HMD
@@ -86,6 +114,7 @@ deaths2 <- deaths %>%
 pops2 <- pops %>%
   mutate(Age = ifelse(Age >= 80, 80, 
                ifelse(Age >= 75, 75,
+               ifelse(Age >= 70, 70,
                ifelse(Age >= 65, 65,
                ifelse(Age >= 60, 60,
                ifelse(Age >= 55, 55,
@@ -99,7 +128,7 @@ pops2 <- pops %>%
                ifelse(Age >= 15, 15,
                ifelse(Age >= 10, 10,
                ifelse(Age >= 5,  5,
-               ifelse(Age >= 1,  1, Age))))))))))))))))) %>%
+               ifelse(Age >= 1,  1, Age)))))))))))))))))) %>%
   group_by(CNTRY, Year, Age) %>%
   summarise(Pop= sum(Total1))
 
@@ -146,11 +175,62 @@ a2 <- left_join(a, GBD_data) %>%
          qx_base = ifelse(Age == 80, 1, mx_base/(1+((width-ax)*mx_base))),
          qx_low = ifelse(Age == 80, 1, mx_low/(1+((width-ax)*mx_low))),
[…]

### 2018-03-23 SCRIPTS/001-ANALYSIS.R
New analysis

@@ -1,0 +1,310 @@
+rm()
+library(tidyverse)
+library(HMDHFDplus)
+library(LifeTables)
+library(xlsx)
+library(tmap)
+library(tmaptools)
+library(RColorBrewer)
+library(data.table)
+
+###   Inputting user info for HMD
+myusername <- userInput()
+mypassword <- userInput()
+
+###   Importing the data from Forizeri et al.
+lancetdata <- tribble(
+  ~CNTRY,~MID,~LOW,~HIGH,
+  "AUT",467,121,797,
+  "BEL",2841,91,5007,
+  "BGR",44,36,63,
+  "CHE",1981,497,3247,
+  "CYP",143,62,279,
+  "CZE",375,48,628,
+  "DEUTNP",7319,620,12132,
+  "DNK",24,4,56,
+  "EST",12,0,22,
+  "ESP",41326,33495,53802,
+  "FIN",23,3,44,
+  "FRATNP",38728,14235,72947,
+  "GRC",2838,1520,5606,
+  "HRV",1145,624,1811,
+  "HUN",559,191,985,
+  "IRL",1566,476,4195,
+  "ISL",3,1,5,
+  "ITA",41965,23694,57401,
+  "LTU",15,1,25,
+  "LUX",392,12,820,
+  "LVA",17,1,30,
+  "MLT",0,0,0,
+  "NLD",1984,126,4114,
+  "NOR",15,4,34,
+  "POL",313,52,561,
+  "PRT",4573,3808,6284,
+  "ROU",177,97,301,
+  "SWE",52,10,113,
+  "SVN",561,251,847,
+  "SVK",63,17,118,
+  "GBR_NP",2515,388,7484
+)
+
+ec_agestandard <- tribble(
+  ~Age, ~StnrdPop,
+  0, 1600,
+  1, 6400,
+  5, 7000,
+  10,7000,
+  15,7000,
+  20,7000,
+  25,7000,
+  30,7000,
+  35,7000,
+  40,7000,
+  45,7000,
+  50,7000,
+  55,6000,
+  60,5000,
+  65,4000,
+  70,3000,
+  75,2000,
+  80,2000
+)
+
+countrycodes <- read.csv("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv") %>%
+  rename(CNTRY = alpha.3,
+         COUNTRY = name) %>%
+  mutate(ISO3 = CNTRY) %>%
[…]

### 2018-03-23 MS-cclifeexpec.Rmd
Added causes of death.

--- 2018-03-21 MS-cclifeexpec.Rmd
@@
 - \usepackage{fancyhdr}
+- \usepackage[table]{xcolor}
+- \usepackage{tabularx}
+- \usepackage{makecell}
+- \usepackage{xcolor}
 - \pagestyle{fancy}
@@
 
-```{r setup, include=FALSE}
+```{r load, include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)
@@
 library(tidyverse)
-library(RColorBrewer)
-library(scales)
+library(HMDHFDplus)
 library(tmap)
 library(tmaptools)
+library(RColorBrewer)
+library(data.table)
+library(kableExtra)
+library(knitr)
+library(xtable)
+```
 
-lt_lancet <- read.csv("data/lt_afterr_09032017.csv")
-lt_lifetables <- read.csv("data/lancet_lifetables_09032017.csv")
+```{r setup, include=FALSE, cache =TRUE}
+knitr::opts_chunk$set(echo = TRUE)
+library(png)
+rm(list=ls())
+library(tidyverse)
+library(HMDHFDplus)
+library(tmap)
+library(tmaptools)
+library(RColorBrewer)
+library(data.table)
 
-lt_lancet$'country name' <- lt_lancet$COUNTRY
-lt_lancet<- lt_lancet[order(lt_lancet$DIF_MID), ]
-lt_lancet$'country name' <- factor(lt_lancet$'country name', levels = lt_lancet$'country name')
+###   Inputting user info for HMD
+
+myusername <- "hauer@uga.edu"
+mypassword <- "fuckhead"
+
+###   Importing the data from Forizeri et al.
+lancetdata <- tribble(
+  ~CNTRY,~MID,~LOW,~HIGH,
+  "AUT",467,121,797,
+  "BEL",2841,91,5007,
+  "BGR",44,36,63,
+  "CHE",1981,497,3247,
+  "CYP",143,62,279,
+  "CZE",375,48,628,
+  "DEUTNP",7319,620,12132,
+  "DNK",24,4,56,
+  "EST",12,0,22,
+  "ESP",41326,33495,53802,
+  "FIN",23,3,44,
+  "FRATNP",38728,14235,72947,
+  "GRC",2838,1520,5606,
+  "HRV",1145,624,1811,
+  "HUN",559,191,985,
+  "IRL",1566,476,4195,
+  "ISL",3,1,5,
+  "ITA",41965,23694,57401,
+  "LTU",15,1,25,
+  "LUX",392,12,820,
+  "LVA",17,1,30,
+  "MLT",0,0,0,
+  "NLD",1984,126,4114,
+  "NOR",15,4,34,
+  "POL",313,52,561,
+  "PRT",4573,3808,6284,
+  "ROU",177,97,301,
+  "SWE",52,10,113,
+  "SVN",561,251,847,
+  "SVK",63,17,118,
+  "GBR_NP",2515,388,7484
+)
+
+# ec_agestandard <- tribble(
+#   ~Age, ~StnrdPop,
+#   0, 1600,
+#   1, 6400,
+#   5, 7000,
+#   10,7000,
+#   15,7000,
+#   20,7000,
+#   25,7000,
+#   30,7000,
+#   35,7000,
+#   40,7000,
+#   45,7000,
+#   50,7000,
+#   55,6000,
+#   60,5000,
+#   65,4000,
+#   70,3000,
+#   75,2000,
+#   80,2000
+# )
+
+ec_agestandard <- tribble(
+  ~Age, ~StnrdPop,
+  0, 1118.427,
+  1, 4338.143,
+  5, 5207.188,
+  10, 5378.67,
+  15, 6095.53,
+  20, 6646.578,
+  25, 7054.462,
+  30, 7211.366,
+  35, 7249.137,
+  40, 7288.966,
+  45, 7207.381,
+  50, 6904.728,
+  55, 6400.144,
+  60, 5798.191,
+  65, 4660.589,
+  70, 4031.35,
+  75, 3292.724,
+  80,4116.422
+)
+
+europe_standarddr <- tribble(
+  ~COUNTRY, 	~Circulatorydisease, 	~Cancer, 	~Heartdisease, 	~LungCancer, 	~Respiratorydiseases, 	~Diseasesofthenervoussystem, 	~Colorectalcancer, 	~Suicide, 	~TransportAccidents, 
+  'EU-28', 	373.6, 	261.5, 	126.3, 	54.4, 	78.3, 	38.6, 	30.5, 	11.3, 	5.8, 
+  'Luxembourg', 	296.9, 	260.7, 	80.3, 	59.6, 	63.8, 	38, 	25.5, 	13.4, 	6, 
+  'Spain', 	245, 	232.7, 	68.2, 	47.8, 	91.7, 	48.5, 	33.6, 	8.2, 	4.3, 
+  'Italy', 	310.1, 	246.6, 	98.3, 	49.4, 	58.3, 	34.3, 	27, 	6.3, 	5.6, 
+  'France', 	202.9, 	245.4, 	49.3, 	50.1, 	52, 	50.2, 	26.1, 	14.1, 	5.1, 
+  'Portugal', 	305.8, 	242.1, 	69.6, 	36.4, 	116.7, 	32.8, 	35, 	11.3, 	7.8, 
+  'Croatia', 	678.6, 	336.4, 	306.5, 	65.2, 	59.7, 	21.3, 	51, 	16.8, 	8.9, 
+  'Ireland', 	309.9, 	288.3, 	147.5, 	61.5, 	125.9, 	48.7, 	32.4, 	11, 	4, 
+  'Slovenia', 	451.3, 	299.9, 	102.8, 	58.6, 	66.3, 	21.1, 	38.4, 	18.9, 	6.7, 
+  'Switzerland', 	280, 	219.6, 	97.8, 	42.1, 	51.3, 	44.5, 	22.8, 	12.8, 	3.6, 
+  'Belgium', 	281.9, 	252.6, 	72.4, 	61.6, 	95.7, 	46.5, 	26.1, 	17.3, 	6.7, 
+  'Greece', 	381.4, 	249.3, 	103, 	61.9, 	108.1, 	20.9, 	23.3, 	5, 	8.6, 
+  'Netherlands', 	271.7, 	282.2, 	62.4, 	66.7, 	74.1, 	48.3, 	32.9, 	11.1, 	4.1, 
+  'Germany', 	403.5, 	253.2, 	142.8, 	51, 	68, 	29.6, 	29, 	11.9, 	4.6, 
+  'Great Britain', 	264.9, 	278.4, 	118.4, 	61.4, 	130.9, 	47.6, 	27.7, 	7.1, 	2.8, 
+  'Austria', 	418.1, 	249.3, 	179.1, 	47.5, 	46.6, 	32.6, 	26.4, 	15.3, 	5.8, 
+  'Bulgaria', 	1131, 	242.4, 	195.4, 	45.5, 	58.1, 	15.3, 	34.9, 	9.9, 	9, 
+  'Czech Republic', 	615.2, 	284.6, 	333.1, 	53.1, 	73.4, 	30.8, 	37.9, 	14.4, 	78.8, 
+  'Denmark', 	256.6, 	300.6, 	81, 	71.7, 	115.7, 	42.9, 	35.2, 	11.9, 	4, 
+  'Estonia', 	699.6, 	299.4, 	295.5, 	55.3, 	43.8, 	21.8, 	36, 	18.3, 	7.5, 
+  'Finland', 	378.8, 	218.6, 	199.2, 	39, 	34.4, 	155, 	22.6, 	14.6, 	5.7, 
+  'Hungary', 	761.5, 	348.1, 	390.6, 	89.8, 	78.6, 	19.9, 	55, 	19.4, 	8.1, 
+  'Latvia', 	882.7, 	299.3, 	442.7, 	46.9, 	35.9, 	15.6, 	34.2, 	19, 	12.4, 
+  'Lithuania', 	848.8, 	276.2, 	564.4, 	46.1, 	42.1, 	20.8, 	32.1, 	31.5, 	10.7, 
+  'Norway', 	272.6, 	252.5, 	95.7, 	50.5, 	88.4, 	45.4, 	36.4, 	7.3, 	4, 
+  'Poland', 	591.4, 	292.3, 	129.1, 	69.2, 	69.1, 	16.5, 	36, 	15.5, 	10.3, 
+  'Slovakia', 	654.6, 	324.1, 	388.8, 	50, 	74.9, 	29.5, 	49.2, 	10.8, 	8.5, 
+  'Sweden', 	338.3, 	234.8, 	131.2, 	38.7, 	58.1, 	42.6, 	29.2, 	12.1, 	3.4, 
+  'Cyprus', 	351.8, 	201, 	108.7, 	37.2, 	86.2, 	26.8, 	16.7, 	4.5, 	6.5, 
+  'Liechtenstein', 	296.4, 	203, 	73.7, 	31.3, 	89.8, 	67.6, 	6.8, 	10.2, 	10.3, 
+  'Malta', 	372.4, 	233.5, 	202.8, 	43.2, 	96.6, 	21, 	28.3, 	8.3, 	2.5, 
+  'Romania', 	951.3, 	273.2, 	320.3, 	54.2, 	78.4, 	21, 	32.4, 	11.4, 	12.3, 
+  'Serbia', 	931.6, 	298.3, 	159.5, 	69.4, 	79.7, 	27.3, 	37.2, 	15.9, 	7.6
+
+)
+
+countrycodes <- read.csv("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv") %>%
+  rename(CNTRY = alpha.3,
+         COUNTRY = name) %>%
+  mutate(ISO3 = CNTRY) %>%
+  select(COUNTRY, CNTRY, ISO3)
+#  read.csv("data/LANcET_CCMORTDATA.csv",colClasses=c("CNTRY"="character"))
+
+###   Getting a country list from the HMD
+Countries <- getHMDcountries()
+
+###   Downloading the DEATHS data from the HMD in 5-year age groups by single-year.
+deaths <- rbindlist(lapply(Countries, function(CNTRY){
+  Dat <- readHMDweb(CNTRY = CNTRY, item = "Deaths_5x1", fixup=TRUE, username = myusername, password = mypassword)
+Dat$CNTRY <- CNTRY
+Dat}))
+
+###   Downloading the POPULATION data from the HMD.
+pops <- rbindlist(lapply(Countries, function(CNTRY){
+  Dat <- readHMDweb(CNTRY = CNTRY, item = "Population", fixup=TRUE, username = myusername, password = mypassword)
+  Dat$CNTRY <- CNTRY
+  Dat}))
+
+###   Downloading the LIFE TABLE data from the HMD in 5-year age groups by single-year.
+lt <- rbindlist(lapply(Countries, function(CNTRY){
+  Dat <- readHMDweb(CNTRY = CNTRY, item = "bltper_5x1", fixup=TRUE, username = myusername, password = mypassword)
+  Dat$CNTRY <- CNTRY
+  Dat}))
+
+###   Topcoding the age groups to 80+ of the LIFE TABLE data, and summing the ax value.
+lt2 <- lt%>%
+  mutate(Age = ifelse(Age >= 80, 80, Age)) %>%
+  group_by(CNTRY, Year, Age)%>%
+  summarise(ax = sum(ax))
+
+###   Topcoding the age groups of the DEATHS data and summing.
+deaths2 <- deaths %>%
+  mutate(Age = ifelse(Age >=80, 80, Age)) %>%
+  group_by(CNTRY, Year, Age) %>%
+  summarise(deaths = sum(Total) )
+
+###   Recoding the Population data to single year of age
+pops2 <- pops %>%
+  mutate(Age = ifelse(Age >= 80, 80, 
+               ifelse(Age >= 75, 75,
+               ifelse(Age >= 70, 70,
+               ifelse(Age >= 65, 65,
+               ifelse(Age >= 60, 60,
+               ifelse(Age >= 55, 55,
+               ifelse(Age >= 50, 50,
+               ifelse(Age >= 45, 45,
+               ifelse(Age >= 40, 40,
+               ifelse(Age >= 35, 35,
+               ifelse(Age >= 30, 30,
+               ifelse(Age >= 25, 25,
+               ifelse(Age >= 20, 20,
+               ifelse(Age >= 15, 15,
+               ifelse(Age >= 10, 10,
+               ifelse(Age >= 5,  5,
+               ifelse(Age >= 1,  1, Age)))))))))))))))))) %>%
+  group_by(CNTRY, Year, Age) %>%
+  summarise(Pop= sum(Total1))
+
+###   Joining the POPULATION, DEATHS, and LIFE TABLE data and subseting to the maximum year in the data.
+a <- left_join(pops2, deaths2)
+a <- left_join(a, lt2)
+a <- group_by(a, CNTRY, Age) %>%
+  filter(Year == max(Year))
+
+###   Importing the GBD data.
+###   Data is 5-year age groups of mortality rates for Environmental heat and cold exposure for 2006-2015. 
+###   Analysis uses the mean of the last 10 years of data for this death distribution.
+GBD_data <- read.csv("data/GBD_data.csv") %>%
+  #mutate(perdist = log(val)) %>%
+  group_by(countrycode, Age) %>%
+  summarise(meanval = mean(val),
+            upperval = mean(upper),
+            lowerval = mean(lower)) %>%
+  group_by(countrycode) %>%
+  mutate(countmeanval = sum(meanval),
+         countupperval = sum(upperval),
+         countlowerval = sum(lowerval),
+         perdist = (meanval / countmeanval)) %>%
+  rename(CNTRY = countrycode)
+
+ggplot(data=GBD_data, aes(x=Age, y=perdist)) +
+  geom_line(aes(group =CNTRY, col=CNTRY), alpha=0.7) +
+  geom_smooth(span=0.3, se=F, col="black") +
+  theme_bw() +
+  scale_y_log10()+
+  labs(y="log(Mortality Percentage Distribution)")
+
+a <-  left_join(a, lancetdata)  %>%
+  select(Age, Pop, CNTRY, deaths, ax, MID, LOW, HIGH) %>%
+   filter(!is.na(MID))
+a2 <- left_join(a, GBD_data) %>%
+  mutate(width = ifelse(Age == 0, 1,
+                 ifelse(Age == 1, 4,5)),
+         mx_base = deaths/Pop,
+         mx_low = (deaths + LOW * perdist)/Pop,
+         mx_mid = (deaths + MID * perdist)/Pop,
+         mx_high = (deaths + HIGH * perdist)/Pop,
+         qx_base = ifelse(Age == 80, 1, mx_base/(1+((width-ax)*mx_base))),
+         qx_low = ifelse(Age == 80, 1, mx_low/(1+((width-ax)*mx_low))),
+         qx_mid = ifelse(Age == 80, 1, mx_mid/(1+((width-ax)*mx_mid))),
+         qx_high = ifelse(Age == 80, 1, mx_high/(1+((width-ax)*mx_high)))) %>%
+  group_by(CNTRY) %>%
+  mutate(lx_base = ifelse(is.na(lag(cumprod(1-qx_base),1)*100000), 100000, lag(cumprod(1-qx_base),1)*100000),
+         dx_base = qx_base*lx_base,
+         Lx_base = ifelse(Age == 80, (lx_base/mx_base), ax * lx_base + (width-ax) * lead(lx_base,1)),
+         Tx_base = rev(cumsum(rev(Lx_base))),
+         ex_base = Tx_base/lx_base,
+         lx_low = ifelse(is.na(lag(cumprod(1-qx_low),1)*100000), 100000, lag(cumprod(1-qx_low),1)*100000),
+         dx_low = qx_low*lx_low,
+         Lx_low = ifelse(Age == 80, (lx_low/mx_low), ax * lx_low + (width-ax) * lead(lx_low,1)),
+         Tx_low = rev(cumsum(rev(Lx_low))),
+         ex_low = Tx_low/lx_low,
+         lx_mid = ifelse(is.na(lag(cumprod(1-qx_mid),1)*100000), 100000, lag(cumprod(1-qx_mid),1)*100000),
+         dx_mid = qx_mid*lx_mid,
+         Lx_mid = ifelse(Age == 80, (lx_mid/mx_mid), ax * lx_mid + (width-ax) * lead(lx_mid,1)),
+         Tx_mid = rev(cumsum(rev(Lx_mid))),
+         ex_mid = Tx_mid/lx_mid,
+         lx_high = ifelse(is.na(lag(cumprod(1-qx_high),1)*100000), 100000, lag(cumprod(1-qx_high),1)*100000),
+         dx_high = qx_high*lx_high,
+         Lx_high = ifelse(Age == 80, (lx_high/mx_high), ax * lx_high + (width-ax) * lead(lx_high,1)),
+         Tx_high = rev(cumsum(rev(Lx_high))),
+         ex_high = Tx_high/lx_high,
+         DIF_LOW = ex_low - ex_base,
+         DIF_MID = ex_mid - ex_base,
+         DIF_HIGH = ex_high - ex_base)
+
+# write.table(a2, "data/r_output2.txt", sep="\t")
+
+lt_lancet <- a2 %>%
+  filter(Age == 0) %>%
+  mutate(Base_e0 = ex_base,
+         LOW_e0 = ex_low,
+         MID_e0 = ex_mid,
+         HIGH_e0 = ex_high) %>%
+  #group_by(CNTRY) %>%
+   dplyr::select(CNTRY, Base_e0, LOW_e0, MID_e0, HIGH_e0, DIF_LOW, DIF_MID, DIF_HIGH) %>%
+  ungroup() %>%
+ arrange(DIF_MID) %>%
+  mutate(RANK = row_number(),
+         dif_mid = round(DIF_MID, 2)) %>%
+  left_join(., countrycodes) %>%
+  mutate(COUNTRY = case_when(
+    CNTRY == "FRATNP" ~ "France",
+    CNTRY == "DEUTNP" ~ "Germany",
+    CNTRY == "GBR_NP" ~ "Great Britain",
+    TRUE ~ as.character(COUNTRY)),
+    ISO3 = case_when(
+      CNTRY == "FRATNP" ~ "FRA",
+      CNTRY == "DEUTNP" ~ "DEU",
+      CNTRY == "GBR_NP" ~ "GBR",
+      TRUE ~ as.character(ISO3)),
+    Test = ISO3)
+
 ```
@@
 
-ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=DIF_MID)) +
+ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=dif_mid)) +
   geom_pointrange(stat="identity") +
@@
 Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].
+
+```{r echo = FALSE, warning= FALSE, message = FALSE}
+
+standard <- left_join(a2, ec_agestandard) %>%
+  left_join(., countrycodes) %>%
+  select(Age, CNTRY, COUNTRY, mx_base, mx_low, mx_mid, mx_high, StnrdPop) %>%
+  mutate(ASDR_base = mx_base * StnrdPop,
+         ASDR_low = mx_low * StnrdPop,
+         ASDR_mid = mx_mid * StnrdPop,
+         ASDR_high = mx_high * StnrdPop) %>%
+  group_by(CNTRY, COUNTRY) %>%
+  summarise(ASDR_base = sum(ASDR_base),
+            ASDR_low = sum(ASDR_low),
+            ASDR_mid = sum(ASDR_mid),
+            ASDR_high = sum(ASDR_high)) %>%
+  mutate(ASDR_low = ASDR_low - ASDR_base,
+         ASDR_mid = ASDR_mid - ASDR_base,
+         ASDR_high = ASDR_high - ASDR_base,
+         COUNTRY = case_when(
+           CNTRY == "FRATNP" ~ "France",
+           CNTRY == "DEUTNP" ~ "Germany",
+           CNTRY == "GBR_NP" ~ "Great Britain",
+           TRUE ~ as.character(COUNTRY))) %>%
+  left_join(., europe_standarddr) %>%
+  ungroup() %>%
+  select(-ASDR_base, -CNTRY)
+
+selectcountries <- c("Spain", "Luxembourg", "Italy", "France", "Portugal", "Ireland", "Slovenia", "Croatia", "Switzerland", "Belgium")
+testtbl<- standard %>%
+  group_by(COUNTRY) %>%
+  gather(COD, ASDR, 2:13) %>%
+  filter(!COD %in% c("Circulatorydisease", "Cancer", "ASDR_high", "ASDR_low"),
+         !COUNTRY == "Iceland") %>%
+  mutate(RANK = rank(desc(ASDR)),
+         COD = case_when(
+           COD == "Heartdisease" ~ "makecell{Heart\\Disease\\",
+           COD == "LungCancer" ~ "makecell{Lung\\Cancer\\",
+           COD == "Respiratorydiseases" ~ "makecell{Respiratory\\Diseases\\",
+           COD == "Diseasesofthenervoussystem" ~ "makecell{Dis. of the\\Nervous Sys\\",
+           COD == "Colorectalcancer" ~ "makecell{Colorectal\\Cancer\\",
+           COD == "Suicide" ~ "makecell{Suicide\\",
+           COD == "TransportAccidents" ~ "makecell{Transport\\Accidents\\",
+           COD == "ASDR_mid"  ~ "cellcolor{blue!25}makecell{cellcolor{blue!25}Climate\\cellcolor{blue!25}Change\\cellcolor{blue!25}"),
+         Value = paste0(COD, " ", round(ASDR,2), "}")) %>%
+  ungroup()%>%
+  arrange(COD, desc(ASDR)) %>%
+  #mutate(COUNTRY = paste0(row_number(),COUNTRY))%>%
+  select(COUNTRY, Value, RANK) %>%
+  spread(RANK, Value) %>%
+  slice(match(selectcountries, COUNTRY))
+
+# library(pander)
+# pander(head(testtbl))
+# xtable(testtbl, caption = "A Caption",
+#       comment=F, include.rownames=F, table.placement="ht",
+#       size=getOption("xtable.size", "tiny"))
+# kable(head(testtbl), format="latex") %>%
+#   # column_spec(2, width = "1cm") %>%
+#   kable_styling(latex_options = "scale_down")
+
+# landscape(kable(head(testtbl), "latex"))
+
+  # column_spec(x, 2:9, width = "10em")
+
+# luxrank1 <- standard$Heartdisease[which(standard$COUNTRY == "Luxembourg")]
+# luxrank2 <- standard$ASDR_mid[which(standard$COUNTRY == "Luxembourg")]
+# luxrank3 <- standard$Respiratorydiseases[which(standard$COUNTRY == "Luxembourg")]
+# luxrank4 <- standard$LungCancer[which(standard$COUNTRY == "Luxembourg")]
+# luxrank5 <- standard$Diseasesofthenervoussystem[which(standard$COUNTRY == "Luxembourg")]
+# luxrank6 <- standard$Colorectalcancer[which(standard$COUNTRY == "Luxembourg")]
+# luxrank7 <- standard$Suicide[which(standard$COUNTRY == "Luxembourg")]
+# luxrank8 <- standard$TransportAccidents[which(standard$COUNTRY == "Luxembourg")]
+
+```
+
+\begin{sidewaystable}
+\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} 
+\rowcolors{2}{gray!25}{white}
+\small
+\begin{tabularx}{\textwidth}{l|XXXXXXXXX}
+  \hline
+   \rowcolor{white}
+ Country & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\ 
+  \hline
+Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
+of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\
+Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
+of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\
+Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
+of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\
+France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis.
+of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\
+Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
+of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\
+Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis.
+of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\
+Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02} & \makecell{Dis.
+of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\
+Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44} & \makecell{Dis.
+of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\
+Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis.
+of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\
+Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis.
+of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\
+   \hline
+\end{tabularx}
+\end{sidewaystable}
 

### 2018-03-25 MS-cclifeexpec.Rmd
added a paragraph for the table.

@@
 - \usepackage{xcolor}
+# - \usepackage{threeparttable}
 - \pagestyle{fancy}
@@
 
-# Results
-We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig.
-1**).
-This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States.
-Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy.
-Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.
-
-```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change.
-We report changes in life expectancies due to climate change for twenty-eight European countries.
-The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}
-
-ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=dif_mid)) +
-  geom_pointrange(stat="identity") +
-  geom_point(stat="identity", fill="black", size=8) +
-   geom_hline(yintercept=0) +
-  geom_text(color="white", size=2) +
-  #scale_x_discrete(limits = rev(levels(the_factor)))
-  scale_x_reverse(breaks = lt_lancet$RANK, labels=lt_lancet$COUNTRY) +
-  coord_flip() +
-  theme_bw() +
-  #ylim(-2, 2) +
-  theme(panel.grid.minor = element_blank()) +
-
-   labs(y=expression(Delta* e[0]),
-        x= "Country")
-```
-
-Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies.
-Whereas many Northern European countries could experience negligible impacts on life expectancy, five European countries could see life expectancy changes in excess of -0.5 years (Spain, Luxembourg, France, Italy, and Ireland).
-This group of countries could see life expectancy changes of more than -1.0 years if climate change hazards are more intense than anticipated.
-In some countries, climate change could thus become a bigger killer than trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or all accident related mortality (-0.84 years) [@arias2013united] by the end of the century.
-
-```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}
-
-data(Europe)
-Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)
-
-bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)
-
-  tm_shape(Europe2) +
-  tm_fill("DIF_MID", title = expression(Delta* e[0 ~italic(MID)]), palette = "YlOrBr", style = "jenks") +
-  tm_borders(alpha = 0.5) +  
-  tm_text("Test" , size="AREA", root=5) +
-  tm_format_Europe() 
-```
-
-While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig.
-2**).
-The most westerly European countries tend to have the greatest reductions.
-Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change.
-Emergent research in Atmospheric Rivers suggest Western Europe could be more susceptible to these types of events [@ramos2015daily].
-Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.
-
-# Discussion
-In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries.
-Previous studies on climate change and excess mortality potentially misrepresent the impact climate change could have on human mortality.
-Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy].
-Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer].
-Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes.
-Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].
-
 ```{r echo = FALSE, warning= FALSE, message = FALSE}
@@
 
+# Results
+We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig.
+1**).
+Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy.
+Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.
+
+The overall reduction in $e_0$ is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States.
+**Table 1** reports the ranking of Climate Change compared to the major causes of death (COD) in European countries using age-standardised death rates and the standard European population [@pace2013revision] for the year 2013.
+In three European countries (Spain, Luxembourg, Italy), Climate Change could become a top three COD.
+In Spain (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Spain")],2)`), Luxembourg (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Luxembourg")],2)`), France (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "France")],2)`), and Italy (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Italy")],2)`) Climate Change could become the leading cause of death if Climate Change is more extreme than anticipated.
+Only all circulatory diseases combined and all cancers combined would cause more deaths.
+
+```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change.
+We report changes in life expectancies due to climate change for twenty-eight European countries.
+The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}
+
+ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=dif_mid)) +
+  geom_pointrange(stat="identity") +
+  geom_point(stat="identity", fill="black", size=8) +
+   geom_hline(yintercept=0) +
+  geom_text(color="white", size=2) +
+  #scale_x_discrete(limits = rev(levels(the_factor)))
+  scale_x_reverse(breaks = lt_lancet$RANK, labels=lt_lancet$COUNTRY) +
+  coord_flip() +
+  theme_bw() +
+  #ylim(-2, 2) +
+  theme(panel.grid.minor = element_blank()) +
+
+   labs(y=expression(Delta* e[0]),
+        x= "Country")
+```
+
+Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies.
+Whereas many Northern European countries could experience negligible impacts on life expectancy, five European countries could see life expectancy changes in excess of -0.5 years (Luxembourg, Spain, Italy, France, and Ireland).
+This group of countries could see life expectancy changes of more than -1.0 years if climate change hazards are more intense than anticipated.
+In some countries, climate change could thus become a bigger killer than trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or all accident related mortality (-0.84 years) [@arias2013united] by the end of the century.
+
+```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}
+
+data(Europe)
+Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)
+
+bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)
+
+  tm_shape(Europe2) +
+  tm_fill("DIF_MID", title = expression(Delta* e[0 ~italic(MID)]), palette = "YlOrBr", style = "jenks") +
+  tm_borders(alpha = 0.5) +  
+  tm_text("Test" , size="AREA", root=5) +
+  tm_format_Europe() 
+```
+
+While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig.
+2**).
+The most westerly European countries tend to have the greatest reductions.
+Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change.
+Emergent research in Atmospheric Rivers suggest Western Europe could be more susceptible to these types of events [@ramos2015daily].
+Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.
+
+# Discussion
+In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries.
+Previous studies on climate change and excess mortality potentially misrepresent the impact climate change could have on human mortality.
+Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy].
+Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer].
+Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes.
+Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].
+
 \begin{sidewaystable}
-\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} 
+\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Climate Change COD is based on the 2080s while the other CODs are based on the 2010s.}
 \rowcolors{2}{gray!25}{white}
@@
 \end{tabularx}
+\vspace{1ex}
+ \raggedright Data for other causes of death come from Eurostat (online data code: hlth_cd_asdr2). These results reflect the Standard European Population for 2013.
 \end{sidewaystable}

### 2018-03-25 MS-cclifeexpec.tex
added a paragraph for the table.

@@
 
-\title{Climate change projected to reduce European life expectancy \thanks{All data and code that supports these conclusions are available upon
-request at \href{mailto:hauer@uga.edu}{\nolinkurl{hauer@uga.edu}}.}  }
+\title{Climate change projected to reduce European life expectancy \thanks{All data and code that supports these conclusions are available as
+supplementary materials.}  }
 
-\author{\Large Mathew E. Hauer\vspace{0.05in} \newline\normalsize\emph{University of Georgia}   \and \Large Alexis R. Santos\vspace{0.05in} \newline\normalsize\emph{Pennsylvania State University}  }
+\author{\Large Mathew E.
+Hauer\textsuperscript{1}*\vspace{0.05in} \newline\normalsize\emph{University of Georgia}   \and \Large Alexis R.
+Santos\textsuperscript{2}\vspace{0.05in} \newline\normalsize\emph{Pennsylvania State University}  }
 
@@
             bookmarks=true,
-            pdfauthor={Mathew E. Hauer (University of Georgia) and Alexis R. Santos (Pennsylvania State University)},
+            pdfauthor={Mathew E. Hauer\textsuperscript{1}* (University of Georgia) and Alexis R. Santos\textsuperscript{2} (Pennsylvania State University)},
              pdfkeywords = {Climate change, Life expectancy, Mortality, Demography, Excess
@@
 \usepackage{rotating}
+\usepackage{fancyhdr}
+\usepackage[table]{xcolor}
+\usepackage{tabularx}
+\usepackage{makecell}
+\usepackage{xcolor}
+\pagestyle{fancy}
+\fancyhead{}
+\fancyhead[CE,CO]{Climate change life expectancy}
 
@@
    \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
-\textbf{\authorfont Mathew E. Hauer} \hskip 15pt \emph{\small University of Georgia}   \par \textbf{\authorfont Alexis R. Santos} \hskip 15pt \emph{\small Pennsylvania State University}
+\textbf{\authorfont Mathew E.
+Hauer\textsuperscript{1}*} \hskip 15pt \emph{\small University of Georgia}   \par \textbf{\authorfont Alexis R.
+Santos\textsuperscript{2}} \hskip 15pt \emph{\small Pennsylvania State University}
 
@@
 the dramatic impact on public health and human mortality from climate
-change. However, life expectancy at birth is more easily communicated to
-the public, understood by the public, and controls for different
-geographic, economic, and demographic groups. No studies have
-comprehensively connected excess mortality due to climate change to
-anticipated reductions in life expectancy. Without properly situating
-the potential loss of life within the contexts of life expectancy, we
-risk misrepresenting cost of climate change on mortality. In this paper,
-we convert excess mortality estimates into potential reductions of life
+change. However, life expectancy at birth is more easily communicated
+and understood by the public, and controls for different geographic,
+economic, and demographic groups. No studies have comprehensively
+connected excess mortality due to climate change to anticipated
+reductions in life expectancy. Without properly situating the potential
+loss of life within the contexts of life expectancy, we risk
+misrepresenting cost of climate change on mortality. In this paper, we
+convert excess mortality estimates into potential reductions of life
 expectancy at birth and report the cost of climate change on human
@@
 
-\noindent \doublespacing \section{Introduction}\label{introduction}
+\noindent \doublespacing \begin{itemize}
+\tightlist
+\item
+  Corresponding author.
+  \href{mailto:hauer@uga.edu}{\nolinkurl{hauer@uga.edu}}. p:
+  706-542-9369.
+\end{itemize}
+
+\textsuperscript{1} Carl Vinson Institute of Government, University of
+Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.
+
+\textsuperscript{2} Department of Sociology and Criminology,
+Pennsylvania State University. 901 Oswald Tower, University Park, PA USA
+16802.
+
+\newpage
+
+\section{Introduction}\label{introduction}
 
@@
 coastal floods, and windstorms under the business-as-usual greenhouse
-gas emission scenario. They modelled long-term, small-area demographic
-changes in European countries that correspond to a middle-of-the-road
-socioeconomic scenario and based their projected mortality on
-extrapolations of an exhaustive examination of contemporary disaster
-databases. Their results represent business-as-usual climate change and
-human development without incorporating potential adaptation. They found
-a potential 150,000+ climate change related fatalities per year by the
-mid 2080s with climate change contributing to 90\% of mortality rise as
-opposed to population changes. These data provide the anticipated
-fatalities but not the effect these deaths might have on life
-expectancy.
+gas emission scenario (SRES A1B). They modelled long-term, small-area
+demographic changes in European countries that correspond to a
+middle-of-the-road socioeconomic scenario (Shared-Socioeconomic Pathway
+2 with medium fertility, medium mortality, medium migration, and the
+Global Education Trend education scenario) and based their projected
+mortality on extrapolations of an exhaustive examination of contemporary
+disaster databases within spatially rectified, high-resolution gridded
+datasets. While the disaster data are not standardized across country,
+Forzeri et al. imputed data in incomplete time periods and countries.
+This imputation could mask spatial variability at the sub-national
+level, but here we use their country-aggregated results. Their results
+represent business-as-usual climate change and human development without
+incorporating potential adaptation. They found a potential 150,000+
+climate change related fatalities per year by the mid 2080s with climate
+change contributing to 90\% of mortality rise as opposed to population
+changes. These data provide the anticipated fatalities but not the
+effect these deaths might have on life expectancy.
 
@@
 
-We find that climate change could alter life expectancy by -0.22 years
-(-2.5 months) in the average European country (\textbf{Fig. 1}). This
-reduction is comparable to mortality due to influenza and pneumonia
-\citep{arias2013united} in the United States. Although the average
-European country could see a change of -0.22 years, several countries
-are likely to experience considerably greater reductions in life
-expectancy. Luxembourg could see a change of up to -2.2 years and the
-medium-variant of climate hazards could cost the average Spaniard 0.98
-years of life.
+We find that climate change could alter life expectancy by -0.23 years
+(-2.5 months (-0.1 - -0.39 years)) in the average European country
+(\textbf{Fig. 1}). Although the average European country could see a
+change of -0.22 years, several countries are likely to experience
+considerably greater reductions in life expectancy. Luxembourg could see
+a change of up to -2.2 years and the medium-variant of climate hazards
+could cost the average Spaniard 0.98 years of life.
+
+The overall reduction in \(e_0\) is comparable to mortality due to
+influenza and pneumonia \citep{arias2013united} in the United States.
+\textbf{Table 1} reports the ranking of Climate Change compared to the
+major causes of death (COD) in European countries using age-standardised
+death rates and the standard European population
+\citep{pace2013revision} for the year 2013. In three European countries
+(Spain, Luxembourg, Italy), Climate Change could become a top three COD.
+In Spain (high = 101.62), Luxembourg (high = 160.93), France (high =
+93.45), and Italy (high = 68.21) Climate Change could become the leading
+cause of death if Climate Change is more extreme than anticipated. Only
+all circulatory diseases combined and all cancers combined would cause
+more deaths.
 
@@
 \centering
-\includegraphics{MS-cclifeexpec_files/figure-latex/unnamed-chunk-1-1.pdf}
+\includegraphics{MS-cclifeexpec_files/figure-latex/unnamed-chunk-2-1.pdf}
 \caption{Change in life expectancy at birth (\(e_0\)) due to
@@
 expectancy, five European countries could see life expectancy changes in
-excess of -0.5 years (Spain, Luxembourg, France, Italy, and Ireland).
+excess of -0.5 years (Luxembourg, Spain, Italy, France, and Ireland).
 This group of countries could see life expectancy changes of more than
@@
 \centering
-\includegraphics{MS-cclifeexpec_files/figure-latex/unnamed-chunk-2-1.pdf}
+\includegraphics{MS-cclifeexpec_files/figure-latex/unnamed-chunk-3-1.pdf}
 \caption{Estimated reduction in life expectancy at birth (\(e_0\)) by
@@
 similar longitudes seem to experience more similar mortality regimes
-from climate change. Additionally, these results point to the importance
-of converting excess mortality into life expectancy to properly quantify
-the effects of climate change mortality.
+from climate change. Emergent research in Atmospheric Rivers suggest
+Western Europe could be more susceptible to these types of events
+\citep{ramos2015daily}. Additionally, these results point to the
+importance of converting excess mortality into life expectancy to
+properly quantify the effects of climate change mortality.
 
@@
 
+\begin{sidewaystable}
+\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Climate Change COD is based on the 2080s while the other CODs are based on the 2010s.}
+\rowcolors{2}{gray!25}{white}
+\small
+\begin{tabularx}{\textwidth}{l|XXXXXXXXX}
+  \hline
+   \rowcolor{white}
+ Country & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\ 
+  \hline
+Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
+of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\
+Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
+of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\
+Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
+of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\
+France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis.
+of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\
+Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
+of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\
+Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis.
+of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\
+Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02} & \makecell{Dis.
+of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\
+Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44} & \makecell{Dis.
+of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\
+Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis.
+of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\
+Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis.
+of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\
+   \hline
+\end{tabularx}
+\vspace{1ex}
+ \raggedright Data for other causes of death come from Eurostat (online data code: hlth_cd_asdr2). These results reflect the Standard European Population for 2013.
+\end{sidewaystable}
+
 Without adaptation measures, our results suggest climate change could
@@
 children born today are likely to still be alive by the end of the
-century and will be in the age groups most threatened by extreme heat -
-the biggest mortality risk associated with climate change
+century and will be in the age groups most threatened by extreme heat
+(65+) - the biggest mortality risk associated with climate change
 \citep{keatinge2000heat}. If climate change unfolds as a more aggressive
@@
 most complete and accurate source of mortality data available. We
-downloaded the most recent 1x5 (by age and year) life table data for
+downloaded the most recent 5x1 (by age and year) life table data for
 twenty-eight European countries. We used the following data elements
-from the five-year life table data in the HMD:
+from the five-year age group life table data in the HMD:
 
@@
 
+\begin{figure}
+\centering
+\includegraphics{MS-cclifeexpec_files/figure-latex/suppfig-1.pdf}
+\caption{\textbf{Mortality distribution of Heat-related mortality in 28
+European countries}. The thick black line is the average\label{suppfig}}
+\end{figure}
+
 \subsection{Methods}\label{methods-1}
@@
 We derived a number of additional variables to accomplish our analysis.
-First, we abridged the life table data from the HMD from 100+ to 80+ to
-conform to the GBD cause-specific mortality schedules.
+First, we abridged the life table data from the HMD from ages 100+ to
+80+ to conform to the GBD cause-specific mortality schedules.
 
@@
 deaths, \(D\), from each age group \(x\) in each country \(i\)
-(\(D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}\)).
+(\(_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}\)).
 
@@
 
-\[ _nm_{x,i,BASE} = _nD_{x,i}^{HMD} / _nP_{x,i}^{HMD} \]
-\[ _n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + \hat{D_i} \cdot _nt_{x,i} ) / _nP_{x,i}^{HMD} \]
+\begin{equation}
+_nm_{x,i,BASE} = _nD_{x,i}^{HMD} \,/\, _nP_{x,i}^{HMD}
+\end{equation}
+
+\begin{equation}
+_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_i} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
+\end{equation}
 
@@
 
-\[_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}\]
-\[_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}\]
+\begin{equation}
+_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}
+\end{equation}
+
+\begin{equation}
+_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}
+\end{equation}
 
 We calculated each additional life table value identically for each
-scenario \(s\) using the standard life table equations:
+scenario \(s\) using standard life table equations:
 
@@
 \singlespacing 
-\bibliography{C:/Users/Matt/Documents/Untitled/ccmort}
+\bibliography{LATEX/ccmort}
 \end{document}

### 2018-03-31 MS-cclifeexpec.Rmd
Update MS-cclifeexpec.Rmd

@@
 library(data.table)
+library(getPass)
 
@@
 
-myusername <- "hauer@uga.edu"
-mypassword <- "fuckhead"
+myusername <- getPass()
+mypassword <- getPass()
 

### 2018-04-05 supplementary_rcode.Rmd
minor changes in prep for submittal.

---
title: "Climate Change Projected to Reduce European Life Expectancy"
output:
  word_document: default
  pdf_document: default
date: "April 5, 2018"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## R Data Load and Processing The following code chunk will 
1. Import the underlying Forzieri et al data. 
2. Import the Standard European Population from EuroStat. 
3. Import the leading causes of death with age-standardized death rates from EuroStat. 
4. Import the Global Burden of Disease data. This data is located on github. 
5. Download all of the Human Mortality Database data.

To successfully run this code, several R pacakgse must be installed and these
packages are located just below this text section but are tidyverse, HMHFDplus,
tmap, tmaptools, RColorBrewer, data.table, and getPass.

To successfully run this code, you must also have an account at the HMD and
input your username and password when prompted.

```{r DataLoadAndProcessing}
rm(list=ls())
library(tidyverse)
library(HMDHFDplus)
library(tmap)
library(tmaptools)
library(RColorBrewer)
library(data.table)
library(getPass)

###   Inputting user info for HMD
myusername <- getPass(msg = "HMD username: ", noblank = FALSE, forcemask = TRUE)
mypassword <- getPass(msg = "HMD password: ", noblank = FALSE, forcemask = FALSE)

###   Importing the data from Forizeri et al.
lancetdata <- tribble(
    ~CNTRY,  ~MID,  ~LOW, ~HIGH,
     "AUT",   467,   121,   797,
     "BEL",  2841,    91,  5007,
     "BGR",    44,    36,    63,
     "CHE",  1981,   497,  3247,
     "CYP",   143,    62,   279,
     "CZE",   375,    48,   628,
  "DEUTNP",  7319,   620, 12132,
     "DNK",    24,     4,    56,
     "EST",    12,     0,    22,
     "ESP", 41326, 33495, 53802,
     "FIN",    23,     3,    44,
  "FRATNP", 38728, 14235, 72947,
     "GRC",  2838,  1520,  5606,
     "HRV",  1145,   624,  1811,
     "HUN",   559,   191,   985,
     "IRL",  1566,   476,  4195,
     "ISL",     3,     1,     5,
     "ITA", 41965, 23694, 57401,
     "LTU",    15,     1,    25,
     "LUX",   392,    12,   820,
     "LVA",    17,     1,    30,
     "MLT",     0,     0,     0,
     "NLD",  1984,   126,  4114,
     "NOR",    15,     4,    34,
     "POL",   313,    52,   561,
     "PRT",  4573,  3808,  6284,
     "ROU",   177,    97,   301,
     "SWE",    52,    10,   113,
     "SVN",   561,   251,   847,
     "SVK",    63,    17,   118,
  "GBR_NP",  2515,   388,  7484
  )

### European Standard Population from EuroStat
ec_agestandard <- tribble(
  ~Age, ~StnrdPop,
     0,  1118.427,
     1,  4338.143,
     5,  5207.188,
    10,   5378.67,
    15,   6095.53,
    20,  6646.578,
    25,  7054.462,
    30,  7211.366,
    35,  7249.137,
    40,  7288.966,
    45,  7207.381,
    50,  6904.728,
    55,  6400.144,
    60,  5798.191,
    65,  4660.589,
    70,   4031.35,
    75,  3292.724,
    80,  4116.422
  )

### COD standardized death rates from EuroStat
europe_standarddr <- tribble(
          ~COUNTRY, ~Circulatorydisease, ~Cancer, ~Heartdisease, ~LungCancer, ~Respiratorydiseases, ~Diseasesofthenervoussystem, ~Colorectalcancer, ~Suicide, ~TransportAccidents,
           "EU-28",               373.6,   261.5,         126.3,        54.4,                 78.3,                        38.6,              30.5,     11.3,                 5.8,
      "Luxembourg",               296.9,   260.7,          80.3,        59.6,                 63.8,                          38,              25.5,     13.4,                   6,
           "Spain",                 245,   232.7,          68.2,        47.8,                 91.7,                        48.5,              33.6,      8.2,                 4.3,
           "Italy",               310.1,   246.6,          98.3,        49.4,                 58.3,                        34.3,                27,      6.3,                 5.6,
          "France",               202.9,   245.4,          49.3,        50.1,                   52,                        50.2,              26.1,     14.1,                 5.1,
        "Portugal",               305.8,   242.1,          69.6,        36.4,                116.7,                        32.8,                35,     11.3,                 7.8,
         "Croatia",               678.6,   336.4,         306.5,        65.2,                 59.7,                        21.3,                51,     16.8,                 8.9,
         "Ireland",               309.9,   288.3,         147.5,        61.5,                125.9,                        48.7,              32.4,       11,                   4,
        "Slovenia",               451.3,   299.9,         102.8,        58.6,                 66.3,                        21.1,              38.4,     18.9,                 6.7,
     "Switzerland",                 280,   219.6,          97.8,        42.1,                 51.3,                        44.5,              22.8,     12.8,                 3.6,
         "Belgium",               281.9,   252.6,          72.4,        61.6,                 95.7,                        46.5,              26.1,     17.3,                 6.7,
          "Greece",               381.4,   249.3,           103,        61.9,                108.1,                        20.9,              23.3,        5,                 8.6,
     "Netherlands",               271.7,   282.2,          62.4,        66.7,                 74.1,                        48.3,              32.9,     11.1,                 4.1,
         "Germany",               403.5,   253.2,         142.8,          51,                   68,                        29.6,                29,     11.9,                 4.6,
   "Great Britain",               264.9,   278.4,         118.4,        61.4,                130.9,                        47.6,              27.7,      7.1,                 2.8,
         "Austria",               418.1,   249.3,         179.1,        47.5,                 46.6,                        32.6,              26.4,     15.3,                 5.8,
        "Bulgaria",                1131,   242.4,         195.4,        45.5,                 58.1,                        15.3,              34.9,      9.9,                   9,
  "Czech Republic",               615.2,   284.6,         333.1,        53.1,                 73.4,                        30.8,              37.9,     14.4,                78.8,
         "Denmark",               256.6,   300.6,            81,        71.7,                115.7,                        42.9,              35.2,     11.9,                   4,
         "Estonia",               699.6,   299.4,         295.5,        55.3,                 43.8,                        21.8,                36,     18.3,                 7.5,
         "Finland",               378.8,   218.6,         199.2,          39,                 34.4,                         155,              22.6,     14.6,                 5.7,
         "Hungary",               761.5,   348.1,         390.6,        89.8,                 78.6,                        19.9,                55,     19.4,                 8.1,
          "Latvia",               882.7,   299.3,         442.7,        46.9,                 35.9,                        15.6,              34.2,       19,                12.4,
       "Lithuania",               848.8,   276.2,         564.4,        46.1,                 42.1,                        20.8,              32.1,     31.5,                10.7,
          "Norway",               272.6,   252.5,          95.7,        50.5,                 88.4,                        45.4,              36.4,      7.3,                   4,
          "Poland",               591.4,   292.3,         129.1,        69.2,                 69.1,                        16.5,                36,     15.5,                10.3,
        "Slovakia",               654.6,   324.1,         388.8,          50,                 74.9,                        29.5,              49.2,     10.8,                 8.5,
          "Sweden",               338.3,   234.8,         131.2,        38.7,                 58.1,                        42.6,              29.2,     12.1,                 3.4,
          "Cyprus",               351.8,     201,         108.7,        37.2,                 86.2,                        26.8,              16.7,      4.5,                 6.5,
   "Liechtenstein",               296.4,     203,          73.7,        31.3,                 89.8,                        67.6,               6.8,     10.2,                10.3,
           "Malta",               372.4,   233.5,         202.8,        43.2,                 96.6,                          21,              28.3,      8.3,                 2.5,
         "Romania",               951.3,   273.2,         320.3,        54.2,                 78.4,                          21,              32.4,     11.4,                12.3,
          "Serbia",               931.6,   298.3,         159.5,        69.4,                 79.7,                        27.3,              37.2,     15.9,                 7.6
  )

countrycodes <- read.csv("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv") %>%
  rename(CNTRY = alpha.3,
         COUNTRY = name) %>%
  mutate(ISO3 = CNTRY) %>%
  select(COUNTRY, CNTRY, ISO3)
#  read.csv("data/LANcET_CCMORTDATA.csv",colClasses=c("CNTRY"="character"))

###   Getting a country list from the HMD
Countries <- getHMDcountries()

###   Downloading the DEATHS data from the HMD in 5-year age groups by single-year.
deaths <- rbindlist(lapply(Countries, function(CNTRY){
  Dat <- readHMDweb(CNTRY = CNTRY, item = "Deaths_5x1", fixup=TRUE, username = myusername, password = mypassword)
  Dat$CNTRY <- CNTRY
  Dat}))

###   Downloading the POPULATION data from the HMD.
pops <- rbindlist(lapply(Countries, function(CNTRY){
  Dat <- readHMDweb(CNTRY = CNTRY, item = "Population", fixup=TRUE, username = myusername, password = mypassword)
  Dat$CNTRY <- CNTRY
  Dat}))

###   Downloading the LIFE TABLE data from the HMD in 5-year age groups by single-year.
lt <- rbindlist(lapply(Countries, function(CNTRY){
  Dat <- readHMDweb(CNTRY = CNTRY, item = "bltper_5x1", fixup=TRUE, username = myusername, password = mypassword)
  Dat$CNTRY <- CNTRY
  Dat}))

###   Topcoding the age groups to 80+ of the LIFE TABLE data, and summing the ax value.
lt2 <- lt%>%
  mutate(Age = ifelse(Age >= 80, 80, Age)) %>%
  group_by(CNTRY, Year, Age)%>%
  summarise(ax = sum(ax))

###   Topcoding the age groups of the DEATHS data and summing.
deaths2 <- deaths %>%
  mutate(Age = ifelse(Age >=80, 80, Age)) %>%
  group_by(CNTRY, Year, Age) %>%
  summarise(deaths = sum(Total) )

###   Recoding the Population data to single year of age
pops2 <- pops %>%
  mutate(Age = case_when(
    Age >= 80 ~ 80,
    Age >= 75 ~ 75,
    Age >= 70 ~ 70,
    Age >= 65 ~ 65,
    Age >= 60 ~ 60,
    Age >= 55 ~ 55,
    Age >= 50 ~ 50,
    Age >= 45 ~ 45,
    Age >= 40 ~ 40,
    Age >= 35 ~ 35,
    Age >= 30 ~ 30,
    Age >= 25 ~ 25,
    Age >= 20 ~ 20,
    Age >= 15 ~ 15,
    Age >= 10 ~ 10,
    Age >= 5 ~5,
    Age >= 1 ~ 1,
    TRUE ~ as.numeric(Age)
  ))  %>%
  group_by(CNTRY, Year, Age) %>%
  summarise(Pop= sum(Total1))

###   Joining the POPULATION, DEATHS, and LIFE TABLE data and subseting to the maximum year in the data.
a <- left_join(pops2, deaths2)
a <- left_join(a, lt2)
a <- group_by(a, CNTRY, Age) %>%
  filter(Year == max(Year))

###   Importing the GBD data.
###   Data is 5-year age groups of mortality rates for Environmental heat and cold exposure for 2006-2015. 
###   Analysis uses the mean of the last 10 years of data for this death distribution.
GBD_data <- read.csv("data/GBD_data.csv") %>%
  #mutate(perdist = log(val)) %>%
  group_by(countrycode, Age) %>%
  summarise(meanval = mean(val),
            upperval = mean(upper),
            lowerval = mean(lower)) %>%
  group_by(countrycode) %>%
  mutate(countmeanval = sum(meanval),
         countupperval = sum(upperval),
         countlowerval = sum(lowerval),
         perdist = (meanval / countmeanval)) %>%
  rename(CNTRY = countrycode)

################################
#
# MAKING THE LIFE TABLES
#
################################
a <-  left_join(a, lancetdata)  %>%
  select(Age, Pop, CNTRY, deaths, ax, MID, LOW, HIGH) %>%
  filter(!is.na(MID))
a2 <- left_join(a, GBD_data) %>%
  mutate(width = ifelse(Age == 0, 1,
                        ifelse(Age == 1, 4,5)),
         mx_base = deaths/Pop,
         mx_low = (deaths + LOW * perdist)/Pop,
         mx_mid = (deaths + MID * perdist)/Pop,
         mx_high = (deaths + HIGH * perdist)/Pop,
         qx_base = ifelse(Age == 80, 1, mx_base/(1+((width-ax)*mx_base))),
         qx_low = ifelse(Age == 80, 1, mx_low/(1+((width-ax)*mx_low))),
         qx_mid = ifelse(Age == 80, 1, mx_mid/(1+((width-ax)*mx_mid))),
         qx_high = ifelse(Age == 80, 1, mx_high/(1+((width-ax)*mx_high)))) %>%
  group_by(CNTRY) %>%
  mutate(lx_base = ifelse(is.na(lag(cumprod(1-qx_base),1)*100000), 100000, lag(cumprod(1-qx_base),1)*100000),
         dx_base = qx_base*lx_base,
         Lx_base = ifelse(Age == 80, (lx_base/mx_base), ax * lx_base + (width-ax) * lead(lx_base,1)),
         Tx_base = rev(cumsum(rev(Lx_base))),
         ex_base = Tx_base/lx_base,
         lx_low = ifelse(is.na(lag(cumprod(1-qx_low),1)*100000), 100000, lag(cumprod(1-qx_low),1)*100000),
         dx_low = qx_low*lx_low,
         Lx_low = ifelse(Age == 80, (lx_low/mx_low), ax * lx_low + (width-ax) * lead(lx_low,1)),
         Tx_low = rev(cumsum(rev(Lx_low))),
         ex_low = Tx_low/lx_low,
         lx_mid = ifelse(is.na(lag(cumprod(1-qx_mid),1)*100000), 100000, lag(cumprod(1-qx_mid),1)*100000),
         dx_mid = qx_mid*lx_mid,
         Lx_mid = ifelse(Age == 80, (lx_mid/mx_mid), ax * lx_mid + (width-ax) * lead(lx_mid,1)),
         Tx_mid = rev(cumsum(rev(Lx_mid))),
         ex_mid = Tx_mid/lx_mid,
         lx_high = ifelse(is.na(lag(cumprod(1-qx_high),1)*100000), 100000, lag(cumprod(1-qx_high),1)*100000),
         dx_high = qx_high*lx_high,
         Lx_high = ifelse(Age == 80, (lx_high/mx_high), ax * lx_high + (width-ax) * lead(lx_high,1)),
         Tx_high = rev(cumsum(rev(Lx_high))),
         ex_high = Tx_high/lx_high,
         DIF_LOW = ex_low - ex_base,
         DIF_MID = ex_mid - ex_base,
         DIF_HIGH = ex_high - ex_base)

lt_lancet <- a2 %>%
  filter(Age == 0) %>%
  mutate(Base_e0 = ex_base,
         LOW_e0 = ex_low,
         MID_e0 = ex_mid,
         HIGH_e0 = ex_high) %>%
  dplyr::select(CNTRY, Base_e0, LOW_e0, MID_e0, HIGH_e0, DIF_LOW, DIF_MID, DIF_HIGH) %>%
  ungroup() %>%
  arrange(DIF_MID) %>%
  mutate(RANK = row_number(),
         dif_mid = round(DIF_MID, 2)) %>%
  left_join(., countrycodes) %>%
  mutate(COUNTRY = case_when(
    CNTRY == "FRATNP" ~ "France",
    CNTRY == "DEUTNP" ~ "Germany",
    CNTRY == "GBR_NP" ~ "Great Britain",
    TRUE ~ as.character(COUNTRY)),
    ISO3 = case_when(
      CNTRY == "FRATNP" ~ "FRA",
      CNTRY == "DEUTNP" ~ "DEU",
      CNTRY == "GBR_NP" ~ "GBR",
      TRUE ~ as.character(ISO3)),
    Test = ISO3)
```

# Figure 1
This chunk of code reproduces Figure 1.
```{r Figure1}

ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=dif_mid)) +
  geom_pointrange(stat="identity") +
  geom_point(stat="identity", fill="black", size=8) +
   geom_hline(yintercept=0) +
  geom_text(color="white", size=2) +
  #scale_x_discrete(limits = rev(levels(the_factor)))
  scale_x_reverse(breaks = lt_lancet$RANK, labels=lt_lancet$COUNTRY) +
  coord_flip() +
  theme_bw() +
  #ylim(-2, 2) +
  theme(panel.grid.minor = element_blank()) +

   labs(y=expression(Delta* e[0]),
        x= "Country")
```

# Figure 2
This chunk of code reproduecs Figure 2.
```{r Figure2}

data(Europe)
Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)

bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)

  tm_shape(Europe2) +
  tm_fill("DIF_MID", title = expression(Delta* e[0 ~italic(MID)]), palette = "YlOrBr", style = "jenks") +
  tm_borders(alpha = 0.5) +  
  tm_text("Test" , size="AREA", root=5) +
  tm_format_Europe() 
```

# Supplementay Figure 1
This chunk of code reproduces the Supplementary Figure
```{r suppfig}
GBD_data <- read.csv("data/GBD_data.csv") %>%
  #mutate(perdist = log(val)) %>%
  group_by(countrycode, Age) %>%
  summarise(meanval = mean(val),
            upperval = mean(upper),
            lowerval = mean(lower)) %>%
  group_by(countrycode) %>%
  mutate(countmeanval = sum(meanval),
         countupperval = sum(upperval),
         countlowerval = sum(lowerval),
         perdist = (meanval / countmeanval)) %>%
  rename(CNTRY = countrycode)

ggplot(data=GBD_data, aes(x=Age, y=perdist)) +
  geom_line(aes(color=CNTRY), alpha=1) +
  geom_smooth(span=0.3, se=F, col="black", lwd =2) +
  theme_bw() +
  theme(legend.position = c(0.8,0.25),
        legend.title=element_blank(),
        legend.text=element_text(size=4),
        legend.key.size = unit(1,"line")) +
  guides(col = guide_legend(ncol = 4)) +
  scale_y_log10()+
  labs(y="log(Mortality Percentage Distribution)",
       #title = "Mortality distribution of Heat-related mortality in 28 European countries",
       caption = "Source: Global Burden of Disease")

```

# Table 1
This chunk of code reproduces the data inside Table 1 in the main document.

Note: the output from this table (the object testtbl) is formatted for latex rather than markdown. To completely reproduce that table, the output from testtbl must be pasted into a latex compiler and all forward slashes ("/") must be converted to backward slashes ("\").
```{r Table1}

standard <- left_join(a2, ec_agestandard) %>%
  left_join(., countrycodes) %>%
  select(Age, CNTRY, COUNTRY, mx_base, mx_low, mx_mid, mx_high, StnrdPop) %>%
  mutate(ASDR_base = mx_base * StnrdPop,
         ASDR_low = mx_low * StnrdPop,
         ASDR_mid = mx_mid * StnrdPop,
         ASDR_high = mx_high * StnrdPop) %>%
  group_by(CNTRY, COUNTRY) %>%
  summarise(ASDR_base = sum(ASDR_base),
            ASDR_low = sum(ASDR_low),
            ASDR_mid = sum(ASDR_mid),
            ASDR_high = sum(ASDR_high)) %>%
  mutate(ASDR_low = ASDR_low - ASDR_base,
         ASDR_mid = ASDR_mid - ASDR_base,
         ASDR_high = ASDR_high - ASDR_base,
         COUNTRY = case_when(
           CNTRY == "FRATNP" ~ "France",
           CNTRY == "DEUTNP" ~ "Germany",
           CNTRY == "GBR_NP" ~ "Great Britain",
           TRUE ~ as.character(COUNTRY))) %>%
  left_join(., europe_standarddr) %>%
  ungroup() %>%
  select(-ASDR_base, -CNTRY)

selectcountries <- c("Spain", "Luxembourg", "Italy", "France", "Portugal", "Ireland", "Slovenia", "Croatia", "Switzerland", "Belgium")
testtbl<- standard %>%
  group_by(COUNTRY) %>%
  gather(COD, ASDR, 2:13) %>%
  filter(!COD %in% c("Circulatorydisease", "Cancer", "ASDR_high", "ASDR_low"),
         !COUNTRY == "Iceland") %>%
  mutate(RANK = rank(desc(ASDR)),
         ASDR = round(ASDR,2),
         COD = case_when(
           COD == "Heartdisease" ~ "makecell{Heart\\Disease\\",
           COD == "LungCancer" ~ "makecell{Lung\\Cancer\\",
           COD == "Respiratorydiseases" ~ "makecell{Respiratory\\Diseases\\",
           COD == "Diseasesofthenervoussystem" ~ "makecell{Dis. of the\\Nervous Sys\\",
           COD == "Colorectalcancer" ~ "makecell{Colorectal\\Cancer\\",
           COD == "Suicide" ~ "makecell{Suicide\\",
           COD == "TransportAccidents" ~ "makecell{Transport\\Accidents\\",
           COD == "ASDR_mid"  ~ "cellcolor{blue!25}makecell{cellcolor{blue!25}Climate\\cellcolor{blue!25}Change\\cellcolor{blue!25}"),
         Value = paste0(COD, " ", ASDR, "}")) %>%
  ungroup()%>%
  arrange(COD, desc(ASDR)) %>%
  #mutate(COUNTRY = paste0(row_number(),COUNTRY))%>%
  select(COUNTRY, Value, RANK) %>%
  spread(RANK, Value) %>%
  slice(match(selectcountries, COUNTRY))

head(testtbl)
```

### 2018-04-05 MS-cclifeexpec.Rmd
blinded.

--- 2018-03-23 MS-cclifeexpec.Rmd
@@
 author:
-- name: Mathew E. Hauer^1^*
-  affiliation: University of Georgia
-- name: Alexis R. Santos^2^
-  affiliation: Pennsylvania State University
+# - name: Mathew E. Hauer^1^*
+#   affiliation: University of Georgia
+# - name: Alexis R. Santos^2^
+#   affiliation: Pennsylvania State University
 
-abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change.
-However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups.
+abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change [@patz2005impact;@aastrom2013attributing;@forzieri2017increasing;@staddon2014climate].
+However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups [@parrish2010peer].
 No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy.
@@
 - \fancyhead{}
-- \fancyhead[CE,CO]{Climate change life expectancy}
+- \fancyhead[C]{Climate change life expectancy}
+
 ---
 
-* Corresponding author. hauer@uga.edu. p: 706-542-9369.
+<!-- * Corresponding author. hauer@uga.edu. p: 706-542-9369. -->
 
-^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.
+<!-- ^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602. -->
 
-^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.
+<!-- ^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802. -->
 
@@
 library(data.table)
+library(getPass)
 
 ###   Inputting user info for HMD
-
-myusername <- "hauer@uga.edu"
-mypassword <- "fuckhead"
+myusername <- getPass(msg = "HMD username: ", noblank = FALSE, forcemask = FALSE)
+mypassword <- getPass(msg = "HMD password: ", noblank = FALSE, forcemask = FALSE)
 
@@
 
-# Introduction
+# Main Text
 
-Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimatin], estimates of displacement [@nicholls2011sea;@hauer2016millions;@hauer2017migration], or human conflict [@barnett2007climate]  but have the potential to lead to the loss of human life [@forzieri2017increasing;@pachauri2014climate].
+Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimating], estimates of displacement [@rigaud2018groundswell], or human conflict [@barnett2007climate]  but have the potential to lead to the loss of human life [@forzieri2017increasing;@pachauri2014climate].
 As impact quantification studies move further from the physical sciences of climate change into the social sciences, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].
 
-Scholars have long tried to estimate the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer].
-Although such estimates are useful, the excess mortality estimates are rather sterile - one death is a tragedy, a million deaths a statistic - and difficult to relate to on an individual level.
-Life expectancy at birth, on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone.
-To date, no studies exist that comprehensively examine the potential reductions of life expectancy due to climate change.
+Scholars have long estimated the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer].
+Although such estimates are useful, the excess mortality estimates are rather sterile -- one death is a tragedy, a million deaths a statistic -- and difficult to relate to on an individual level.
+Life expectancy at birth ($e_0$), on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone.
+To our knowledge, no studies exist that comprehensively examine the potential reductions of life expectancy due to climate change.
 Without properly situating the potential loss of life within the contexts of metrics such as life expectancy, we risk underestimating the impact of climate change on human mortality.
 
-Using published data on excess mortality, we connect climate change excess mortality to life expectancy in a mortality model.
+Using published data on excess mortality[@forzieri2017increasing], we connect climate change excess mortality to life expectancy in a mortality model.
 By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in twenty-eight European countries.
-This approach allows us to quantify the impact of climate change on human longevity and answer the following question: How much could climate change mortality reduce life expectancy at birth?
+This approach allows us to quantify the impact of climate change on human longevity and answer the following questions: What is the cost of inaction on climate change on human longevity?
+and How does climate change related mortality compare to other mortality vectors?
 Our results can situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.
 
-# Methods
-We estimate changes in life expectancy by combining three primary datasets: excess mortality data from Forzieri et al.
+<!-- # Methods -->
+We estimate changes in life expectancy by combining three primary datasets: excess mortality data due to meterological hazards related to climate change from Forzieri et al.
 [@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD].
@@
 [@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario (SRES A1B).
-They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario (Shared-Socioeconomic Pathway 2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario) and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases within spatially rectified, high-resolution gridded datasets.
+They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario (SSP2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario) and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases within spatially rectified, high-resolution gridded datasets.
 While the disaster data are not standardized across country, Forzeri et al.
@@
 The HMD comprises only complete, official vital event statistics and represents the most accurate and complete compilation of mortality data available.
-We allocate excess multi-hazard mortality for each country from supplementary table S8 [@forzieri2017increasing] based on the observed mortality schedule for environmental heat and cold exposure deaths from the GBD for each country in the previous decade.
+We allocate excess multi-hazard mortality for each country from supplementary table S8 in Forzieri et al.
+[@forzieri2017increasing] based on the observed mortality schedule for environmental heat and cold exposure deaths from the GBD for each country in the previous decade.
 
-We then recalculated age-specific mortality rates in five-year life tables by adding the climate change mortality onto the underlying mortality data and create four subsequent life tables (baseline, low, mid, high) using standard life table techniques [@wunsch2013life] based on the published confidence intervals from [@forzieri2017increasing].
+We then recalculated age-specific mortality rates in five-year life tables by adding the climate change mortality onto the underlying mortality data and create four subsequent life tables ($baseline$, $low$, $mid$, $high$) using standard life table techniques [@wunsch2013life] based on the published confidence intervals from [@forzieri2017increasing].
 We used a variant of the cause-deleted life tables approach [@brand2005approximations;@beltran2008integrated] to measure the impact of excess mortality on life expectancy at birth or $e_0$.
@@
 
-# Results
-We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig.
-1**).
+<!-- # Results -->
+We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**).
 This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States.
-Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy.
-Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.
+Although the average European country could see a change of `r round(mean(lt_lancet$DIF_MID),2)` years, several countries are likely to experience considerably greater reductions in life expectancy.
+Luxembourg could see a reduction of up to `r round(lt_lancet$DIF_HIGH[which(lt_lancet$CNTRY == "LUX")],2)*-1` years and the medium-variant of climate hazards could cost the average Spaniard `r round(lt_lancet$DIF_MID[which(lt_lancet$CNTRY == "ESP")],2)*-1` years of life.
 
-```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change.
+```{r figure1, echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change.
 We report changes in life expectancies due to climate change for twenty-eight European countries.
-The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}
+The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.\\label{figure1}")}
 
@@
 
-```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}
+```{r figure2, echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.\\label{map}")}
 
@@
 
-While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig.
-2**).
+**\autoref{Table1}** reports the rankings of Age Standardised Death Rates (ASDR) for the leading causes of death in these European countries (ASDR's for other CODs come from Eurostat (online data code: hlth_cd_asdr2)) with our estimates of climate change included.
+In many European countries, climate change could emerge as a top five killer.
+Only combined CODs, such as all circulatory diseases or all cancers, exhibit higher ASDRs if climatic hazards are more severe than anticipated for Spain, Luxembourg, and France.
+
+While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**\autoref{map}**).
 The most westerly European countries tend to have the greatest reductions.
@@
 
-# Discussion
+<!-- # Discussion -->
 In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries.
-Previous studies on climate change and excess mortality potentially misrepresent the impact climate change could have on human mortality.
+Previous studies on climate change and excess mortality potentially miscommunicate the impact climate change could have on human mortality.
 Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy].
@@
   mutate(RANK = rank(desc(ASDR)),
+         ASDR = round(ASDR,2),
          COD = case_when(
@@
            COD == "ASDR_mid"  ~ "cellcolor{blue!25}makecell{cellcolor{blue!25}Climate\\cellcolor{blue!25}Change\\cellcolor{blue!25}"),
-         Value = paste0(COD, " ", round(ASDR,2), "}")) %>%
+         Value = paste0(COD, " ", ASDR, "}")) %>%
   ungroup()%>%
@@
 \begin{sidewaystable}
-\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} 
+\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1}
 \rowcolors{2}{gray!25}{white}
@@
   \hline
-Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
+Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05\\\cellcolor{blue!25}\begin{tiny}(63.26 - 101.6)\end{tiny}} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
 of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\
-Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
+Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93\\\cellcolor{blue!25}\begin{tiny}(2.36 - 160.9)\end{tiny}} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
 of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\
-Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
+Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87\\\cellcolor{blue!25}\begin{tiny}(28.15 - 68.2)\end{tiny}} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
 of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\
 France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis.
-of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\
-Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
+of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61\\\cellcolor{blue!25}\begin{tiny}(18.24 - 93.5)\end{tiny}} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\
+Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95\\\cellcolor{blue!25}\begin{tiny}(29.93 - 49.4)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
 of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\
 Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis.
-of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\
-Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02} & \makecell{Dis.
+of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65\\\cellcolor{blue!25}\begin{tiny}(13.57 - 119.6)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\
+Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02\\\cellcolor{blue!25}\begin{tiny}(11.19 - 37.8)\end{tiny}} & \makecell{Dis.
 of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\
-Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44} & \makecell{Dis.
+Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44\\\cellcolor{blue!25}\begin{tiny}(13.32 - 38.6)\end{tiny}} & \makecell{Dis.
 of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\
 Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis.
-of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\
+of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88\\\cellcolor{blue!25}\begin{tiny}(5.74 - 37.5)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\
 Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis.
-of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\
+of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23\\\cellcolor{blue!25}\begin{tiny}(0.68 - 37.4)\end{tiny}} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\
    \hline
@@
 Without adaptation measures, our results suggest climate change could emerge as a significant new mortality vector and could pose a major public health threat for some European countries by the end of the century, echoing previous findings [@forzieri2017increasing;@patz2005impact].
-Life expectancy has steadily risen across the world for the last century [@oeppen2002broken] and our results suggest that climate change alone could spur a sharp reversal in these trends.
-We expect one European countries to see life expectancy reductions more than one year under the middle scenario (Luxembourg), but if climate change has a greater impact on mortality than anticipated, five European countries (Luxembourg, Spain, Italy, France, Ireland) could see life expectancy reductions more than one year with Luxembourg experiencing a reduction of more than two.
+Life expectancy has steadily risen across the world for the last century [@gerland2014world] and our results suggest that climate change alone could spur a sharp reversal in these trends in some countries.
+We expect two European countries to see life expectancy reductions more than one year under the middle scenario (Luxembourg and Spain), but if climate change has a greater impact on mortality than anticipated, five European countries (Luxembourg, Spain, Italy, France, Ireland) could see life expectancy reductions more than one year with Luxembourg experiencing a reduction of more than two.
+These findings highlight the hyperlocalized impacts of climate change [@kendon2014heavier;@rosenzweig2010cities;@forzieri2017increasing]
 
 Reductions such as those should not be taken lightly.
-Many of the children born today are likely to still be alive by the end of the century and will be in the age groups most threatened by extreme heat (65+) - the biggest mortality risk associated with climate change [@keatinge2000heat].
-If climate change unfolds as a more aggressive mortality vector, only major cardiovascular diseases and malignant neoplasms would contribute more to life expectancy reductions in numerous European countries [@arias2013united].
+Many of the children born today are likely to still be alive by the end of the century and will be in the age groups (aged 65+) most threatened by the biggest mortality risk associated with climate change [@keatinge2000heat] -- extreme heat.
+If climate change unfolds as a more aggressive mortality vector, only all circulatory diseases combined or all cancers combined would contribute more to mortality rates in numerous European countries.
 This would make climate change one of the most aggressive new mortality vectors to emerge over the last quarter-century, representing a major threat to public health in many parts of Europe.
@@
 
-# Supplemenatary Information
+# Methods
 ## Data

### 2018-04-05 MS-cclifeexpec.Rmd

@@
     fig_caption: true
-    latex_engine: pdflatex
+    latex_engine: xelatex
     template: LATEX/svm-latex-ms.tex
@@
 biblio-style: nature
+# use apsr or nature
 fig_caption: true
-# use apsr or nature
 header-includes:
@@
 ###   Inputting user info for HMD
+
 myusername <- getPass(msg = "HMD username: ", noblank = FALSE, forcemask = FALSE)
@@
 <!-- # Results -->
-We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**).
+We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` to `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**).
 This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States.
@@
 
+<!-- # Results -->
+We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig.
+1**).
+Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy.
+Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.
+
+The overall reduction in $e_0$ is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States.
+**Table 1** reports the ranking of Climate Change compared to the major causes of death (COD) in European countries using age-standardised death rates and the standard European population [@pace2013revision] for the year 2013.
+In three European countries (Spain, Luxembourg, Italy), Climate Change could become a top three COD.
+In Spain (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Spain")],2)`), Luxembourg (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Luxembourg")],2)`), France (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "France")],2)`), and Italy (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Italy")],2)`) Climate Change could become the leading cause of death if Climate Change is more extreme than anticipated.
+Only all circulatory diseases combined and all cancers combined would cause more deaths.
+
+```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change.
+We report changes in life expectancies due to climate change for twenty-eight European countries.
+The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}
+
+ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=dif_mid)) +
+  geom_pointrange(stat="identity") +
+  geom_point(stat="identity", fill="black", size=8) +
+   geom_hline(yintercept=0) +
+  geom_text(color="white", size=2) +
+  #scale_x_discrete(limits = rev(levels(the_factor)))
+  scale_x_reverse(breaks = lt_lancet$RANK, labels=lt_lancet$COUNTRY) +
+  coord_flip() +
+  theme_bw() +
+  #ylim(-2, 2) +
+  theme(panel.grid.minor = element_blank()) +
+
+   labs(y=expression(Delta* e[0]),
+        x= "Country")
+```
+
+Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies.
+Whereas many Northern European countries could experience negligible impacts on life expectancy, five European countries could see life expectancy changes in excess of -0.5 years (Luxembourg, Spain, Italy, France, and Ireland).
+This group of countries could see life expectancy changes of more than -1.0 years if climate change hazards are more intense than anticipated.
+In some countries, climate change could thus become a bigger killer than trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or all accident related mortality (-0.84 years) [@arias2013united] by the end of the century.
+
+```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}
+
+data(Europe)
+Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)
+
+bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)
+
+  tm_shape(Europe2) +
+  tm_fill("DIF_MID", title = expression(Delta* e[0 ~italic(MID)]), palette = "YlOrBr", style = "jenks") +
+  tm_borders(alpha = 0.5) +  
+  tm_text("Test" , size="AREA", root=5) +
+  tm_format_Europe() 
+```
+
+While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig.
+2**).
+The most westerly European countries tend to have the greatest reductions.
+Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change.
+Emergent research in Atmospheric Rivers suggest Western Europe could be more susceptible to these types of events [@ramos2015daily].
+Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.
+
+<!-- # Discussion -->
+In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries.
+Previous studies on climate change and excess mortality potentially misrepresent the impact climate change could have on human mortality.
+Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy].
+Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer].
+Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes.
+Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].
+
 \begin{sidewaystable}
+
 \caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1}
@@
 \end{tabularx}
+
 \end{sidewaystable}

### 2018-04-05 MS-cclifeexpec.tex

@@
 
-\author{\Large Mathew E.
-Hauer\textsuperscript{1}*\vspace{0.05in} \newline\normalsize\emph{University of Georgia}   \and \Large Alexis R.
-Santos\textsuperscript{2}\vspace{0.05in} \newline\normalsize\emph{Pennsylvania State University}  }
+\author{}
 
@@
             bookmarks=true,
-            pdfauthor={Mathew E. Hauer\textsuperscript{1}* (University of Georgia) and Alexis R. Santos\textsuperscript{2} (Pennsylvania State University)},
+            pdfauthor={},
              pdfkeywords = {Climate change, Life expectancy, Mortality, Demography, Excess
@@
 \fancyhead{}
-\fancyhead[CE,CO]{Climate change life expectancy}
+\fancyhead[C]{Climate change life expectancy}
 
@@
    \vskip 13.5pt\relax \normalsize\fontsize{11}{12}
-\textbf{\authorfont Mathew E.
-Hauer\textsuperscript{1}*} \hskip 15pt \emph{\small University of Georgia}   \par \textbf{\authorfont Alexis R.
-Santos\textsuperscript{2}} \hskip 15pt \emph{\small Pennsylvania State University}
 
@@
 the dramatic impact on public health and human mortality from climate
-change. However, life expectancy at birth is more easily communicated
-and understood by the public, and controls for different geographic,
-economic, and demographic groups. No studies have comprehensively
-connected excess mortality due to climate change to anticipated
-reductions in life expectancy. Without properly situating the potential
-loss of life within the contexts of life expectancy, we risk
-misrepresenting cost of climate change on mortality. In this paper, we
-convert excess mortality estimates into potential reductions of life
+change
+\citep{patz2005impact, aastrom2013attributing, forzieri2017increasing, staddon2014climate}.
+However, life expectancy at birth is more easily communicated and
+understood by the public, and controls for different geographic,
+economic, and demographic groups \citep{parrish2010peer}. No studies
+have comprehensively connected excess mortality due to climate change to
+anticipated reductions in life expectancy. Without properly situating
+the potential loss of life within the contexts of life expectancy, we
+risk misrepresenting cost of climate change on mortality. In this paper,
+we convert excess mortality estimates into potential reductions of life
 expectancy at birth and report the cost of climate change on human
@@
 
-\noindent \doublespacing \begin{itemize}
-\tightlist
-\item
-  Corresponding author.
-  \href{mailto:hauer@uga.edu}{\nolinkurl{hauer@uga.edu}}. p:
-  706-542-9369.
-\end{itemize}
+\noindent \doublespacing \newpage
 
-\textsuperscript{1} Carl Vinson Institute of Government, University of
-Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.
-
-\textsuperscript{2} Department of Sociology and Criminology,
-Pennsylvania State University. 901 Oswald Tower, University Park, PA USA
-16802.
-
-\newpage
-
-\section{Introduction}\label{introduction}
+\section{Main Text}\label{main-text}
 
 Climate change's implications on humanity go far beyond estimates of
-economic damages \citep{hsiang2017estimatin}, estimates of displacement
-\citep{nicholls2011sea, hauer2016millions, hauer2017migration}, or human
-conflict \citep{barnett2007climate} but have the potential to lead to
-the loss of human life
-\citep{forzieri2017increasing, pachauri2014climate}. As impact
-quantification studies move further from the physical sciences of
+economic damages \citep{hsiang2017estimating}, estimates of displacement
+\citep{rigaud2018groundswell}, or human conflict
+\citep{barnett2007climate} but have the potential to lead to the loss of
+human life \citep{forzieri2017increasing, pachauri2014climate}. As
+impact quantification studies move further from the physical sciences of
 climate change into the social sciences, properly quantifying and
@@
 
-Scholars have long tried to estimate the mortality risks associated with
-climate change and typically use excess or extra mortality
+Scholars have long estimated the mortality risks associated with climate
+change and typically use excess or extra mortality
 \citep{forzieri2017increasing, wilson2017climate, mcmichael2006climate, zanobetti2012summer}.
 Although such estimates are useful, the excess mortality estimates are
-rather sterile - one death is a tragedy, a million deaths a statistic -
-and difficult to relate to on an individual level. Life expectancy at
-birth, on the other hand, provides potent comparisons of mortality
-vectors and converts excess mortality into an intuitively understood
-metric, relatable to everyone. To date, no studies exist that
-comprehensively examine the potential reductions of life expectancy due
-to climate change. Without properly situating the potential loss of life
-within the contexts of metrics such as life expectancy, we risk
-underestimating the impact of climate change on human mortality.
+rather sterile -- one death is a tragedy, a million deaths a statistic
+-- and difficult to relate to on an individual level. Life expectancy at
+birth (\(e_0\)), on the other hand, provides potent comparisons of
+mortality vectors and converts excess mortality into an intuitively
+understood metric, relatable to everyone. To our knowledge, no studies
+exist that comprehensively examine the potential reductions of life
+expectancy due to climate change. Without properly situating the
+potential loss of life within the contexts of metrics such as life
+expectancy, we risk underestimating the impact of climate change on
+human mortality.
 
-Using published data on excess mortality, we connect climate change
-excess mortality to life expectancy in a mortality model. By estimating
-the increase in age-specific mortality rates associated with previously
-published excess death estimates, we can assess how much climate change
-could reduce the anticipated longevity of the average person in
-twenty-eight European countries. This approach allows us to quantify the
-impact of climate change on human longevity and answer the following
-question: How much could climate change mortality reduce life expectancy
-at birth? Our results can situate climate change mortality within the
-broader context of human mortality and can be used to inform public
-health interventions to prevent such futures.
-
-\section{Methods}\label{methods}
+Using published data on excess mortality\citep{forzieri2017increasing},
+we connect climate change excess mortality to life expectancy in a
+mortality model. By estimating the increase in age-specific mortality
+rates associated with previously published excess death estimates, we
+can assess how much climate change could reduce the anticipated
+longevity of the average person in twenty-eight European countries. This
+approach allows us to quantify the impact of climate change on human
+longevity and answer the following questions: What is the cost of
+inaction on climate change on human longevity? and How does climate
+change related mortality compare to other mortality vectors? Our results
+can situate climate change mortality within the broader context of human
+mortality and can be used to inform public health interventions to
+prevent such futures.
 
 We estimate changes in life expectancy by combining three primary
-datasets: excess mortality data from Forzieri et al.
-\citep{forzieri2017increasing}, cause-of-death distributions from the
-Global Burden of Disease study (GBD) \citep{GBD, wang2012age}, and life
-tables from the Human Mortality Database (HMD) \citep{HMD}. Forzieri et
-al. \citep{forzieri2017increasing} focused on the projected excess
-mortality of heatwaves, coldwaves, wildfires, droughts, river and
-coastal floods, and windstorms under the business-as-usual greenhouse
-gas emission scenario (SRES A1B). They modelled long-term, small-area
-demographic changes in European countries that correspond to a
-middle-of-the-road socioeconomic scenario (Shared-Socioeconomic Pathway
-2 with medium fertility, medium mortality, medium migration, and the
-Global Education Trend education scenario) and based their projected
-mortality on extrapolations of an exhaustive examination of contemporary
-disaster databases within spatially rectified, high-resolution gridded
-datasets. While the disaster data are not standardized across country,
-Forzeri et al. imputed data in incomplete time periods and countries.
-This imputation could mask spatial variability at the sub-national
-level, but here we use their country-aggregated results. Their results
-represent business-as-usual climate change and human development without
-incorporating potential adaptation. They found a potential 150,000+
-climate change related fatalities per year by the mid 2080s with climate
-change contributing to 90\% of mortality rise as opposed to population
-changes. These data provide the anticipated fatalities but not the
-effect these deaths might have on life expectancy.
+datasets: excess mortality data due to meterological hazards related to
+climate change from Forzieri et al. \citep{forzieri2017increasing},
+cause-of-death distributions from the Global Burden of Disease study
+(GBD) \citep{GBD, wang2012age}, and life tables from the Human Mortality
+Database (HMD) \citep{HMD}. Forzieri et al.
+\citep{forzieri2017increasing} focused on the projected excess mortality
+of heatwaves, coldwaves, wildfires, droughts, river and coastal floods,
+and windstorms under the business-as-usual greenhouse gas emission
+scenario (SRES A1B). They modelled long-term, small-area demographic
+changes in European countries that correspond to a middle-of-the-road
+socioeconomic scenario (SSP2 with medium fertility, medium mortality,
+medium migration, and the Global Education Trend education scenario) and
+based their projected mortality on extrapolations of an exhaustive
+examination of contemporary disaster databases within spatially
+rectified, high-resolution gridded datasets. While the disaster data are
+not standardized across country, Forzeri et al. imputed data in
+incomplete time periods and countries. This imputation could mask
+spatial variability at the sub-national level, but here we use their
+country-aggregated results. Their results represent business-as-usual
+climate change and human development without incorporating potential
+adaptation. They found a potential 150,000+ climate change related
+fatalities per year by the mid 2080s with climate change contributing to
+90\% of mortality rise as opposed to population changes. These data
+provide the anticipated fatalities but not the effect these deaths might
+have on life expectancy.
 
@@
 complete compilation of mortality data available. We allocate excess
-multi-hazard mortality for each country from supplementary table S8
-\citep{forzieri2017increasing} based on the observed mortality schedule
-for environmental heat and cold exposure deaths from the GBD for each
-country in the previous decade.
+multi-hazard mortality for each country from supplementary table S8 in
+Forzieri et al. \citep{forzieri2017increasing} based on the observed
+mortality schedule for environmental heat and cold exposure deaths from
+the GBD for each country in the previous decade.
 
@@
 tables by adding the climate change mortality onto the underlying
-mortality data and create four subsequent life tables (baseline, low,
-mid, high) using standard life table techniques \citep{wunsch2013life}
-based on the published confidence intervals from
+mortality data and create four subsequent life tables (\(baseline\),
+\(low\), \(mid\), \(high\)) using standard life table techniques
+\citep{wunsch2013life} based on the published confidence intervals from
 \citep{forzieri2017increasing}. We used a variant of the cause-deleted
@@
 
-\section{Results}\label{results}
+We find that climate change could alter life expectancy by -0.23 years
+(-2.5 months (-0.1 to -0.39 years)) in the average European country
+(\textbf{\autoref{figure1}}). This reduction is comparable to mortality
+due to influenza and pneumonia \citep{arias2013united} in the United
+States. Although the average European country could see a change of
+-0.23 years, several countries are likely to experience considerably
+greater reductions in life expectancy. Luxembourg could see a reduction
+of up to 2.2 years and the medium-variant of climate hazards could cost
+the average Spaniard 1.03 years of life.
+
+\begin{figure}
+\centering
+\includegraphics{MS-cclifeexpec_files/figure-latex/figure1-1.pdf}
+\caption{Change in life expectancy at birth (\(e_0\)) due to
+business-as-usual climate change. We report changes in life expectancies
+due to climate change for twenty-eight European countries. The central
+values represent the ensemble median while the stems represent the upper
+and lower bounds of the inter-model climate variability.\label{figure1}}
+\end{figure}
+
+Our results also suggest climate change mortality differentials are
+likely to unfold along highly uneven geographies. Whereas many Northern
+European countries could experience negligible impacts on life
+expectancy, five European countries could see life expectancy changes in
+excess of -0.5 years (Spain, Luxembourg, France, Italy, and Ireland).
+This group of countries could see life expectancy changes of more than
+-1.0 years if climate change hazards are more intense than anticipated.
+In some countries, climate change could thus become a bigger killer than
+trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute
+myocardial infarction (-0.87 years), or all accident related mortality
+(-0.84 years) \citep{arias2013united} by the end of the century.
+
+\begin{figure}
+\centering
+\includegraphics{MS-cclifeexpec_files/figure-latex/figure2-1.pdf}
+\caption{Estimated reduction in life expectancy at birth (\(e_0\)) by
+the 2080s under the \(MID\) scenario.\label{map}}
+\end{figure}
+
+\textbf{\autoref{Table1}} reports the rankings of Age Standardised Death
+Rates (ASDR) for the leading causes of death in these European countries
+(ASDR's for other CODs come from Eurostat (online data code:
+hlth\_cd\_asdr2)) with our estimates of climate change included. In many
+European countries, climate change could emerge as a top five killer.
+Only combined CODs, such as all circulatory diseases or all cancers,
+exhibit higher ASDRs if climatic hazards are more severe than
+anticipated for Spain, Luxembourg, and France.
+
+While Forzieri et al's \citep{forzieri2017increasing} results suggest
+that the greatest climate change related mortality will unfold along a
+north-south gradient, we find the greatest reduction in life expectancy
+unfolds along an east-west gradient (\textbf{\autoref{map}}). The most
+westerly European countries tend to have the greatest reductions.
+Neighboring pairs of countries at similar latitudes could experience
+vastly different mortality regimes, but neighboring pairs of countries
+at similar longitudes seem to experience more similar mortality regimes
+from climate change. Emergent research in Atmospheric Rivers suggest
+Western Europe could be more susceptible to these types of events
+\citep{ramos2015daily}. Additionally, these results point to the
+importance of converting excess mortality into life expectancy to
+properly quantify the effects of climate change mortality.
+
+In this article, we demonstrate the impact climate change could have on
+life expectancy at birth in twenty-eight European countries. Previous
+studies on climate change and excess mortality potentially
+miscommunicate the impact climate change could have on human mortality.
+Contrary to excess mortality estimates, life expectancy is routinely
+used as a primary metric for communicating overall health outcomes and
+enjoys widespread use by major international organizations
+\citep{world2015world, marmot2012building, salomon2012healthy}. Life
+expectancy and its derivatives are the recommended metrics for
+population health \citep{parrish2010peer}. Additionally, it connects
+mortality estimates into intuitively understandable metrics, translating
+global estimates of mortality into individual outcomes. Our work reveals
+the extent to which climate change could reduce the average person's
+longevity by the end of the century, the first study of its kind,
+significantly expanding our understanding of climate change and public
+health; thus linking two of the major areas in current developmental and
+sustainability discussion at both national and international levels
+\citep{abel2016meeting}.
 
@@
 
-\section{Discussion}\label{discussion}
-
 In this article, we demonstrate the impact climate change could have on
@@
 \begin{sidewaystable}
-\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Climate Change COD is based on the 2080s while the other CODs are based on the 2010s.}
+
+\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1}
 \rowcolors{2}{gray!25}{white}
@@
   \hline
-Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
+Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05\\\cellcolor{blue!25}\begin{tiny}(63.26 - 101.6)\end{tiny}} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
 of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\
-Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
+Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93\\\cellcolor{blue!25}\begin{tiny}(2.36 - 160.9)\end{tiny}} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
 of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\
-Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
+Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87\\\cellcolor{blue!25}\begin{tiny}(28.15 - 68.2)\end{tiny}} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
 of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\
 France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis.
-of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\
-Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
+of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61\\\cellcolor{blue!25}\begin{tiny}(18.24 - 93.5)\end{tiny}} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\
+Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95\\\cellcolor{blue!25}\begin{tiny}(29.93 - 49.4)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
 of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\
 Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis.
-of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\
-Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02} & \makecell{Dis.
+of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65\\\cellcolor{blue!25}\begin{tiny}(13.57 - 119.6)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\
+Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02\\\cellcolor{blue!25}\begin{tiny}(11.19 - 37.8)\end{tiny}} & \makecell{Dis.
 of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\
-Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44} & \makecell{Dis.
+Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44\\\cellcolor{blue!25}\begin{tiny}(13.32 - 38.6)\end{tiny}} & \makecell{Dis.
 of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\
 Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis.
-of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\
+of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88\\\cellcolor{blue!25}\begin{tiny}(5.74 - 37.5)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\
 Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis.
-of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\
+of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23\\\cellcolor{blue!25}\begin{tiny}(0.68 - 37.4)\end{tiny}} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\
    \hline
 \end{tabularx}
-\vspace{1ex}
- \raggedright Data for other causes of death come from Eurostat (online data code: hlth_cd_asdr2). These results reflect the Standard European Population for 2013.
+
 \end{sidewaystable}
@@
 steadily risen across the world for the last century
-\citep{oeppen2002broken} and our results suggest that climate change
-alone could spur a sharp reversal in these trends. We expect one
-European countries to see life expectancy reductions more than one year
-under the middle scenario (Luxembourg), but if climate change has a
-greater impact on mortality than anticipated, five European countries
-(Luxembourg, Spain, Italy, France, Ireland) could see life expectancy
-reductions more than one year with Luxembourg experiencing a reduction
-of more than two.
+\citep{gerland2014world} and our results suggest that climate change
+alone could spur a sharp reversal in these trends in some countries. We
+expect two European countries to see life expectancy reductions more
+than one year under the middle scenario (Luxembourg and Spain), but if
+climate change has a greater impact on mortality than anticipated, five
+European countries (Luxembourg, Spain, Italy, France, Ireland) could see
+life expectancy reductions more than one year with Luxembourg
+experiencing a reduction of more than two. These findings highlight the
+hyperlocalized impacts of climate change
+\citep{kendon2014heavier, rosenzweig2010cities, forzieri2017increasing}
 
@@
 children born today are likely to still be alive by the end of the
-century and will be in the age groups most threatened by extreme heat
-(65+) - the biggest mortality risk associated with climate change
-\citep{keatinge2000heat}. If climate change unfolds as a more aggressive
-mortality vector, only major cardiovascular diseases and malignant
-neoplasms would contribute more to life expectancy reductions in
-numerous European countries \citep{arias2013united}. This would make
-climate change one of the most aggressive new mortality vectors to
-emerge over the last quarter-century, representing a major threat to
-public health in many parts of Europe.
+century and will be in the age groups (aged 65+) most threatened by the
+biggest mortality risk associated with climate change
+\citep{keatinge2000heat} -- extreme heat. If climate change unfolds as a
+more aggressive mortality vector, only all circulatory diseases combined
+or all cancers combined would contribute more to mortality rates in
+numerous European countries. This would make climate change one of the
+most aggressive new mortality vectors to emerge over the last
+quarter-century, representing a major threat to public health in many
+parts of Europe.
 
@@
 
-\section{Supplemenatary Information}\label{supplemenatary-information}
+\section{Methods}\label{methods}
 
@@
 \singlespacing 
-\bibliography{LATEX/ccmort}
+\bibliography{LATEX/ccmort.bib}
 \end{document}

### 2018-04-18 PRESENTATIONS/presentation.Rmd
paa presentation

--- 2018-03-31 MS-cclifeexpec.Rmd
@@
 ---
-output: 
-  pdf_document:
-    citation_package: natbib
-    keep_tex: true
-    fig_caption: true
-    latex_engine: pdflatex
-    template: LATEX/svm-latex-ms.tex
 title: "Climate change projected to reduce European life expectancy"
-thanks: "All data and code that supports these conclusions are available as supplementary materials."
-author:
-- name: Mathew E. Hauer^1^*
-  affiliation: University of Georgia
-- name: Alexis R. Santos^2^
-  affiliation: Pennsylvania State University
+author: | 
+ | Mathew E. Hauer^1^ & Alexis R. Santos^2^
+ | ^1^University of Georgia | ^2^Pennsylvania State University
+date: "April 26, 2018"
+output:   beamer_presentation
+classoption: "aspectratio=169"
 
-abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change.
-However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups.
-No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy.
-Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting cost of climate change on mortality.
-In this paper, we convert excess mortality estimates into potential reductions of life expectancy at birth and report the cost of climate change on human longevity.
-We find climate change could become the third largest life expectancy reducer behind heart disease and cancer in some European countries.
-Thus, the cost of inaction on climate change could approach one year of life in some countries."
-
-keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
-date: "`r format(Sys.time(), '%B %d, %Y')`"
-geometry: margin=1in
-#fontfamily: mathpazo
-fontsize: 12pt
-spacing: double
-bibliography: LATEX/ccmort.bib
-biblio-style: nature
-fig_caption: true
-# use apsr or nature
-header-includes:
-- \usepackage[all]{nowidow}
-- \usepackage{rotating}
-- \usepackage{fancyhdr}
-- \usepackage[table]{xcolor}
-- \usepackage{tabularx}
-- \usepackage{makecell}
-- \usepackage{xcolor}
-# - \usepackage{threeparttable}
-- \pagestyle{fancy}
-- \fancyhead{}
-- \fancyhead[CE,CO]{Climate change life expectancy}
 ---
 
-* Corresponding author. hauer@uga.edu. p: 706-542-9369.
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE)
+```
 
-^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.
+```{r setup2, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 
-^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.
-
-\newpage
-
-```{r load, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(png)
 rm(list=ls())
 library(tidyverse)
+library(ggrepel)
+library(grid)
+library(gridExtra)
 library(HMDHFDplus)
+library(scales)
+library(dplyr)
+library(openxlsx) 
 library(tmap)
 library(tmaptools)
-library(RColorBrewer)
-library(data.table)
-library(kableExtra)
-library(knitr)
-library(xtable)
+library(tigris)
+library(censusapi)
+library(tidycensus)
+library(cowplot)
+library(magick)
+library(pdftools)
+library(readxl)
 ```
 
-```{r setup, include=FALSE, cache =TRUE}
+```{r setup3, include=FALSE, cache =TRUE}
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-myusername <- getPass()
-mypassword <- getPass()
+myusername <- getPass(msg = "HMD username: ", noblank = FALSE, forcemask = FALSE)
+mypassword <- getPass(msg = "HMD password: ", noblank = FALSE, forcemask = FALSE)
 
@@
 ###   Analysis uses the mean of the last 10 years of data for this death distribution.
-GBD_data <- read.csv("data/GBD_data.csv") %>%
+GBD_data <- read.csv("../data/GBD_data.csv") %>%
   #mutate(perdist = log(val)) %>%
@@
 
-# Introduction
+## Climate Change and Human Health
 
-Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimatin], estimates of displacement [@nicholls2011sea;@hauer2016millions;@hauer2017migration], or human conflict [@barnett2007climate]  but have the potential to lead to the loss of human life [@forzieri2017increasing;@pachauri2014climate].
-As impact quantification studies move further from the physical sciences of climate change into the social sciences, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].
+![](./presentation_files/figure-beamer/climate_change_health_impacts600w.jpg)\
 
-Scholars have long tried to estimate the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer].
-Although such estimates are useful, the excess mortality estimates are rather sterile - one death is a tragedy, a million deaths a statistic - and difficult to relate to on an individual level.
-Life expectancy at birth, on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone.
-To date, no studies exist that comprehensively examine the potential reductions of life expectancy due to climate change.
-Without properly situating the potential loss of life within the contexts of metrics such as life expectancy, we risk underestimating the impact of climate change on human mortality.
+morbidity $\neq$ mortality
 
-Using published data on excess mortality, we connect climate change excess mortality to life expectancy in a mortality model.
-By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in twenty-eight European countries.
-This approach allows us to quantify the impact of climate change on human longevity and answer the following question: How much could climate change mortality reduce life expectancy at birth?
-Our results can situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.
+## Climate Change and Mortality
 
-# Methods
-We estimate changes in life expectancy by combining three primary datasets: excess mortality data from Forzieri et al.
-[@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD].
-Forzieri et al.
-[@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario (SRES A1B).
-They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario (Shared-Socioeconomic Pathway 2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario) and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases within spatially rectified, high-resolution gridded datasets.
-While the disaster data are not standardized across country, Forzeri et al.
-imputed data in incomplete time periods and countries.
-This imputation could mask spatial variability at the sub-national level, but here we use their country-aggregated results.
-Their results represent business-as-usual climate change and human development without incorporating potential adaptation.
-They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes.
-These data provide the anticipated fatalities but not the effect these deaths might have on life expectancy.
+Typically described as excess or "extra" mortality. Most studies quantify exposure rather than mortality.
 
-To convert excess mortality to life expectancy, we use data gathered from the HMD [@HMD] for corresponding European countries in five-year life tables in conjunction with cause-specific mortality data from the GBD [@GBD;@wang2012age].
-The HMD comprises only complete, official vital event statistics and represents the most accurate and complete compilation of mortality data available.
-We allocate excess multi-hazard mortality for each country from supplementary table S8 [@forzieri2017increasing] based on the observed mortality schedule for environmental heat and cold exposure deaths from the GBD for each country in the previous decade.
+Some studies do publish projected mortality:
 
-We then recalculated age-specific mortality rates in five-year life tables by adding the climate change mortality onto the underlying mortality data and create four subsequent life tables (baseline, low, mid, high) using standard life table techniques [@wunsch2013life] based on the published confidence intervals from [@forzieri2017increasing].
-We used a variant of the cause-deleted life tables approach [@brand2005approximations;@beltran2008integrated] to measure the impact of excess mortality on life expectancy at birth or $e_0$.
-Changes in life expectancy are reported based on differentials from $e_0$ between the 2080s and the most recent complete life table in the HMD.
-Complete life tables for all countries are available in the Supplementary Dataset and all equations are located in the Supplementary Information.
+- Forzieri et al (Europe), six-climate hazards: 150k deaths by 2080s
+- Silva et al (Global), ozone-related mortality: 375k by 2100
 
-```{r echo = FALSE, warning= FALSE, message = FALSE}
+These estimates are useful but sterile -- *one death is a tragedy, a million deaths a statistic*.
 
-standard <- left_join(a2, ec_agestandard) %>%
-  left_join(., countrycodes) %>%
-  select(Age, CNTRY, COUNTRY, mx_base, mx_low, mx_mid, mx_high, StnrdPop) %>%
-  mutate(ASDR_base = mx_base * StnrdPop,
-         ASDR_low = mx_low * StnrdPop,
-         ASDR_mid = mx_mid * StnrdPop,
-         ASDR_high = mx_high * StnrdPop) %>%
-  group_by(CNTRY, COUNTRY) %>%
-  summarise(ASDR_base = sum(ASDR_base),
-            ASDR_low = sum(ASDR_low),
-            ASDR_mid = sum(ASDR_mid),
-            ASDR_high = sum(ASDR_high)) %>%
-  mutate(ASDR_low = ASDR_low - ASDR_base,
-         ASDR_mid = ASDR_mid - ASDR_base,
-         ASDR_high = ASDR_high - ASDR_base,
-         COUNTRY = case_when(
-           CNTRY == "FRATNP" ~ "France",
-           CNTRY == "DEUTNP" ~ "Germany",
-           CNTRY == "GBR_NP" ~ "Great Britain",
-           TRUE ~ as.character(COUNTRY))) %>%
-  left_join(., europe_standarddr) %>%
-  ungroup() %>%
-  select(-ASDR_base, -CNTRY)
+## Climate Change and Life Expectancy
 
-selectcountries <- c("Spain", "Luxembourg", "Italy", "France", "Portugal", "Ireland", "Slovenia", "Croatia", "Switzerland", "Belgium")
-testtbl<- standard %>%
-  group_by(COUNTRY) %>%
-  gather(COD, ASDR, 2:13) %>%
-  filter(!COD %in% c("Circulatorydisease", "Cancer", "ASDR_high", "ASDR_low"),
-         !COUNTRY == "Iceland") %>%
-  mutate(RANK = rank(desc(ASDR)),
-         COD = case_when(
-           COD == "Heartdisease" ~ "makecell{Heart\\Disease\\",
-           COD == "LungCancer" ~ "makecell{Lung\\Cancer\\",
-           COD == "Respiratorydiseases" ~ "makecell{Respiratory\\Diseases\\",
-           COD == "Diseasesofthenervoussystem" ~ "makecell{Dis. of the\\Nervous Sys\\",
-           COD == "Colorectalcancer" ~ "makecell{Colorectal\\Cancer\\",
-           COD == "Suicide" ~ "makecell{Suicide\\",
-           COD == "TransportAccidents" ~ "makecell{Transport\\Accidents\\",
-           COD == "ASDR_mid"  ~ "cellcolor{blue!25}makecell{cellcolor{blue!25}Climate\\cellcolor{blue!25}Change\\cellcolor{blue!25}"),
-         Value = paste0(COD, " ", round(ASDR,2), "}")) %>%
-  ungroup()%>%
-  arrange(COD, desc(ASDR)) %>%
-  #mutate(COUNTRY = paste0(row_number(),COUNTRY))%>%
-  select(COUNTRY, Value, RANK) %>%
-  spread(RANK, Value) %>%
-  slice(match(selectcountries, COUNTRY))
+Life expectancy at birth ($e_0$) is the preferred mortality metric.
 
-# library(pander)
-# pander(head(testtbl))
-# xtable(testtbl, caption = "A Caption",
-#       comment=F, include.rownames=F, table.placement="ht",
-#       size=getOption("xtable.size", "tiny"))
-# kable(head(testtbl), format="latex") %>%
-#   # column_spec(2, width = "1cm") %>%
-#   kable_styling(latex_options = "scale_down")
+To our knowledge, *no studies quantify the possible impact of climate change on human longevity*.
 
-# landscape(kable(head(testtbl), "latex"))
+Without properly situating the potential loss of life within metrics like life expectancy, we risk miscommunicating the impact of climate change on human mortality.
 
-  # column_spec(x, 2:9, width = "10em")
+We ask and answer the following questions:
 
-# luxrank1 <- standard$Heartdisease[which(standard$COUNTRY == "Luxembourg")]
-# luxrank2 <- standard$ASDR_mid[which(standard$COUNTRY == "Luxembourg")]
-# luxrank3 <- standard$Respiratorydiseases[which(standard$COUNTRY == "Luxembourg")]
-# luxrank4 <- standard$LungCancer[which(standard$COUNTRY == "Luxembourg")]
-# luxrank5 <- standard$Diseasesofthenervoussystem[which(standard$COUNTRY == "Luxembourg")]
-# luxrank6 <- standard$Colorectalcancer[which(standard$COUNTRY == "Luxembourg")]
-# luxrank7 <- standard$Suicide[which(standard$COUNTRY == "Luxembourg")]
-# luxrank8 <- standard$TransportAccidents[which(standard$COUNTRY == "Luxembourg")]
+- What is the cost of inaction on climate change on human longevity?
+- How does climate change related mortality compare to other mortality vectors?
 
-```
+## Data 1
 
-# Results
-We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig.
-1**).
-Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy.
-Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.
+We use data from three primary sources:
 
-The overall reduction in $e_0$ is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States.
-**Table 1** reports the ranking of Climate Change compared to the major causes of death (COD) in European countries using age-standardised death rates and the standard European population [@pace2013revision] for the year 2013.
-In three European countries (Spain, Luxembourg, Italy), Climate Change could become a top three COD.
-In Spain (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Spain")],2)`), Luxembourg (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Luxembourg")],2)`), France (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "France")],2)`), and Italy (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Italy")],2)`) Climate Change could become the leading cause of death if Climate Change is more extreme than anticipated.
-Only all circulatory diseases combined and all cancers combined would cause more deaths.
+1.
+Forzieri et al projected excess mortality due to six climate vectors (heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms) under conventional greenhouse gas emission scenario (SRES A1B).
+(150k deaths per year by 2080s)
 
-```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change.
-We report changes in life expectancies due to climate change for twenty-eight European countries.
-The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}
+![](./presentation_files/figure-beamer/forzieri.PNG){ width=45% } via: Forzieri et al (2017)
 
-ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=dif_mid)) +
-  geom_pointrange(stat="identity") +
-  geom_point(stat="identity", fill="black", size=8) +
-   geom_hline(yintercept=0) +
-  geom_text(color="white", size=2) +
-  #scale_x_discrete(limits = rev(levels(the_factor)))
-  scale_x_reverse(breaks = lt_lancet$RANK, labels=lt_lancet$COUNTRY) +
-  coord_flip() +
-  theme_bw() +
-  #ylim(-2, 2) +
-  theme(panel.grid.minor = element_blank()) +
+## Data 2
 
-   labs(y=expression(Delta* e[0]),
-        x= "Country")
-```
+2. To convert Forzieri et al's excess mortality to life expectancy, we use the Human Mortality Database (HMD) for underlying life tables.
 
-Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies.
-Whereas many Northern European countries could experience negligible impacts on life expectancy, five European countries could see life expectancy changes in excess of -0.5 years (Luxembourg, Spain, Italy, France, and Ireland).
-This group of countries could see life expectancy changes of more than -1.0 years if climate change hazards are more intense than anticipated.
-In some countries, climate change could thus become a bigger killer than trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or all accident related mortality (-0.84 years) [@arias2013united] by the end of the century.
+- HMD for the correspoding 28 European countries in 5x1 (by age and year) life tables for the most recent life table.
 
-```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}
+    - Abridge the life table to $[0,1,5,...,80+]$
 
-data(Europe)
-Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)
+- $_nP_x$, population aged $x$ to $x+n$
 
-bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)
+- $_na_x$, average length of survival between ages $x$ and $x+n$
 
-  tm_shape(Europe2) +
-  tm_fill("DIF_MID", title = expression(Delta* e[0 ~italic(MID)]), palette = "YlOrBr", style = "jenks") +
-  tm_borders(alpha = 0.5) +  
-  tm_text("Test" , size="AREA", root=5) +
-  tm_format_Europe() 
-```
+- $_nd_x$, number of deaths between ages $x$ and $x+n$
 
-While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig.
-2**).
-The most westerly European countries tend to have the greatest reductions.
-Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change.
-Emergent research in Atmospheric Rivers suggest Western Europe could be more susceptible to these types of events [@ramos2015daily].
-Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.
+## Data 3
 
-# Discussion
-In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries.
-Previous studies on climate change and excess mortality potentially misrepresent the impact climate change could have on human mortality.
-Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy].
-Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer].
-Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes.
-Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].
+3. The Global Burden of Disease (GBD) provides cause-specific mortality by age/sex/geography/year for 249 causes of death.
 
-\begin{sidewaystable}
-\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Climate Change COD is based on the 2080s while the other CODs are based on the 2010s.}
-\rowcolors{2}{gray!25}{white}
-\small
-\begin{tabularx}{\textwidth}{l|XXXXXXXXX}
-  \hline
-   \rowcolor{white}
- Country & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\ 
-  \hline
-Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
-of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\
-Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
-of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\
-Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
-of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\
-France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis.
-of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\
-Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
-of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\
-Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis.
-of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\
-Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02} & \makecell{Dis.
-of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\
-Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44} & \makecell{Dis.
-of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\
-Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis.
-of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\
-Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis.
-of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\
-   \hline
-\end{tabularx}
-\vspace{1ex}
- \raggedright Data for other causes of death come from Eurostat (online data code: hlth_cd_asdr2). These results reflect the Standard European Population for 2013.
-\end{sidewaystable}
+- We gathered ata on age-specific deaths and mortality rates for environmental heat and cold exposure (cause C.2.9) for 2006-2015 for each country in the study.
 
-Without adaptation measures, our results suggest climate change could emerge as a significant new mortality vector and could pose a major public health threat for some European countries by the end of the century, echoing previous findings [@forzieri2017increasing;@patz2005impact].
-Life expectancy has steadily risen across the world for the last century [@oeppen2002broken] and our results suggest that climate change alone could spur a sharp reversal in these trends.
-We expect one European countries to see life expectancy reductions more than one year under the middle scenario (Luxembourg), but if climate change has a greater impact on mortality than anticipated, five European countries (Luxembourg, Spain, Italy, France, Ireland) could see life expectancy reductions more than one year with Luxembourg experiencing a reduction of more than two.
-
-Reductions such as those should not be taken lightly.
-Many of the children born today are likely to still be alive by the end of the century and will be in the age groups most threatened by extreme heat (65+) - the biggest mortality risk associated with climate change [@keatinge2000heat].
-If climate change unfolds as a more aggressive mortality vector, only major cardiovascular diseases and malignant neoplasms would contribute more to life expectancy reductions in numerous European countries [@arias2013united].
-This would make climate change one of the most aggressive new mortality vectors to emerge over the last quarter-century, representing a major threat to public health in many parts of Europe.
-
-Prospective studies on the emerging threat from climate change rely on linking contemporary mortality with future mortality.
-However, climate change could reshape future mortality through other causes of death.
-Climate change affects health behaviors that in turn increase mortality risk through increased alcohol and substance abuse, violent behavior, insecurity, increase in post-traumatic stress due to weather-related trauma, increase in stress due to climate change and schizophrenia, increase in the use of medications that reduce the ability to perspire and sweat, etc.
-[@patz2005impact] The International Classifications of Diseases and Related Health Problems (ICD) does not contain ``climate change" as an official cause of death, so we can only speculate that the impact of climate change could be larger than reported here.
-Although we do not model these potential impacts, our results could thus be considered conservative.
-
-We also share the concerns of Lee et al.
-[@lee2017comprehensive] concerning the business-as-usual climatic assumptions.
-It is likely that many countries and communities will deploy a wide variety of adaptation measures [@haines2006climate;@kovats2003methods;@ebi2006approach].
-These adaptation measures rely on accurate information about the potential mortality vectors.
-Our models and those produced by Forzieri et al.
-[@forzieri2017increasing] present plausible scenarios on the potential impact of climate change on human mortality and provide crucial information to public health officials, national governments, and international organizations.
-The time frames associated with climate change allow ample time for this potential health crisis to be averted.
-
-Finally, we would like to point out that future research should not only transform excess mortality into life expectancy decrements.
-Given the influence of climate change in diseases and causes of death, it is imperative to quantify the extent to which climate change will derive in increasing costs for health care systems in these countries.
-The health care structures are being taxed by population aging [@rechel2009can] and rising health care costs, yet it remains unclear how climate change will exacerbate these pressures.
-
-\newpage
-
-# Supplemenatary Information
-## Data
-To create life expectancy differentials, we used three sources of data.
-The first is the baseline life table data.
-Baseline life tables comes from the Human Mortality Database (HMD) [@HMD].
-These datasets comprise only official vital event registration and are considered the most complete and accurate source of mortality data available.
-We downloaded the most recent 5x1 (by age and year) life table data for twenty-eight European countries.
-We used the following data elements from the five-year age group life table data in the HMD:
-
-\begin{enumerate}
-\item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
-\item $_na_x$ or the average length of survival between ages $x$ and $x+n$.
-\item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
-\end{enumerate}
-
-The second dataset comes from published projections of climate-related mortality in Europe [@forzieri2017increasing].
-Forzieri et al.
-[@forzieri2017increasing] combined data from the International Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool (NatCatSERVICE) to create exposure and fatality statistics related to six weather-related hazards -- heatwaves, cold waves, droughts, wildfires, river and coastal floods, and windstorms.
-They downscaled these exposures to  1km grid scales and integrated them into small-area demographic projections for Europe out to 2100.
-Heatwaves and coldwaves account for the vast majority of projected fatalities.
-The number of future anticipated deaths are available in their Supplementary Table S8 [@forzieri2017increasing] and provide the magnitude of deaths for our study.
-
-The third dataset comes from the GBD [@GBD;@wang2012age].
-The GBD is the most comprehensive worldwide dataset of epidemiological data produced and provides cause-specific mortality by age/sex/geography/year for 249 causes of death in 195 countries and territories.
-We gathered data on age-specific deaths and mortality rates ($m_x$) for mortality from environmental heat and cold exposure (cause C.2.9) for the period 2006-2015 for each country in the study.
-These data provide a mortality schedule to fit the projected climate change excess mortality from Fozieri et al [@forzieri2017increasing].
-
-```{r suppfig, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("**Mortality distribution of Heat-related mortality in 28 European countries**.
-The thick black line is the average\\label{suppfig}")}
-GBD_data <- read.csv("data/GBD_data.csv") %>%
+```{r suppfig, echo= FALSE, message = FALSE, warning = FALSE, fig.width = 4, fig.height=2, fig.align = "center"}
+GBD_data <- read.csv("../data/GBD_data.csv") %>%
   #mutate(perdist = log(val)) %>%
@@
   theme_bw() +
-  theme(legend.position = c(0.8,0.25),
+  theme(legend.position = "none",
         legend.title=element_blank(),
-        legend.text=element_text(size=4),
+        legend.text=element_text(size=2),
         legend.key.size = unit(1,"line")) +
-  guides(col = guide_legend(ncol = 4)) +
+  # guides(col = guide_legend(ncol = 4)) +
   scale_y_log10()+
-  labs(y="log(Mortality Percentage Distribution)",
-       #title = "Mortality distribution of Heat-related mortality in 28 European countries",
-       caption = "Source: Global Burden of Disease")
+  labs(y="log(Mortality)")
+       #title = "Mortality distribution of Heat-related mortality in 28 European countries")
 
@@
 ## Methods
-We derived a number of additional variables to accomplish our analysis.
-First, we abridged the life table data from the HMD from ages 100+ to 80+ to conform to the GBD cause-specific mortality schedules.
 
-From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}$).
+- From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}$).
 
-We then derived additional $m_x$ rates for each age group $x$ in each country $i$ for each scenario $s$ ($BASE$, $LOW$, $MID$, $HIGH$) from Forzieri et al [@forzieri2017increasing].
+- We then derived additional $m_x$ rates for each age group $x$ in each country $i$ for each scenario $s$ ($BASE$, $LOW$, $MID$, $HIGH$) from Forzieri et al.
 
 \begin{equation}
-_nm_{x,i,BASE} = _nD_{x,i}^{HMD} \,/\, _nP_{x,i}^{HMD}
+_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + \mathbf{(\hat{D_i} \,\cdot\, _nt_{x,i})} ) \,/\, _nP_{x,i}^{HMD}
 \end{equation}
 
-\begin{equation}
-_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_i} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
-\end{equation}
+- Calculate the remaining life table using standard life table techniques and simply subtract $e_{x,i,s}$ from $e_{x,i,BASE}$ to arrive a $\Delta e_x$.
 
-where $D_{x,i}$ is the number of deaths in age group $x$ in country $i$ from the HMD, $P_{x,i}$ is the population in age group $x$ from the HMD, $\hat{D_x}$ is the number of deaths from Forzieri et al.
-[@forzieri2017increasing], and $t_{x,i}$ is the proportion of mortality experienced in each age group $x$ from the GBD.
-Thus, the anticipated additional mortality due to climate change is added to each age group based on the underlying mortality cause-specific mortality schedule observed between 2006-2015 in the GBD.
+## Results
 
-We then calculated $q_x$ values, or the probability of dying, for each scenario $s$ for each country.
+```{r results1, echo= FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
+ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=dif_mid)) +
+  geom_pointrange(stat="identity") +
+  geom_point(stat="identity", fill="black", size=15) +
+   geom_hline(yintercept=0) +
+  geom_text(color="white", size=4.5) +
+  #scale_x_discrete(limits = rev(levels(the_factor)))
+  scale_x_reverse(breaks = lt_lancet$RANK, labels=lt_lancet$COUNTRY) +
+  coord_flip() +
+  theme_bw() +
+  #ylim(-2, 2) +
+  theme(panel.grid.minor = element_blank(),
+        text = element_text(size=20)) +
+    labs(y=expression(Delta* e[0]),
+        x= "Country")
 
-\begin{equation}
-_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}
-\end{equation}
+#, fig.width = 10, fig.height=6,
 
-\begin{equation}
-_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}
-\end{equation}
+```
 
-We calculated each additional life table value identically for each scenario $s$ using standard life table equations:
+## Results
+```{r results2, echo= FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
+data(Europe)
+Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)
 
-$$_nd_{x,i,s} = _nl_{x,i,s} \cdot _nq_{x,i,s}$$
-$$_nl_{x,i,s} = _nl_{x-1,i,s} - _nd_{x-1,i,s}$$
-$$_nL_{x,i,s} = _na_{x,i,s} \cdot _nl_{x,i,s} + ((n-_na_{x,i,s}) \cdot _nl_{x,i,s})$$
-$$_nT_{x,i,s} = _nL_{x,i,s} + _nT_{x+1,i,s}$$
-$$_ne_{x,i,s} = _nT_{x,i,s} / _nl_{x,i,s}$$
+bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)
 
-To determine the differences between life expectancy compared to the baseline, we simply subtract $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].
+  tm_shape(Europe2) +
+  tm_fill("DIF_MID", title = expression(Delta* e[0 ~italic(MID)]), palette = "YlOrBr", style = "jenks") +
+  tm_borders(alpha = 0.5) +  
+  tm_text("Test" , size="AREA", root=5) +
+  tm_format_Europe()
+
+```
+
+## Results
+
+![](./presentation_files/figure-beamer/table2.PNG){ width=100% }\
+
+## Conclusion
+
+- Two European countries could see reductions of more than one-year ($MID$) but five countries could see more than one-year reduction under higher scenarios ($high$).
+
+- Thus, **the cost of inaction on climate change could approach *one year of life* in some countries**.
+
+- Life expectancy connects mortality estimates into more intuitively understandable metrics, translating global estimates of mortality into individual outcomes.
+
+- Without adaptation measures, our results suggest climate change could emerge as a significant new mortality vector for some European countries by the end of the century.
+
+## Thank you!
+
+**Mathew E Hauer**
+Applied Demographer 
+Carl Vinson Institute of Government \
+hauer@uga.edu
+
+**Alexis R Santos **
+Applied Demographer
+Pennsylvania State University

### 2018-05-14 MS-cclifeexpec.Rmd
duplicate figures removed.

@@
 author:
-# - name: Mathew E. Hauer^1^*
-#   affiliation: University of Georgia
-# - name: Alexis R. Santos^2^
-#   affiliation: Pennsylvania State University
+- name: Mathew E. Hauer^1^*
+  affiliation: University of Georgia
+- name: Alexis R. Santos^2^
+  affiliation: Pennsylvania State University
 
@@
 
-<!-- * Corresponding author. hauer@uga.edu. p: 706-542-9369. -->
+* Corresponding author. hauer@uga.edu. p: 706-542-9369.
 
-<!-- ^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602. -->
+^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.
 
-<!-- ^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802. -->
+^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.
 
@@
 
-```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change.
-We report changes in life expectancies due to climate change for twenty-eight European countries.
-The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}
-
-ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=dif_mid)) +
-  geom_pointrange(stat="identity") +
-  geom_point(stat="identity", fill="black", size=8) +
-   geom_hline(yintercept=0) +
-  geom_text(color="white", size=2) +
-  #scale_x_discrete(limits = rev(levels(the_factor)))
-  scale_x_reverse(breaks = lt_lancet$RANK, labels=lt_lancet$COUNTRY) +
-  coord_flip() +
-  theme_bw() +
-  #ylim(-2, 2) +
-  theme(panel.grid.minor = element_blank()) +
-
-   labs(y=expression(Delta* e[0]),
-        x= "Country")
-```
-
 Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies.
@@
 In some countries, climate change could thus become a bigger killer than trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or all accident related mortality (-0.84 years) [@arias2013united] by the end of the century.
-
-```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}
-
-data(Europe)
-Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)
-
-bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)
-
-  tm_shape(Europe2) +
-  tm_fill("DIF_MID", title = expression(Delta* e[0 ~italic(MID)]), palette = "YlOrBr", style = "jenks") +
-  tm_borders(alpha = 0.5) +  
-  tm_text("Test" , size="AREA", root=5) +
-  tm_format_Europe() 
-```
 

### 2018-05-14 MS-cclifeexpec.tex
duplicate figures removed.

@@
 
-\author{}
+\author{\Large Mathew E.
+Hauer\textsuperscript{1}*\vspace{0.05in} \newline\normalsize\emph{University of Georgia}   \and \Large Alexis R.
+Santos\textsuperscript{2}\vspace{0.05in} \newline\normalsize\emph{Pennsylvania State University}  }
 
@@
             bookmarks=true,
-            pdfauthor={},
+            pdfauthor={Mathew E. Hauer\textsuperscript{1}* (University of Georgia) and Alexis R. Santos\textsuperscript{2} (Pennsylvania State University)},
              pdfkeywords = {Climate change, Life expectancy, Mortality, Demography, Excess
@@
    \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
+\textbf{\authorfont Mathew E.
+Hauer\textsuperscript{1}*} \hskip 15pt \emph{\small University of Georgia}   \par \textbf{\authorfont Alexis R.
+Santos\textsuperscript{2}} \hskip 15pt \emph{\small Pennsylvania State University}
 
@@
 
-\noindent \doublespacing \newpage
+\noindent \doublespacing \begin{itemize}
+\tightlist
+\item
+  Corresponding author.
+  \href{mailto:hauer@uga.edu}{\nolinkurl{hauer@uga.edu}}. p:
+  706-542-9369.
+\end{itemize}
+
+\textsuperscript{1} Carl Vinson Institute of Government, University of
+Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.
+
+\textsuperscript{2} Department of Sociology and Criminology,
+Pennsylvania State University. 901 Oswald Tower, University Park, PA USA
+16802.
+
+\newpage
 
@@
 
-\begin{figure}
-\centering
-\includegraphics{MS-cclifeexpec_files/figure-latex/unnamed-chunk-2-1.pdf}
-\caption{Change in life expectancy at birth (\(e_0\)) due to
-business-as-usual climate change. We report changes in life expectancies
-due to climate change for twenty-eight European countries. The central
-values represent the ensemble median while the stems represent the upper
-and lower bounds of the inter-model climate variability.}
-\end{figure}
-
 Our results also suggest climate change mortality differentials are
@@
 (-0.84 years) \citep{arias2013united} by the end of the century.
-
-\begin{figure}
-\centering
-\includegraphics{MS-cclifeexpec_files/figure-latex/unnamed-chunk-3-1.pdf}
-\caption{Estimated reduction in life expectancy at birth (\(e_0\)) by
-the 2080s under the \(MID\) scenario.}
-\end{figure}
 

### 2018-06-13 MS-cclifeexpec_paa.Rmd
from computer.

--- 2018-05-14 MS-cclifeexpec.Rmd
@@
 
-<!-- # Methods -->
+# Methods
 We estimate changes in life expectancy by combining three primary datasets: excess mortality data due to meterological hazards related to climate change from Forzieri et al.
@@
 
-<!-- # Results -->
+# Results
 We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` to `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**).
@@
 
-<!-- # Results -->
+# Results
 We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig.
@@
 
+```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change.
+We report changes in life expectancies due to climate change for twenty-eight European countries.
+The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}
+
+ggplot(lt_lancet, aes(x=RANK, y=DIF_MID, ymin = DIF_LOW, ymax = DIF_HIGH, label=dif_mid)) +
+  geom_pointrange(stat="identity") +
+  geom_point(stat="identity", fill="black", size=8) +
+   geom_hline(yintercept=0) +
+  geom_text(color="white", size=2) +
+  #scale_x_discrete(limits = rev(levels(the_factor)))
+  scale_x_reverse(breaks = lt_lancet$RANK, labels=lt_lancet$COUNTRY) +
+  coord_flip() +
+  theme_bw() +
+  #ylim(-2, 2) +
+  theme(panel.grid.minor = element_blank()) +
+
+   labs(y=expression(Delta* e[0]),
+        x= "Country")
+```
+
 Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies.
@@
 In some countries, climate change could thus become a bigger killer than trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or all accident related mortality (-0.84 years) [@arias2013united] by the end of the century.
+
+```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}
+
+data(Europe)
+Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)
+
+bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)
+
+  tm_shape(Europe2) +
+  tm_fill("DIF_MID", title = expression(Delta* e[0 ~italic(MID)]), palette = "YlOrBr", style = "jenks") +
+  tm_borders(alpha = 0.5) +  
+  tm_text("Test" , size="AREA", root=5) +
+  tm_format_Europe() 
+```
 

### 2018-06-13 MS-cclifeexpec_paa.tex
from computer.

--- 2018-05-14 MS-cclifeexpec.tex
@@
 
+\section{Methods}\label{methods}
+
 We estimate changes in life expectancy by combining three primary
@@
 
+\section{Results}\label{results}
+
 We find that climate change could alter life expectancy by -0.23 years
@@
 \centering
-\includegraphics{MS-cclifeexpec_files/figure-latex/figure1-1.pdf}
+\includegraphics{MS-cclifeexpec_paa_files/figure-latex/figure1-1.pdf}
 \caption{Change in life expectancy at birth (\(e_0\)) due to
@@
 \centering
-\includegraphics{MS-cclifeexpec_files/figure-latex/figure2-1.pdf}
+\includegraphics{MS-cclifeexpec_paa_files/figure-latex/figure2-1.pdf}
 \caption{Estimated reduction in life expectancy at birth (\(e_0\)) by
@@
 
+\section{Results}\label{results-1}
+
 We find that climate change could alter life expectancy by -0.23 years
@@
 
+\begin{figure}
+\centering
+\includegraphics{MS-cclifeexpec_paa_files/figure-latex/unnamed-chunk-2-1.pdf}
+\caption{Change in life expectancy at birth (\(e_0\)) due to
+business-as-usual climate change. We report changes in life expectancies
+due to climate change for twenty-eight European countries. The central
+values represent the ensemble median while the stems represent the upper
+and lower bounds of the inter-model climate variability.}
+\end{figure}
+
 Our results also suggest climate change mortality differentials are
@@
 (-0.84 years) \citep{arias2013united} by the end of the century.
+
+\begin{figure}
+\centering
+\includegraphics{MS-cclifeexpec_paa_files/figure-latex/unnamed-chunk-3-1.pdf}
+\caption{Estimated reduction in life expectancy at birth (\(e_0\)) by
+the 2080s under the \(MID\) scenario.}
+\end{figure}
 
@@
 
-\section{Methods}\label{methods}
+\section{Methods}\label{methods-1}
 
@@
 \centering
-\includegraphics{MS-cclifeexpec_files/figure-latex/suppfig-1.pdf}
+\includegraphics{MS-cclifeexpec_paa_files/figure-latex/suppfig-1.pdf}
 \caption{\textbf{Mortality distribution of Heat-related mortality in 28
@@
 
-\subsection{Methods}\label{methods-1}
+\subsection{Methods}\label{methods-2}
 

### 2018-06-13 PRESENTATIONS/presentation.Rmd
from computer.

@@
 ---
-title: "Climate change projected to reduce European life expectancy"
+title: Climate Change Projected to Reduce European Life Expectancy
+# subtitle: (A Subtitle Would Go Here if This Were a Class)
 author: | 
  | Mathew E. Hauer^1^ & Alexis R. Santos^2^
+institute:  | 
  | ^1^University of Georgia | ^2^Pennsylvania State University
-date: "April 26, 2018"
-output:   beamer_presentation
-classoption: "aspectratio=169"
-
+titlegraphic1: logo-uga-color.eps
+titlegraphic2: pennstate.png
+fontsize: 10pt
+output:
+ beamer_presentation:
+    template: ../PRESENTATIONS/presentation_template.tex
+    keep_tex: TRUE
+make149: true
 ---
@@
 
-**Mathew E Hauer**
-Applied Demographer 
-Carl Vinson Institute of Government \
+**Mathew E Hauer**\newline
+Applied Demographer \newline
+Carl Vinson Institute of Government \newline
 hauer@uga.edu
+\newline
+\newline
 
-**Alexis R Santos **
-Applied Demographer
+**Alexis R Santos **\newline
+Applied Demographer\newline
 Pennsylvania State University

### 2018-06-13 SCRIPTS/001-ANALYSIS_md.Rmd
from computer.

--- 2018-04-05 supplementary_rcode.Rmd
@@
 ---
-title: "Climate Change Projected to Reduce European Life Expectancy"
-output:
-  word_document: default
-  pdf_document: default
+title: "Untitled"
+author: "Mathew E. Hauer"
 date: "April 5, 2018"
+output: pdf_document
 ---
@@
 
-## R Data Load and Processing The following code chunk will 
-1. Import the underlying Forzieri et al data. 
-2. Import the Standard European Population from EuroStat. 
-3. Import the leading causes of death with age-standardized death rates from EuroStat. 
-4. Import the Global Burden of Disease data. This data is located on github. 
-5. Download all of the Human Mortality Database data.
+## R Markdown
 
-To successfully run this code, several R pacakgse must be installed and these
-packages are located just below this text section but are tidyverse, HMHFDplus,
-tmap, tmaptools, RColorBrewer, data.table, and getPass.
+This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
 
-To successfully run this code, you must also have an account at the HMD and
-input your username and password when prompted.
+When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
+You can embed an R code chunk like this:
 
@@
 
-# Figure 1
-This chunk of code reproduces Figure 1.
 ```{r Figure1}
@@
 
-# Figure 2
-This chunk of code reproduecs Figure 2.
 ```{r Figure2}
@@
 
-# Supplementay Figure 1
-This chunk of code reproduces the Supplementary Figure
 ```{r suppfig}
@@
 
-# Table 1
-This chunk of code reproduces the data inside Table 1 in the main document.
-
-Note: the output from this table (the object testtbl) is formatted for latex rather than markdown.
-To completely reproduce that table, the output from testtbl must be pasted into a latex compiler and all forward slashes ("/") must be converted to backward slashes ("\").
 ```{r Table1}

### 2018-08-22 MS-cclifeexpec.Rmd
New branch

--- 2018-04-05 MS-cclifeexpec.Rmd
@@
     fig_caption: true
-    latex_engine: pdflatex
+    latex_engine: xelatex
     template: LATEX/svm-latex-ms.tex
-title: "Climate change projected to reduce European life expectancy"
+title: "Inaction on climate change could approach one year of life in some European countries"
 thanks: "All data and code that supports these conclusions are available as supplementary materials."
 author:
-# - name: Mathew E. Hauer^1^*
-#   affiliation: University of Georgia
-# - name: Alexis R. Santos^2^
-#   affiliation: Pennsylvania State University
+- name: Mathew E. Hauer^1^*
+  affiliation: University of Georgia
+- name: Alexis R. Santos^2^
+  affiliation: Pennsylvania State University
 
-abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change [@patz2005impact;@aastrom2013attributing;@forzieri2017increasing;@staddon2014climate].
-However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups [@parrish2010peer].
+abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change.
+However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups.
 No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy.
 Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting cost of climate change on mortality.
-In this paper, we convert excess mortality estimates into potential reductions of life expectancy at birth and report the cost of climate change on human longevity.
-We find climate change could become the third largest life expectancy reducer behind heart disease and cancer in some European countries.
-Thus, the cost of inaction on climate change could approach one year of life in some countries."
+In this paper, we convert excess mortality estimates due to increaes in extreme weather from climate change into potential reductions of life expectancy at birth and report the cost of climate change on human longevity.
+We find climate change extremes could become the third largest life expectancy reducer behind heart disease and cancer in some European countries.
+Thus, the cost of inaction on climate change could approach one year of life in some countries and possibly exceed that."
 
@@
 bibliography: LATEX/ccmort.bib
-biblio-style: nature
+biblio-style: apsr
+# use apsr or nature
 fig_caption: true
-# use apsr or nature
 header-includes:
@@
 
-<!-- * Corresponding author. hauer@uga.edu. p: 706-542-9369. -->
+* Corresponding author. hauer@uga.edu. p: 706-542-9369.
 
-<!-- ^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602. -->
+^1^ Department of Sociology, Florida State University. 201 N. Milledge Ave. Athens, GA USA 30602.
 
-<!-- ^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802. -->
+^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.
 
@@
   mutate(ISO3 = CNTRY) %>%
-  select(COUNTRY, CNTRY, ISO3)
+  dplyr::select(COUNTRY, CNTRY, ISO3)
 #  read.csv("data/LANcET_CCMORTDATA.csv",colClasses=c("CNTRY"="character"))
@@
 
-ggplot(data=GBD_data, aes(x=Age, y=perdist)) +
-  geom_line(aes(group =CNTRY, col=CNTRY), alpha=0.7) +
-  geom_smooth(span=0.3, se=F, col="black") +
-  theme_bw() +
-  scale_y_log10()+
-  labs(y="log(Mortality Percentage Distribution)")
+# ggplot(data=GBD_data, aes(x=Age, y=perdist)) +
+#   geom_line(aes(group =CNTRY, col=CNTRY), alpha=0.7) +
+#   geom_smooth(span=0.3, se=F, col="black") +
+#   theme_bw() +
+#   scale_y_log10()+
+#   labs(y="log(Mortality Percentage Distribution)")
 
 a <-  left_join(a, lancetdata)  %>%
-  select(Age, Pop, CNTRY, deaths, ax, MID, LOW, HIGH) %>%
+  dplyr::select(Age, Pop, CNTRY, deaths, ax, MID, LOW, HIGH, Year) %>%
    filter(!is.na(MID))
@@
 Scholars have long estimated the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer].
+
 Although such estimates are useful, the excess mortality estimates are rather sterile -- one death is a tragedy, a million deaths a statistic -- and difficult to relate to on an individual level.
@@
 
-Using published data on excess mortality[@forzieri2017increasing], we connect climate change excess mortality to life expectancy in a mortality model.
+Using published data on excess mortality [@forzieri2017increasing], we connect climate change excess mortality to life expectancy in a mortality model.
 By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in twenty-eight European countries.
@@
 and How does climate change related mortality compare to other mortality vectors?
-Our results can situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.
+Our results situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.
 
-<!-- # Methods -->
+Our results suggest that climate change could emerge as a potent new mortality vector in the 20th century.
+Mitigation strategies (ie reductions in greenhouse gas emisions) and adaptation strategies (ie outreach efforts, retrofitting public buildings, etc.) would help prevent this new public health concern.
+
+# Background
+## Climate change and human mortality
+
+# Methods and Materials
 We estimate changes in life expectancy by combining three primary datasets: excess mortality data due to meterological hazards related to climate change from Forzieri et al.
@@
 
+\begin{table}
+\caption{\textbf{Most recent mortality data available by country in the HMD.}} \label{Table2}
+\begin{tabular}{ll|ll|ll|ll}
+Country & Year & Country & Year & Country & Year & Country & Year \\
+\hline
+AUT     & 2014 & ESP     & 2014 & HUN     & 2014 & NLD     & 2016 \\
+BEL     & 2015 & EST     & 2014 & IRL     & 2014 & NOR     & 2014 \\
+BGR     & 2010 & FIN     & 2015 & ISL     & 2016 & POL     & 2014 \\
+CHE     & 2016 & FRATNP  & 2015 & ITA     & 2014 & PRT     & 2015 \\
+CZE     & 2016 & GBR     & 2016 & LTU     & 2014 & SVK     & 2014 \\
+DEUT    & 2015 & GRC     & 2013 & LUX     & 2014 & SVN     & 2014 \\
+DNK     & 2016 & HRV     & 2016 & LVA     & 2014 & SWE     & 2016 \\
+\hline
+\end{tabular}
+\end{table}
+
 We then recalculated age-specific mortality rates in five-year life tables by adding the climate change mortality onto the underlying mortality data and create four subsequent life tables ($baseline$, $low$, $mid$, $high$) using standard life table techniques [@wunsch2013life] based on the published confidence intervals from [@forzieri2017increasing].
 We used a variant of the cause-deleted life tables approach [@brand2005approximations;@beltran2008integrated] to measure the impact of excess mortality on life expectancy at birth or $e_0$.
-Changes in life expectancy are reported based on differentials from $e_0$ between the 2080s and the most recent complete life table in the HMD.
-Complete life tables for all countries are available in the Supplementary Dataset and all equations are located in the Supplementary Information.
+Changes in life expectancy are reported based on differentials from $e_0$ between the 2080s and the most recent complete life table in the HMD (see **Table 1** for the years used).
+Complete life tables for all countries are available in the *Supplementary Dataset* and all computer code to replicate our results are available online[^1].
 
-<!-- # Results -->
-We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**).
-This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States.
+[^1]: Replication files are located here: <https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1>
+
+## Data
+To create life expectancy differentials, we used three sources of data.
+The first is the baseline life table data.
+Baseline life tables comes from the Human Mortality Database (HMD) [@HMD].
+These datasets comprise only official vital event registration and are considered the most complete and accurate source of mortality data available.
+We downloaded the most recent 5x1 (by age and year) life table data for twenty-eight European countries.
+We used the following data elements from the five-year age group life table data in the HMD:
+
+\begin{enumerate}
+\item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
+\item $_na_x$ or the average length of survival between ages $x$ and $x+n$.
+\item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
+\end{enumerate}
+
+The second dataset comes from published projections of climate-related mortality in Europe [@forzieri2017increasing].
+Forzieri et al.
+[@forzieri2017increasing] combined data from the International Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool (NatCatSERVICE) to create exposure and fatality statistics related to six weather-related hazards -- heatwaves, cold waves, droughts, wildfires, river and coastal floods, and windstorms.
+They downscaled these exposures to  1km grid scales and integrated them into small-area demographic projections for Europe out to 2100.
+Heatwaves and coldwaves account for the vast majority of projected fatalities.
+The number of future anticipated deaths are available in their Supplementary Table S8 [@forzieri2017increasing] and provide the magnitude of deaths for our study.
+
+The third dataset comes from the GBD [@GBD;@wang2012age].
+The GBD is the most comprehensive worldwide dataset of epidemiological data produced and provides cause-specific mortality by age/sex/geography/year for 249 causes of death in 195 countries and territories.
+We gathered data on age-specific deaths and mortality rates ($m_x$) for mortality from environmental heat and cold exposure (cause C.2.9) for the period 2006-2015 for each country in the study.
+These data provide a mortality schedule to fit the projected climate change excess mortality from Fozieri et al [@forzieri2017increasing].
+Thus, we assume that the age-specific mortality schedules observed between 2006 and 2015 due to environmental heat and cold exposure in the GBD (see \autoref{suppfig}) are likely to remain unchanged in the future.
+
+```{r suppfig, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("**Mortality distribution of Heat-related mortality in 28 European countries**.
+The thick black line is the average\\label{suppfig}")}
+GBD_data <- read.csv("data/GBD_data.csv") %>%
+  #mutate(perdist = log(val)) %>%
+  group_by(countrycode, Age) %>%
+  summarise(meanval = mean(val),
+            upperval = mean(upper),
+            lowerval = mean(lower)) %>%
+  group_by(countrycode) %>%
+  mutate(countmeanval = sum(meanval),
+         countupperval = sum(upperval),
+         countlowerval = sum(lowerval),
+         perdist = (meanval / countmeanval)) %>%
+  rename(CNTRY = countrycode)
+
+ggplot(data=GBD_data, aes(x=Age, y=perdist)) +
+  geom_line(aes(color=CNTRY), alpha=1) +
+  geom_smooth(span=0.3, se=F, col="black", lwd =2) +
+  theme_bw() +
+  theme(legend.position = c(0.8,0.25),
+        legend.title=element_blank(),
+        legend.text=element_text(size=4),
+        legend.key.size = unit(1,"line")) +
+  guides(col = guide_legend(ncol = 4)) +
+  scale_y_log10()+
+  labs(y="log(Mortality Percentage Distribution)",
+       #title = "Mortality distribution of Heat-related mortality in 28 European countries",
+       caption = "Source: Global Burden of Disease")
+
+```
+
+## Methods
+We derived a number of additional variables to accomplish our analysis.
+First, we abridged the life table data from the HMD from ages 100+ to 80+ to conform to the GBD cause-specific mortality schedules.
+
+From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_{\alpha,i}^{GBD}}$).
+
+We then derived additional $m_x$ rates for each age group $x$ in each country $i$ for each scenario $s$ ($BASE$, $LOW$, $MID$, $HIGH$) from Forzieri et al [@forzieri2017increasing].
+
+\begin{equation}
+_nm_{x,i,BASE} = _nD_{x,i}^{HMD} \,/\, _nP_{x,i}^{HMD}
+\end{equation}
+
+\begin{equation}
+_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_{i,s}} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
+\end{equation}
+
+where $D_{x,i}$ is the number of deaths in age group $x$ in country $i$ from the HMD, $P_{x,i}$ is the population in age group $x$ from the HMD, $\hat{D_{i,s}}$ is the number of deaths from Forzieri et al.
+[@forzieri2017increasing] under scenario s, and $t_{x,i}$ is the proportion of mortality experienced in each age group $x$ from the GBD.
+Thus, the anticipated additional mortality due to climate change is added to each age group based on the underlying cause-specific mortality schedule observed between 2006-2015 in the GBD.
+
+We then calculated $q_x$ values, or the probability of dying, for each scenario $s$ for each country.
+
+\begin{equation}
+_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}
+\end{equation}
+
+\begin{equation}
+_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}
+\end{equation}
+
+We calculated each additional life table value identically for each scenario $s$ using standard life table equations:
+
+$$_nd_{x,i,s} = _nl_{x,i,s} \cdot _nq_{x,i,s}$$
+$$_nl_{x,i,s} = _nl_{x-1,i,s} - _nd_{x-1,i,s}$$
+$$_nL_{x,i,s} = _na_{x,i,s} \cdot _nl_{x,i,s} + ((n-_na_{x,i,s}) \cdot _nl_{x,i,s})$$
+$$_nT_{x,i,s} = _nL_{x,i,s} + _nT_{x+1,i,s}$$
+$$_ne_{x,i,s} = _nT_{x,i,s} / _nl_{x,i,s}$$
+
+To determine the differences between life expectancy compared to the baseline, we simply subtract $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].
+
+# Results
+We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` to `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**).
+This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States for the year 2000.
 Although the average European country could see a change of `r round(mean(lt_lancet$DIF_MID),2)` years, several countries are likely to experience considerably greater reductions in life expectancy.
@@
 
-```{r figure1, echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change.
+```{r figure1, echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change in the 2080s compared to the present $e_0$.
 We report changes in life expectancies due to climate change for twenty-eight European countries.
@@
 
-Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies.
+Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies (\autoref{map}).
 Whereas many Northern European countries could experience negligible impacts on life expectancy, five European countries could see life expectancy changes in excess of -0.5 years (Spain, Luxembourg, France, Italy, and Ireland).
@@
 
-data(Europe)
+# Europe <- shapefile("data/borders/TM_WORLD_BORDERS-0.3.SHP")
+load(file= "data/Europe.rda")
 Europe2 <- append_data(Europe, lt_lancet, key.shp = "iso_a3", key.data = "ISO3", ignore.na = TRUE)
 
-bbeur <- bb(Europe2, xlim = c(0,0.75), ylim=c(0,1), relative = TRUE)
+# bbeur <- bb(Europe2,  xlim=c(.42, .64), ylim=c(.7, 0.93), relative = TRUE)
 
@@
   tm_borders(alpha = 0.5) +  
-  tm_text("Test" , size="AREA", root=5) +
-  tm_format_Europe() 
+  tm_text("Test" , size="AREA", root=5, legend.size.show= F) +
+  tm_legend(legend.frame=T) +
+    NULL
 ```
@@
 
-<!-- # Discussion -->
+# Discussion
 In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries.
@@
 Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes.
-Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].
+Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].
 
@@
 \begin{sidewaystable}
+
 \caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1}
@@
 
+These results should also be considered conservative when compared to the broader impact of climate change on human longevity.
+The disaster databases that Forzieri et al [@forzieri2017increasing] use in generating their excess mortality estimates probably, though the disaster databases are unclear, only account for the deaths certified as *directly* caused by these hazards and are unlikely to capture the overall numbers associated with these deaths.
+If the certification of deaths due to weather extremes is similar to the certication in the present, than our results and those of Forzieri et al reflect the exess mortality directly attributable to climatic extremes.
+Despite this limitation, the impact of these extremes on $e_0$ is considerable, even if conservative.
+
 Finally, we would like to point out that future research should not only transform excess mortality into life expectancy decrements.
@@
 \newpage
-
-# Methods
-## Data
-To create life expectancy differentials, we used three sources of data.
-The first is the baseline life table data.
-Baseline life tables comes from the Human Mortality Database (HMD) [@HMD].
-These datasets comprise only official vital event registration and are considered the most complete and accurate source of mortality data available.
-We downloaded the most recent 5x1 (by age and year) life table data for twenty-eight European countries.
-We used the following data elements from the five-year age group life table data in the HMD:
-
-\begin{enumerate}
-\item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
-\item $_na_x$ or the average length of survival between ages $x$ and $x+n$.
-\item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
-\end{enumerate}
-
-The second dataset comes from published projections of climate-related mortality in Europe [@forzieri2017increasing].
-Forzieri et al.
-[@forzieri2017increasing] combined data from the International Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool (NatCatSERVICE) to create exposure and fatality statistics related to six weather-related hazards -- heatwaves, cold waves, droughts, wildfires, river and coastal floods, and windstorms.
-They downscaled these exposures to  1km grid scales and integrated them into small-area demographic projections for Europe out to 2100.
-Heatwaves and coldwaves account for the vast majority of projected fatalities.
-The number of future anticipated deaths are available in their Supplementary Table S8 [@forzieri2017increasing] and provide the magnitude of deaths for our study.
-
-The third dataset comes from the GBD [@GBD;@wang2012age].
-The GBD is the most comprehensive worldwide dataset of epidemiological data produced and provides cause-specific mortality by age/sex/geography/year for 249 causes of death in 195 countries and territories.
-We gathered data on age-specific deaths and mortality rates ($m_x$) for mortality from environmental heat and cold exposure (cause C.2.9) for the period 2006-2015 for each country in the study.
-These data provide a mortality schedule to fit the projected climate change excess mortality from Fozieri et al [@forzieri2017increasing].
-
-```{r suppfig, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("**Mortality distribution of Heat-related mortality in 28 European countries**.
-The thick black line is the average\\label{suppfig}")}
-GBD_data <- read.csv("data/GBD_data.csv") %>%
-  #mutate(perdist = log(val)) %>%
-  group_by(countrycode, Age) %>%
-  summarise(meanval = mean(val),
-            upperval = mean(upper),
-            lowerval = mean(lower)) %>%
-  group_by(countrycode) %>%
-  mutate(countmeanval = sum(meanval),
-         countupperval = sum(upperval),
-         countlowerval = sum(lowerval),
-         perdist = (meanval / countmeanval)) %>%
-  rename(CNTRY = countrycode)
-
-ggplot(data=GBD_data, aes(x=Age, y=perdist)) +
-  geom_line(aes(color=CNTRY), alpha=1) +
-  geom_smooth(span=0.3, se=F, col="black", lwd =2) +
-  theme_bw() +
-  theme(legend.position = c(0.8,0.25),
-        legend.title=element_blank(),
-        legend.text=element_text(size=4),
-        legend.key.size = unit(1,"line")) +
-  guides(col = guide_legend(ncol = 4)) +
-  scale_y_log10()+
-  labs(y="log(Mortality Percentage Distribution)",
-       #title = "Mortality distribution of Heat-related mortality in 28 European countries",
-       caption = "Source: Global Burden of Disease")
-
-```
-
-## Methods
-We derived a number of additional variables to accomplish our analysis.
-First, we abridged the life table data from the HMD from ages 100+ to 80+ to conform to the GBD cause-specific mortality schedules.
-
-From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}$).
-
-We then derived additional $m_x$ rates for each age group $x$ in each country $i$ for each scenario $s$ ($BASE$, $LOW$, $MID$, $HIGH$) from Forzieri et al [@forzieri2017increasing].
-
-\begin{equation}
-_nm_{x,i,BASE} = _nD_{x,i}^{HMD} \,/\, _nP_{x,i}^{HMD}
-\end{equation}
-
-\begin{equation}
-_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_i} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
-\end{equation}
-
-where $D_{x,i}$ is the number of deaths in age group $x$ in country $i$ from the HMD, $P_{x,i}$ is the population in age group $x$ from the HMD, $\hat{D_x}$ is the number of deaths from Forzieri et al.
-[@forzieri2017increasing], and $t_{x,i}$ is the proportion of mortality experienced in each age group $x$ from the GBD.
-Thus, the anticipated additional mortality due to climate change is added to each age group based on the underlying mortality cause-specific mortality schedule observed between 2006-2015 in the GBD.
-
-We then calculated $q_x$ values, or the probability of dying, for each scenario $s$ for each country.
-
-\begin{equation}
-_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}
-\end{equation}
-
-\begin{equation}
-_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}
-\end{equation}
-
-We calculated each additional life table value identically for each scenario $s$ using standard life table equations:
-
-$$_nd_{x,i,s} = _nl_{x,i,s} \cdot _nq_{x,i,s}$$
-$$_nl_{x,i,s} = _nl_{x-1,i,s} - _nd_{x-1,i,s}$$
-$$_nL_{x,i,s} = _na_{x,i,s} \cdot _nl_{x,i,s} + ((n-_na_{x,i,s}) \cdot _nl_{x,i,s})$$
-$$_nT_{x,i,s} = _nL_{x,i,s} + _nT_{x+1,i,s}$$
-$$_ne_{x,i,s} = _nT_{x,i,s} / _nl_{x,i,s}$$
-
-To determine the differences between life expectancy compared to the baseline, we simply subtract $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].

### 2018-08-22 MS-cclifeexpec.tex
New branch

@@
 
-\title{Climate change projected to reduce European life expectancy \thanks{All data and code that supports these conclusions are available as
+\title{Inaction on climate change could approach one year of life in some
+European countries \thanks{All data and code that supports these conclusions are available as
 supplementary materials.}  }
@@
 \usepackage{natbib}
-\bibliographystyle{nature}
+\bibliographystyle{apsr}
 \usepackage[strings]{underscore} % protect underscores in most circumstances
@@
 mortality, Europe, Public health},  
-            pdftitle={Climate change projected to reduce European life expectancy},
+            pdftitle={Inaction on climate change could approach one year of life in some
+European countries},
             colorlinks=true,
@@
 the dramatic impact on public health and human mortality from climate
-change
-\citep{patz2005impact, aastrom2013attributing, forzieri2017increasing, staddon2014climate}.
-However, life expectancy at birth is more easily communicated and
-understood by the public, and controls for different geographic,
-economic, and demographic groups \citep{parrish2010peer}. No studies
-have comprehensively connected excess mortality due to climate change to
-anticipated reductions in life expectancy. Without properly situating
-the potential loss of life within the contexts of life expectancy, we
-risk misrepresenting cost of climate change on mortality. In this paper,
-we convert excess mortality estimates into potential reductions of life
-expectancy at birth and report the cost of climate change on human
-longevity. We find climate change could become the third largest life
-expectancy reducer behind heart disease and cancer in some European
-countries. Thus, the cost of inaction on climate change could approach
-one year of life in some countries.
+change. However, life expectancy at birth is more easily communicated
+and understood by the public, and controls for different geographic,
+economic, and demographic groups. No studies have comprehensively
+connected excess mortality due to climate change to anticipated
+reductions in life expectancy. Without properly situating the potential
+loss of life within the contexts of life expectancy, we risk
+misrepresenting cost of climate change on mortality. In this paper, we
+convert excess mortality estimates due to increaes in extreme weather
+from climate change into potential reductions of life expectancy at
+birth and report the cost of climate change on human longevity. We find
+climate change extremes could become the third largest life expectancy
+reducer behind heart disease and cancer in some European countries.
+Thus, the cost of inaction on climate change could approach one year of
+life in some countries and possibly exceed that.
 
@@
 
-\textsuperscript{1} Carl Vinson Institute of Government, University of
-Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.
+\textsuperscript{1} Department of Sociology, Florida State University.
+201 N. Milledge Ave. Athens, GA USA 30602.
 
@@
 
-\section{Main Text}\label{main-text}
+\hypertarget{main-text}{%
+\section{Main Text}\label{main-text}}
 
@@
 
-Using published data on excess mortality\citep{forzieri2017increasing},
+Using published data on excess mortality \citep{forzieri2017increasing},
 we connect climate change excess mortality to life expectancy in a
@@
 change related mortality compare to other mortality vectors? Our results
-can situate climate change mortality within the broader context of human
+situate climate change mortality within the broader context of human
 mortality and can be used to inform public health interventions to
 prevent such futures.
+
+Our results suggest that climate change could emerge as a potent new
+mortality vector in the 20th century. Mitigation strategies (ie
+reductions in greenhouse gas emisions) and adaptation strategies (ie
+outreach efforts, retrofitting public buildings, etc.) would help
+prevent this new public health concern.
+
+\hypertarget{background}{%
+\section{Background}\label{background}}
+
+\hypertarget{climate-change-and-human-mortality}{%
+\subsection{Climate change and human
+mortality}\label{climate-change-and-human-mortality}}
+
+\hypertarget{methods-and-materials}{%
+\section{Methods and Materials}\label{methods-and-materials}}
 
@@
 rectified, high-resolution gridded datasets. While the disaster data are
-not standardized across country, Forzeri et al. imputed data in
+not standardized across country, Forzeri et al.~imputed data in
 incomplete time periods and countries. This imputation could mask
@@
 
+\begin{table}
+\caption{\textbf{Most recent mortality data available by country in the HMD.}} \label{Table2}
+\begin{tabular}{ll|ll|ll|ll}
+Country & Year & Country & Year & Country & Year & Country & Year \\
+\hline
+AUT     & 2014 & ESP     & 2014 & HUN     & 2014 & NLD     & 2016 \\
+BEL     & 2015 & EST     & 2014 & IRL     & 2014 & NOR     & 2014 \\
+BGR     & 2010 & FIN     & 2015 & ISL     & 2016 & POL     & 2014 \\
+CHE     & 2016 & FRATNP  & 2015 & ITA     & 2014 & PRT     & 2015 \\
+CZE     & 2016 & GBR     & 2016 & LTU     & 2014 & SVK     & 2014 \\
+DEUT    & 2015 & GRC     & 2013 & LUX     & 2014 & SVN     & 2014 \\
+DNK     & 2016 & HRV     & 2016 & LVA     & 2014 & SWE     & 2016 \\
+\hline
+\end{tabular}
+\end{table}
+
 We then recalculated age-specific mortality rates in five-year life
@@
 \(e_0\) between the 2080s and the most recent complete life table in the
-HMD. Complete life tables for all countries are available in the
-Supplementary Dataset and all equations are located in the Supplementary
-Information.
+HMD (see \textbf{Table 1} for the years used). Complete life tables for
+all countries are available in the \emph{Supplementary Dataset} and all
+computer code to replicate our results are available online\footnote{Replication
+  files are located here:
+  \url{https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1}}.
+
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
+To create life expectancy differentials, we used three sources of data.
+The first is the baseline life table data. Baseline life tables comes
+from the Human Mortality Database (HMD) \citep{HMD}. These datasets
+comprise only official vital event registration and are considered the
+most complete and accurate source of mortality data available. We
+downloaded the most recent 5x1 (by age and year) life table data for
+twenty-eight European countries. We used the following data elements
+from the five-year age group life table data in the HMD:
+
+\begin{enumerate}
+\item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
+\item $_na_x$ or the average length of survival between ages $x$ and $x+n$.
+\item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
+\end{enumerate}
+
+The second dataset comes from published projections of climate-related
+mortality in Europe \citep{forzieri2017increasing}. Forzieri et al.
+\citep{forzieri2017increasing} combined data from the International
+Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool
+(NatCatSERVICE) to create exposure and fatality statistics related to
+six weather-related hazards -- heatwaves, cold waves, droughts,
+wildfires, river and coastal floods, and windstorms. They downscaled
+these exposures to 1km grid scales and integrated them into small-area
+demographic projections for Europe out to 2100. Heatwaves and coldwaves
+account for the vast majority of projected fatalities. The number of
+future anticipated deaths are available in their Supplementary Table S8
+\citep{forzieri2017increasing} and provide the magnitude of deaths for
+our study.
+
+The third dataset comes from the GBD \citep{GBD, wang2012age}. The GBD
+is the most comprehensive worldwide dataset of epidemiological data
+produced and provides cause-specific mortality by age/sex/geography/year
+for 249 causes of death in 195 countries and territories. We gathered
+data on age-specific deaths and mortality rates (\(m_x\)) for mortality
+from environmental heat and cold exposure (cause C.2.9) for the period
+2006-2015 for each country in the study. These data provide a mortality
+schedule to fit the projected climate change excess mortality from
+Fozieri et al \citep{forzieri2017increasing}. Thus, we assume that the
+age-specific mortality schedules observed between 2006 and 2015 due to
+environmental heat and cold exposure in the GBD (see \autoref{suppfig})
+are likely to remain unchanged in the future.
+
+\begin{figure}
+\centering
+\includegraphics{MS-cclifeexpec_files/figure-latex/suppfig-1.pdf}
+\caption{\textbf{Mortality distribution of Heat-related mortality in 28
+European countries}. The thick black line is the average\label{suppfig}}
+\end{figure}
+
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
+
+We derived a number of additional variables to accomplish our analysis.
+First, we abridged the life table data from the HMD from ages 100+ to
+80+ to conform to the GBD cause-specific mortality schedules.
+
+From the GBD data, we derived variable \(t_{x,i}\) as the proportion of
+deaths, \(D\), from each age group \(x\) in each country \(i\)
+(\(_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_{\alpha,i}^{GBD}}\)).
+
+We then derived additional \(m_x\) rates for each age group \(x\) in
+each country \(i\) for each scenario \(s\) (\(BASE\), \(LOW\), \(MID\),
+\(HIGH\)) from Forzieri et al \citep{forzieri2017increasing}.
+
+\begin{equation}
+_nm_{x,i,BASE} = _nD_{x,i}^{HMD} \,/\, _nP_{x,i}^{HMD}
+\end{equation}
+
+\begin{equation}
+_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_{i,s}} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
+\end{equation}
+
+where \(D_{x,i}\) is the number of deaths in age group \(x\) in country
+\(i\) from the HMD, \(P_{x,i}\) is the population in age group \(x\)
+from the HMD, \(\hat{D_{i,s}}\) is the number of deaths from Forzieri et
+al. \citep{forzieri2017increasing} under scenario s, and \(t_{x,i}\) is
+the proportion of mortality experienced in each age group \(x\) from the
+GBD. Thus, the anticipated additional mortality due to climate change is
+added to each age group based on the underlying cause-specific mortality
+schedule observed between 2006-2015 in the GBD.
+
+We then calculated \(q_x\) values, or the probability of dying, for each
+scenario \(s\) for each country.
+
+\begin{equation}
+_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}
+\end{equation}
+
+\begin{equation}
+_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}
+\end{equation}
+
+We calculated each additional life table value identically for each
+scenario \(s\) using standard life table equations:
+
+\[_nd_{x,i,s} = _nl_{x,i,s} \cdot _nq_{x,i,s}\]
+\[_nl_{x,i,s} = _nl_{x-1,i,s} - _nd_{x-1,i,s}\]
+\[_nL_{x,i,s} = _na_{x,i,s} \cdot _nl_{x,i,s} + ((n-_na_{x,i,s}) \cdot _nl_{x,i,s})\]
+\[_nT_{x,i,s} = _nL_{x,i,s} + _nT_{x+1,i,s}\]
+\[_ne_{x,i,s} = _nT_{x,i,s} / _nl_{x,i,s}\]
+
+To determine the differences between life expectancy compared to the
+baseline, we simply subtract \(e_{x,i,s}\) (for scenario \(LOW\),
+\(MID\), and \(HIGH\)) from \(e_{x,i,BASE}\) as described by
+\citep{beltran2008integrated}.
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 due to influenza and pneumonia \citep{arias2013united} in the United
-States. Although the average European country could see a change of
--0.23 years, several countries are likely to experience considerably
-greater reductions in life expectancy. Luxembourg could see a reduction
-of up to 2.2 years and the medium-variant of climate hazards could cost
-the average Spaniard 1.03 years of life.
+States for the year 2000. Although the average European country could
+see a change of -0.23 years, several countries are likely to experience
+considerably greater reductions in life expectancy. Luxembourg could see
+a reduction of up to 2.2 years and the medium-variant of climate hazards
+could cost the average Spaniard 1.03 years of life.
 
@@
 \caption{Change in life expectancy at birth (\(e_0\)) due to
-business-as-usual climate change. We report changes in life expectancies
-due to climate change for twenty-eight European countries. The central
-values represent the ensemble median while the stems represent the upper
-and lower bounds of the inter-model climate variability.\label{figure1}}
+business-as-usual climate change in the 2080s compared to the present
+\(e_0\). We report changes in life expectancies due to climate change
+for twenty-eight European countries. The central values represent the
+ensemble median while the stems represent the upper and lower bounds of
+the inter-model climate variability.\label{figure1}}
 \end{figure}
@@
 Our results also suggest climate change mortality differentials are
-likely to unfold along highly uneven geographies. Whereas many Northern
-European countries could experience negligible impacts on life
-expectancy, five European countries could see life expectancy changes in
-excess of -0.5 years (Spain, Luxembourg, France, Italy, and Ireland).
-This group of countries could see life expectancy changes of more than
--1.0 years if climate change hazards are more intense than anticipated.
-In some countries, climate change could thus become a bigger killer than
-trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute
-myocardial infarction (-0.87 years), or all accident related mortality
-(-0.84 years) \citep{arias2013united} by the end of the century.
+likely to unfold along highly uneven geographies (\autoref{map}).
+Whereas many Northern European countries could experience negligible
+impacts on life expectancy, five European countries could see life
+expectancy changes in excess of -0.5 years (Spain, Luxembourg, France,
+Italy, and Ireland). This group of countries could see life expectancy
+changes of more than -1.0 years if climate change hazards are more
+intense than anticipated. In some countries, climate change could thus
+become a bigger killer than trachea, bronchus, and lung malignant
+neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or
+all accident related mortality (-0.84 years) \citep{arias2013united} by
+the end of the century.
 
@@
 
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
 In this article, we demonstrate the impact climate change could have on
@@
 the extent to which climate change could reduce the average person's
-longevity by the end of the century, the first study of its kind,
-significantly expanding our understanding of climate change and public
-health; thus linking two of the major areas in current developmental and
-sustainability discussion at both national and international levels
-\citep{abel2016meeting}.
-
-We find that climate change could alter life expectancy by -0.23 years
-(-2.5 months (-0.1 - -0.39 years)) in the average European country
-(\textbf{Fig. 1}). Although the average European country could see a
-change of -0.22 years, several countries are likely to experience
-considerably greater reductions in life expectancy. Luxembourg could see
-a change of up to -2.2 years and the medium-variant of climate hazards
-could cost the average Spaniard 0.98 years of life.
-
-The overall reduction in \(e_0\) is comparable to mortality due to
-influenza and pneumonia \citep{arias2013united} in the United States.
-\textbf{Table 1} reports the ranking of Climate Change compared to the
-major causes of death (COD) in European countries using age-standardised
-death rates and the standard European population
-\citep{pace2013revision} for the year 2013. In three European countries
-(Spain, Luxembourg, Italy), Climate Change could become a top three COD.
-In Spain (high = 101.62), Luxembourg (high = 160.93), France (high =
-93.45), and Italy (high = 68.21) Climate Change could become the leading
-cause of death if Climate Change is more extreme than anticipated. Only
-all circulatory diseases combined and all cancers combined would cause
-more deaths.
-
-Our results also suggest climate change mortality differentials are
-likely to unfold along highly uneven geographies. Whereas many Northern
-European countries could experience negligible impacts on life
-expectancy, five European countries could see life expectancy changes in
-excess of -0.5 years (Luxembourg, Spain, Italy, France, and Ireland).
-This group of countries could see life expectancy changes of more than
--1.0 years if climate change hazards are more intense than anticipated.
-In some countries, climate change could thus become a bigger killer than
-trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute
-myocardial infarction (-0.87 years), or all accident related mortality
-(-0.84 years) \citep{arias2013united} by the end of the century.
-
-While Forzieri et al's \citep{forzieri2017increasing} results suggest
-that the greatest climate change related mortality will unfold along a
-north-south gradient, we find the greatest reduction in life expectancy
-unfolds along an east-west gradient (\textbf{Fig. 2}). The most westerly
-European countries tend to have the greatest reductions. Neighboring
-pairs of countries at similar latitudes could experience vastly
-different mortality regimes, but neighboring pairs of countries at
-similar longitudes seem to experience more similar mortality regimes
-from climate change. Emergent research in Atmospheric Rivers suggest
-Western Europe could be more susceptible to these types of events
-\citep{ramos2015daily}. Additionally, these results point to the
-importance of converting excess mortality into life expectancy to
-properly quantify the effects of climate change mortality.
-
-In this article, we demonstrate the impact climate change could have on
-life expectancy at birth in twenty-eight European countries. Previous
-studies on climate change and excess mortality potentially misrepresent
-the impact climate change could have on human mortality. Contrary to
-excess mortality estimates, life expectancy is routinely used as a
-primary metric for communicating overall health outcomes and enjoys
-widespread use by major international organizations
-\citep{world2015world, marmot2012building, salomon2012healthy}. Life
-expectancy and its derivatives are the recommended metrics for
-population health \citep{parrish2010peer}. Additionally, it connects
-mortality estimates into intuitively understandable metrics, translating
-global estimates of mortality into individual outcomes. Our work reveals
-the extent to which climate change could reduce the average person's
-longevity by the end of the century, the first study of its kind,
-significantly expanding our understanding of climate change and public
-health; thus linking two of the major areas in current developmental and
-sustainability discussion at both national and international levels
-\citep{abel2016meeting}.
+longevity by the end of the century, expanding our understanding of
+climate change and public health; thus linking two of the major areas in
+current developmental and sustainability discussion at both national and
+international levels \citep{abel2016meeting}.
 
@@
 
+These results should also be considered conservative when compared to
+the broader impact of climate change on human longevity. The disaster
+databases that Forzieri et al \citep{forzieri2017increasing} use in
+generating their excess mortality estimates probably, though the
+disaster databases are unclear, only account for the deaths certified as
+\emph{directly} caused by these hazards and are unlikely to capture the
+overall numbers associated with these deaths. If the certification of
+deaths due to weather extremes is similar to the certication in the
+present, than our results and those of Forzieri et al reflect the exess
+mortality directly attributable to climatic extremes. Despite this
+limitation, the impact of these extremes on \(e_0\) is considerable,
+even if conservative.
+
 Finally, we would like to point out that future research should not only
@@
 \newpage
-
-\section{Methods}\label{methods}
-
-\subsection{Data}\label{data}
-
-To create life expectancy differentials, we used three sources of data.
-The first is the baseline life table data. Baseline life tables comes
-from the Human Mortality Database (HMD) \citep{HMD}. These datasets
-comprise only official vital event registration and are considered the
-most complete and accurate source of mortality data available. We
-downloaded the most recent 5x1 (by age and year) life table data for
-twenty-eight European countries. We used the following data elements
-from the five-year age group life table data in the HMD:
-
-\begin{enumerate}
-\item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
-\item $_na_x$ or the average length of survival between ages $x$ and $x+n$.
-\item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
-\end{enumerate}
-
-The second dataset comes from published projections of climate-related
-mortality in Europe \citep{forzieri2017increasing}. Forzieri et al.
-\citep{forzieri2017increasing} combined data from the International
-Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool
-(NatCatSERVICE) to create exposure and fatality statistics related to
-six weather-related hazards -- heatwaves, cold waves, droughts,
-wildfires, river and coastal floods, and windstorms. They downscaled
-these exposures to 1km grid scales and integrated them into small-area
-demographic projections for Europe out to 2100. Heatwaves and coldwaves
-account for the vast majority of projected fatalities. The number of
-future anticipated deaths are available in their Supplementary Table S8
-\citep{forzieri2017increasing} and provide the magnitude of deaths for
-our study.
-
-The third dataset comes from the GBD \citep{GBD, wang2012age}. The GBD
-is the most comprehensive worldwide dataset of epidemiological data
-produced and provides cause-specific mortality by age/sex/geography/year
-for 249 causes of death in 195 countries and territories. We gathered
-data on age-specific deaths and mortality rates (\(m_x\)) for mortality
-from environmental heat and cold exposure (cause C.2.9) for the period
-2006-2015 for each country in the study. These data provide a mortality
-schedule to fit the projected climate change excess mortality from
-Fozieri et al \citep{forzieri2017increasing}.
-
-\begin{figure}
-\centering
-\includegraphics{MS-cclifeexpec_files/figure-latex/suppfig-1.pdf}
-\caption{\textbf{Mortality distribution of Heat-related mortality in 28
-European countries}. The thick black line is the average\label{suppfig}}
-\end{figure}
-
-\subsection{Methods}\label{methods-1}
-
-We derived a number of additional variables to accomplish our analysis.
-First, we abridged the life table data from the HMD from ages 100+ to
-80+ to conform to the GBD cause-specific mortality schedules.
-
-From the GBD data, we derived variable \(t_{x,i}\) as the proportion of
-deaths, \(D\), from each age group \(x\) in each country \(i\)
-(\(_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}\)).
-
-We then derived additional \(m_x\) rates for each age group \(x\) in
-each country \(i\) for each scenario \(s\) (\(BASE\), \(LOW\), \(MID\),
-\(HIGH\)) from Forzieri et al \citep{forzieri2017increasing}.
-
-\begin{equation}
-_nm_{x,i,BASE} = _nD_{x,i}^{HMD} \,/\, _nP_{x,i}^{HMD}
-\end{equation}
-
-\begin{equation}
-_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_i} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
-\end{equation}
-
-where \(D_{x,i}\) is the number of deaths in age group \(x\) in country
-\(i\) from the HMD, \(P_{x,i}\) is the population in age group \(x\)
-from the HMD, \(\hat{D_x}\) is the number of deaths from Forzieri et al.
-\citep{forzieri2017increasing}, and \(t_{x,i}\) is the proportion of
-mortality experienced in each age group \(x\) from the GBD. Thus, the
-anticipated additional mortality due to climate change is added to each
-age group based on the underlying mortality cause-specific mortality
-schedule observed between 2006-2015 in the GBD.
-
-We then calculated \(q_x\) values, or the probability of dying, for each
-scenario \(s\) for each country.
-
-\begin{equation}
-_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}
-\end{equation}
-
-\begin{equation}
-_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}
-\end{equation}
-
-We calculated each additional life table value identically for each
-scenario \(s\) using standard life table equations:
-
-\[_nd_{x,i,s} = _nl_{x,i,s} \cdot _nq_{x,i,s}\]
-\[_nl_{x,i,s} = _nl_{x-1,i,s} - _nd_{x-1,i,s}\]
-\[_nL_{x,i,s} = _na_{x,i,s} \cdot _nl_{x,i,s} + ((n-_na_{x,i,s}) \cdot _nl_{x,i,s})\]
-\[_nT_{x,i,s} = _nL_{x,i,s} + _nT_{x+1,i,s}\]
-\[_ne_{x,i,s} = _nT_{x,i,s} / _nl_{x,i,s}\]
-
-To determine the differences between life expectancy compared to the
-baseline, we simply subtract \(e_{x,i,s}\) (for scenario \(LOW\),
-\(MID\), and \(HIGH\)) from \(e_{x,i,BASE}\) as described by
-\citep{beltran2008integrated}.
 \newpage

### 2018-08-22 R/000-libraries.R

@@ -1,0 +1,71 @@
+######################################################
+##
+## Inaction on climate change could approach one year of life in some European countries
+##       Prepare R Session
+##
+## Authors: Mathew Hauer, mehauer@fsu.edu
+##          
+## Last updated: 08/22/2018
+##
+######################################################
+
+
+#Clean Up Workspace 
+rm(list = ls()) # Remove Previous Workspace
+gc(reset = TRUE) # Garbage Collection
+
+
+# R Workspace Options
+options(scipen = 12) # Scientific Notation
+options(digits = 6) # Specify Digits
+options(java.parameters = "-Xmx1000m") # Increase Java Heap Size
+
+
+#Functions, Libraries, & Parallel Computing 
+## Functions 
+# Specify Number of Digits (Forward)
+numb_digits_F <- function(x,y){
+  numb_digits_F <- do.call("paste0", list(paste0(rep(0, y - nchar(x)), collapse = ""), x))
+  numb_digits_F <- ifelse(nchar(x) < y, numb_digits_F, x)
+}
+
+# Remove Double Space 
+numb_spaces <- function(x) gsub("[[:space:]]{2,}", " ", x)
+
+
+# Install and load pacman for easier package loading and installation
+if (!require("pacman", character.only = TRUE)){
+  install.packages("pacman", dep = TRUE)
+  if (!require("pacman", character.only = TRUE))
+    stop("Package not found")
+}
+
+
+# Libraries
+pkgs <- c(
+  "tidyverse",   # Tidyverse
+  "data.table",  # Data Management/Manipulation
+  "openxlsx",    # Microsoft Excel Files
+  "HMDHFDplus",  # HMD data
+  "tmap",        # Mapping
+  "tmapptools",  # Mapping tools
+  "RColorBrewer",# Colors
+  "kableExtra",  # Table making
+  "knitr",       # Knitting
+  "xtable",      # Table making
+  "getPass"      # To get the password/ids
+  
+)
+
+# Install missing packages
+# Will only run if at least one package is missing
+
+if(!sum(!p_isinstalled(pkgs))==0){
+  p_install(
+    package = pkgs[!p_isinstalled(pkgs)], 
+    character.only = TRUE
+  )
+}
+
+# load the packages
+p_load(pkgs, character.only = TRUE)

### 2018-10-23 MS-cclifeexpec.Rmd
minor edits.

@@
   pdf_document:
-    citation_package: natbib
+    # citation_package: natbib
     keep_tex: true
     fig_caption: true
-    latex_engine: xelatex
+    latex_engine: pdflatex
     template: LATEX/svm-latex-ms.tex
-title: "Inaction on climate change could approach one year of life in some European countries"
-thanks: "All data and code that supports these conclusions are available as supplementary materials."
+title: "Inaction on climate change could cost one year of life in some European countries"
+thanks: "All data and code that supports these conclusions are available as supplementary materials. The authors would like to thank E. Carlson for his helpful comments."
 author:
-- name: Mathew E. Hauer^1^*
-  affiliation: University of Georgia
-- name: Alexis R. Santos^2^
+- name: Mathew E. Hauer^1,2^*
+  affiliation: Florida State University
+- name: Alexis R. Santos^3^
   affiliation: Pennsylvania State University
 
-abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change.
-However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups.
-No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy.
-Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting cost of climate change on mortality.
-In this paper, we convert excess mortality estimates due to increaes in extreme weather from climate change into potential reductions of life expectancy at birth and report the cost of climate change on human longevity.
-We find climate change extremes could become the third largest life expectancy reducer behind heart disease and cancer in some European countries.
-Thus, the cost of inaction on climate change could approach one year of life in some countries and possibly exceed that."
+abstract: "Climate change-related excess mortality estimates clearly demonstrate a dramatic impact on public health and human mortality.
+However, life expectancy at birth is more easily communicated and understood by the public.
+By properly situating climate change mortality within the contexts of life expectancy, we better represent the cost of climate change on longevity.
+In this paper, we convert excess mortality estimates due to increases in extreme weather from climate change into potential reductions in life expectancy at birth in twenty-eight European countries.
+We project climate change extremes to reduce life expectancy at birth by 0.24 years for the average European country with differences in excess of 1.0 years in some countries by 2100.
+We only estimate the impact of mortality directly related to climate extremes, making our estimates conservative.
+Thus, the cost of inaction on climate change could approach, and likely to exceed, one year of life in some European countries."
 
+# "Climate change-related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change.
+However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, age, and demographic groups.
+Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting the cost of climate change on mortality.
+In this paper, we convert excess mortality estimates due to increases in extreme weather from climate change into potential reductions of life expectancy at birth in twenty-eight European countries.
+We project climate change extremes to reduce life expectancy at birth by 0.24 years for the average European country but could exceed 1.0 years in some by 2100.
+These results are conservative when compared to the broader impact of climate change on human longevity as we only estimate the impact of mortality directly related to climate extremes.
+Thus, the cost of inaction on climate change could approach one year of life in some countries and likely to exceed that."
 keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
@@
 bibliography: LATEX/ccmort.bib
-biblio-style: apsr
+# biblio-style: apsr
+csl : LATEX/american-sociological-association.csl
 # use apsr or nature
@@
 
-* Corresponding author. hauer@uga.edu. p: 706-542-9369.
+* Corresponding author. mehauer@fsu.edu p: 850-644-7103.
 
-^1^ Department of Sociology, Florida State University. 201 N. Milledge Ave. Athens, GA USA 30602.
+^1^ Department of Sociology, Florida State University. 113 Colleggiate Way, Bellamy Building, Tallahassee FL 30602.
 
-^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.
+^2^ Center for Demography and Population Health, Florida State University.
+
+^3^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.
 
@@
 
-Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimating], estimates of displacement [@rigaud2018groundswell], or human conflict [@barnett2007climate]  but have the potential to lead to the loss of human life [@forzieri2017increasing;@pachauri2014climate].
-As impact quantification studies move further from the physical sciences of climate change into the social sciences, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].
+Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimating], estimates of displacement [@rigaud2018groundswell], or human conflict [@barnett2007climate]  but have the potential to contribute to the loss of human life [@forzieri2017increasing;@pachauri2014climate].
+As impact quantification studies move further from the physical sciences of climate change, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].
 
 Scholars have long estimated the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer].
-
-Although such estimates are useful, the excess mortality estimates are rather sterile -- one death is a tragedy, a million deaths a statistic -- and difficult to relate to on an individual level.
+For example, Forzieri et al.
+[-@forzieri2017increasing] projected climate change extremes to increase European mortality by 150,000+ persons per year by the end of the century.
+Although such estimates are useful, the excess mortality estimates are rather sterile -- one death is a tragedy, a million deaths a statistic -- and difficult to relate to on a personal level.
 Life expectancy at birth ($e_0$), on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone.
@@
 By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in twenty-eight European countries.
-This approach allows us to quantify the impact of climate change on human longevity and answer the following questions: What is the cost of inaction on climate change on human longevity?
-and How does climate change related mortality compare to other mortality vectors?
+This approach allows us to quantify the impact of climate change on human longevity and answer the following question: What is the cost of inaction on climate change on human longevity?
 Our results situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.
 
-Our results suggest that climate change could emerge as a potent new mortality vector in the 20th century.
+Our results suggest that climate change could emerge as a potent public health threat in the 20th century.
 Mitigation strategies (ie reductions in greenhouse gas emisions) and adaptation strategies (ie outreach efforts, retrofitting public buildings, etc.) would help prevent this new public health concern.
 
-# Background
-## Climate change and human mortality
+# Methods and Materials
 
-# Methods and Materials
+## Data
+
 We estimate changes in life expectancy by combining three primary datasets: excess mortality data due to meterological hazards related to climate change from Forzieri et al.
-[@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD].
+[-@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD].
+
 Forzieri et al.
-[@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario (SRES A1B).
-They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario (SSP2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario) and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases within spatially rectified, high-resolution gridded datasets.
+[-@forzieri2017increasing] combined data from the International Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool (NatCatSERVICE) to create exposure and fatality statistics related to six weather-related hazards -- heatwaves, cold waves, droughts, wildfires, river and coastal floods, and windstorms -- and estimated excess mortality associated with those hazards during a baseline period of 1981-2010 in 28 European countries.
+They only considered fatalities *directly attributable* to the event and *exclude* increased deaths from common causes that were observed to rise during the event, such as cardiovascular or respiratory deaths[^1].
 While the disaster data are not standardized across country, Forzeri et al.
@@
 This imputation could mask spatial variability at the sub-national level, but here we use their country-aggregated results.
-Their results represent business-as-usual climate change and human development without incorporating potential adaptation.
+They downscaled these exposures rates to 1km grid cells and integrated them into small-area demographic projections for Europe out to 2100 that correspond to a middle-of-the-road socioeconomic scenario (SSP2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario [@jiang2014internal;@samir2017human;@o2017roads]) and based their projected mortality on extrapolations of their calculated exposures.
+Their results represent business-as-usual climate change and human development without incorporating potential adaptations to climate change.
 They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes.
-These data provide the anticipated fatalities but not the effect these deaths might have on life expectancy.
-
-To convert excess mortality to life expectancy, we use data gathered from the HMD [@HMD] for corresponding European countries in five-year life tables in conjunction with cause-specific mortality data from the GBD [@GBD;@wang2012age].
-The HMD comprises only complete, official vital event statistics and represents the most accurate and complete compilation of mortality data available.
-We allocate excess multi-hazard mortality for each country from supplementary table S8 in Forzieri et al.
-[@forzieri2017increasing] based on the observed mortality schedule for environmental heat and cold exposure deaths from the GBD for each country in the previous decade.
-
-\begin{table}
-\caption{\textbf{Most recent mortality data available by country in the HMD.}} \label{Table2}
-\begin{tabular}{ll|ll|ll|ll}
-Country & Year & Country & Year & Country & Year & Country & Year \\
-\hline
-AUT     & 2014 & ESP     & 2014 & HUN     & 2014 & NLD     & 2016 \\
-BEL     & 2015 & EST     & 2014 & IRL     & 2014 & NOR     & 2014 \\
-BGR     & 2010 & FIN     & 2015 & ISL     & 2016 & POL     & 2014 \\
-CHE     & 2016 & FRATNP  & 2015 & ITA     & 2014 & PRT     & 2015 \\
-CZE     & 2016 & GBR     & 2016 & LTU     & 2014 & SVK     & 2014 \\
-DEUT    & 2015 & GRC     & 2013 & LUX     & 2014 & SVN     & 2014 \\
-DNK     & 2016 & HRV     & 2016 & LVA     & 2014 & SWE     & 2016 \\
-\hline
-\end{tabular}
-\end{table}
-
-We then recalculated age-specific mortality rates in five-year life tables by adding the climate change mortality onto the underlying mortality data and create four subsequent life tables ($baseline$, $low$, $mid$, $high$) using standard life table techniques [@wunsch2013life] based on the published confidence intervals from [@forzieri2017increasing].
-We used a variant of the cause-deleted life tables approach [@brand2005approximations;@beltran2008integrated] to measure the impact of excess mortality on life expectancy at birth or $e_0$.
-Changes in life expectancy are reported based on differentials from $e_0$ between the 2080s and the most recent complete life table in the HMD (see **Table 1** for the years used).
-Complete life tables for all countries are available in the *Supplementary Dataset* and all computer code to replicate our results are available online[^1].
-
-[^1]: Replication files are located here: <https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1>
-
-## Data
-To create life expectancy differentials, we used three sources of data.
-The first is the baseline life table data.
-Baseline life tables comes from the Human Mortality Database (HMD) [@HMD].
-These datasets comprise only official vital event registration and are considered the most complete and accurate source of mortality data available.
-We downloaded the most recent 5x1 (by age and year) life table data for twenty-eight European countries.
-We used the following data elements from the five-year age group life table data in the HMD:
-
-\begin{enumerate}
-\item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
-\item $_na_x$ or the average length of survival between ages $x$ and $x+n$.
-\item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
-\end{enumerate}
-
-The second dataset comes from published projections of climate-related mortality in Europe [@forzieri2017increasing].
-Forzieri et al.
-[@forzieri2017increasing] combined data from the International Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool (NatCatSERVICE) to create exposure and fatality statistics related to six weather-related hazards -- heatwaves, cold waves, droughts, wildfires, river and coastal floods, and windstorms.
-They downscaled these exposures to  1km grid scales and integrated them into small-area demographic projections for Europe out to 2100.
-Heatwaves and coldwaves account for the vast majority of projected fatalities.
 The number of future anticipated deaths are available in their Supplementary Table S8 [@forzieri2017increasing] and provide the magnitude of deaths for our study.
 
-The third dataset comes from the GBD [@GBD;@wang2012age].
-The GBD is the most comprehensive worldwide dataset of epidemiological data produced and provides cause-specific mortality by age/sex/geography/year for 249 causes of death in 195 countries and territories.
-We gathered data on age-specific deaths and mortality rates ($m_x$) for mortality from environmental heat and cold exposure (cause C.2.9) for the period 2006-2015 for each country in the study.
-These data provide a mortality schedule to fit the projected climate change excess mortality from Fozieri et al [@forzieri2017increasing].
-Thus, we assume that the age-specific mortality schedules observed between 2006 and 2015 due to environmental heat and cold exposure in the GBD (see \autoref{suppfig}) are likely to remain unchanged in the future.
+[^1]: "The fatality data from the two databases considered is likely to not include increased deaths from common causes that were observed to rise at the population level but for which individual deaths could not be attributed to the event.
+For example,...
+increased risk of cardiovascular and respiratory deaths ...
+may be severly underreported in the EMDAT and NatCatSERVICE." [see @forzieri2017increasing, pp.
+9, *Supplementary Materials*].
 
@@
 
+The Forzieri et al.
+[-@forzieri2017increasing] data contain only mortality totals and report no age detail, precluding a direct conversion from excess mortality to life expectancy.
+To convert excess mortality to life expectancy, we allocate excess multi-hazard mortality for each country based on the observed age-specific mortality schedule for environmental heat and cold exposure deaths from the GBD [@GBD;@wang2012age].
+Forzieri et al.
+found that extreme heat (as opposed to other five weather-related hazards) account for 99% of the anticipated excess mortality by the end of the century.
+The GBD is the most comprehensive worldwide dataset of epidemiological data produced and provides cause-specific mortality by age/sex/geography/year for 249 causes of death in 195 countries and territories.
+We gathered data on age-specific deaths and mortality rates ($m_x$) for mortality from environmental heat and cold exposure (cause C.2.9) for the period 2006-2015 for each country in the study.
+These data provide a mortality schedule to fit the projected climate change excess mortality from Fozieri et al [@forzieri2017increasing].
+Thus, we assume that the age-specific mortality schedules observed between 2006 and 2015 due to environmental heat and cold exposure in the GBD (see \autoref{suppfig}) are likely to remain unchanged in the future.
+We only use the GBD to distribute the Forzieri et al.
+deaths by age.
+
+With the Forzier et al data and the GBD we estimate the anticipated age-specific mortality schedules due to climate change in twenty-eight European countries but we cannot compare life expectancy differentials in the absence of Climate Change using those two datasets alone.
+To create baseline mortality schedules and life expectancy, we gather data from the Human Mortality Database [@HMD] for the corresponding European countries.
+The HMD comprises only complete, official vital event statistics and represents the most accurate and complete compilation of mortality data available.
+We downloaded the most recent 5x1 (age by year) life table data for the twenty-eight European countries (see \autoref{Table2} for a list of countries and most recent mortality data).
+
+\begin{table}
+\caption{\textbf{Most recent mortality data available by country in the HMD.}} \label{Table2}
+\begin{tabular}{ll|ll|ll|ll}
+Country & Year & Country & Year & Country & Year & Country & Year \\
+\hline
+AUT     & 2014 & ESP     & 2014 & HUN     & 2014 & NLD     & 2016 \\
+BEL     & 2015 & EST     & 2014 & IRL     & 2014 & NOR     & 2014 \\
+BGR     & 2010 & FIN     & 2015 & ISL     & 2016 & POL     & 2014 \\
+CHE     & 2016 & FRATNP  & 2015 & ITA     & 2014 & PRT     & 2015 \\
+CZE     & 2016 & GBR     & 2016 & LTU     & 2014 & SVK     & 2014 \\
+DEUT    & 2015 & GRC     & 2013 & LUX     & 2014 & SVN     & 2014 \\
+DNK     & 2016 & HRV     & 2016 & LVA     & 2014 & SWE     & 2016 \\
+\hline
+\end{tabular}
+\end{table}
+We used the following data elements from the five-year age group life table data in the HMD:
+
+\begin{enumerate}
+\item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
+\item $_na_x$ or the average length of survival between ages $x$ and $x+n$.
+\item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$.
+\end{enumerate}
+
 ## Methods
+
 We derived a number of additional variables to accomplish our analysis.
@@
 
-To determine the differences between life expectancy compared to the baseline, we simply subtract $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].
+To determine the differences between life expectancy compared to the baseline, we simply subtract $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].Complete life tables for all countries are available in the *Supplementary Dataset* and all computer code to replicate our results are available online[^2].
+
+[^2]: Replication files are located here: <https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1>
 
 # Results
-We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` to `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**).
+We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` to `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**) by 2100.
 This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States for the year 2000.
@@
 
-**\autoref{Table1}** reports the rankings of Age Standardised Death Rates (ASDR) for the leading causes of death in these European countries (ASDR's for other CODs come from Eurostat (online data code: hlth_cd_asdr2)) with our estimates of climate change included.
+<!-- **\autoref{Table1}** reports the rankings of Age Standardised Death Rates (ASDR) for the leading causes of death in these European countries (ASDR's for other CODs come from Eurostat (online data code: hlth_cd_asdr2)) with our estimates of climate change included.
 In many European countries, climate change could emerge as a top five killer.
 Only combined CODs, such as all circulatory diseases or all cancers, exhibit higher ASDRs if climatic hazards are more severe than anticipated for Spain, Luxembourg, and France.
+-->
 
@@
 # Discussion
-In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries.
+In this article, we demonstrate the impact climate change could have on life expectancy at birth due to extreme weather events in twenty-eight European countries.
 Previous studies on climate change and excess mortality potentially miscommunicate the impact climate change could have on human mortality.
@@
 Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer].
-Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes.
+Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into personably relatable outcomes.
 Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].
@@
 
-\begin{sidewaystable}
+<!-- \begin{sidewaystable} -->
 
-\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1}
-\rowcolors{2}{gray!25}{white}
-\small
-\begin{tabularx}{\textwidth}{l|XXXXXXXXX}
-  \hline
-   \rowcolor{white}
- Country & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\ 
-  \hline
-Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05\\\cellcolor{blue!25}\begin{tiny}(63.26 - 101.6)\end{tiny}} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
-of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\
-Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93\\\cellcolor{blue!25}\begin{tiny}(2.36 - 160.9)\end{tiny}} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
-of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\
-Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87\\\cellcolor{blue!25}\begin{tiny}(28.15 - 68.2)\end{tiny}} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
-of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\
-France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis.
-of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61\\\cellcolor{blue!25}\begin{tiny}(18.24 - 93.5)\end{tiny}} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\
-Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95\\\cellcolor{blue!25}\begin{tiny}(29.93 - 49.4)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
-of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\
-Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis.
-of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65\\\cellcolor{blue!25}\begin{tiny}(13.57 - 119.6)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\
-Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02\\\cellcolor{blue!25}\begin{tiny}(11.19 - 37.8)\end{tiny}} & \makecell{Dis.
-of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\
-Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44\\\cellcolor{blue!25}\begin{tiny}(13.32 - 38.6)\end{tiny}} & \makecell{Dis.
-of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\
-Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis.
-of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88\\\cellcolor{blue!25}\begin{tiny}(5.74 - 37.5)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\
-Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis.
-of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23\\\cellcolor{blue!25}\begin{tiny}(0.68 - 37.4)\end{tiny}} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\
-   \hline
-\end{tabularx}
+<!-- \caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1} -->
+<!-- \rowcolors{2}{gray!25}{white} -->
+<!-- \small -->
+<!-- \begin{tabularx}{\textwidth}{l|XXXXXXXXX} -->
+<!--   \hline -->
+<!--    \rowcolor{white} -->
+<!--  Country & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\  -->
+<!--   \hline -->
+<!-- Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 78.05\\\cellcolor{blue!25}\begin{tiny}(63.26 - 101.6)\end{tiny}} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
+of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\  -->
+<!-- Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 76.93\\\cellcolor{blue!25}\begin{tiny}(2.36 - 160.9)\end{tiny}} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
+of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\  -->
+<!-- Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 49.87\\\cellcolor{blue!25}\begin{tiny}(28.15 - 68.2)\end{tiny}} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
+of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\  -->
+<!-- France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis.
+of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 49.61\\\cellcolor{blue!25}\begin{tiny}(18.24 - 93.5)\end{tiny}} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\  -->
+<!-- Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 35.95\\\cellcolor{blue!25}\begin{tiny}(29.93 - 49.4)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
+of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\  -->
+<!-- Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis.
+of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 44.65\\\cellcolor{blue!25}\begin{tiny}(13.57 - 119.6)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\  -->
+<!-- Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 25.02\\\cellcolor{blue!25}\begin{tiny}(11.19 - 37.8)\end{tiny}} & \makecell{Dis.
+of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\  -->
+<!-- Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 24.44\\\cellcolor{blue!25}\begin{tiny}(13.32 - 38.6)\end{tiny}} & \makecell{Dis.
+of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\  -->
+<!-- Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis.
+of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 22.88\\\cellcolor{blue!25}\begin{tiny}(5.74 - 37.5)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\  -->
+<!-- Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis.
+of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 21.23\\\cellcolor{blue!25}\begin{tiny}(0.68 - 37.4)\end{tiny}} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\  -->
+<!--    \hline -->
+<!-- \end{tabularx} -->
 
-\end{sidewaystable}
+<!-- \end{sidewaystable} -->
 
@@
 Life expectancy has steadily risen across the world for the last century [@gerland2014world] and our results suggest that climate change alone could spur a sharp reversal in these trends in some countries.
-We expect two European countries to see life expectancy reductions more than one year under the middle scenario (Luxembourg and Spain), but if climate change has a greater impact on mortality than anticipated, five European countries (Luxembourg, Spain, Italy, France, Ireland) could see life expectancy reductions more than one year with Luxembourg experiencing a reduction of more than two.
-These findings highlight the hyperlocalized impacts of climate change [@kendon2014heavier;@rosenzweig2010cities;@forzieri2017increasing]
+We expect two European countries to see life expectancy reductions of more than one year under the middle scenario (Luxembourg and Spain), but if climate change has a greater impact on mortality than anticipated, five European countries (Luxembourg, Spain, Italy, France, Ireland) could see life expectancy reductions of more than one year with Luxembourg experiencing a reduction of more than two.
+These findings highlight the hyperlocalized impacts of climate change [@kendon2014heavier;@rosenzweig2010cities;@forzieri2017increasing].
 

### 2018-10-23 MS-cclifeexpec.tex
minor edits.

@@
 
-\title{Inaction on climate change could approach one year of life in some
-European countries \thanks{All data and code that supports these conclusions are available as
-supplementary materials.}  }
+\title{Inaction on climate change could cost one year of life in some European
+countries \thanks{All data and code that supports these conclusions are available as
+supplementary materials. The authors would like to thank E. Carlson for
+his helpful comments.}  }
 
 \author{\Large Mathew E.
-Hauer\textsuperscript{1}*\vspace{0.05in} \newline\normalsize\emph{University of Georgia}   \and \Large Alexis R.
-Santos\textsuperscript{2}\vspace{0.05in} \newline\normalsize\emph{Pennsylvania State University}  }
+Hauer\textsuperscript{1,2}*\vspace{0.05in} \newline\normalsize\emph{Florida State University}   \and \Large Alexis R.
+Santos\textsuperscript{3}\vspace{0.05in} \newline\normalsize\emph{Pennsylvania State University}  }
 
@@
 \titleformat*{\subparagraph}{\normalsize\itshape}
-
-\usepackage{natbib}
-\bibliographystyle{apsr}
-\usepackage[strings]{underscore} % protect underscores in most circumstances
 
@@
             bookmarks=true,
-            pdfauthor={Mathew E. Hauer\textsuperscript{1}* (University of Georgia) and Alexis R. Santos\textsuperscript{2} (Pennsylvania State University)},
+            pdfauthor={Mathew E. Hauer\textsuperscript{1,2}* (Florida State University) and Alexis R. Santos\textsuperscript{3} (Pennsylvania State University)},
              pdfkeywords = {Climate change, Life expectancy, Mortality, Demography, Excess
 mortality, Europe, Public health},  
-            pdftitle={Inaction on climate change could approach one year of life in some
-European countries},
+            pdftitle={Inaction on climate change could cost one year of life in some European
+countries},
             colorlinks=true,
@@
 \textbf{\authorfont Mathew E.
-Hauer\textsuperscript{1}*} \hskip 15pt \emph{\small University of Georgia}   \par \textbf{\authorfont Alexis R.
-Santos\textsuperscript{2}} \hskip 15pt \emph{\small Pennsylvania State University}
+Hauer\textsuperscript{1,2}*} \hskip 15pt \emph{\small Florida State University}   \par \textbf{\authorfont Alexis R.
+Santos\textsuperscript{3}} \hskip 15pt \emph{\small Pennsylvania State University}
 
@@
 
-\noindent Climate change related estimates of excess mortality clearly demonstrate
-the dramatic impact on public health and human mortality from climate
-change. However, life expectancy at birth is more easily communicated
-and understood by the public, and controls for different geographic,
-economic, and demographic groups. No studies have comprehensively
-connected excess mortality due to climate change to anticipated
-reductions in life expectancy. Without properly situating the potential
-loss of life within the contexts of life expectancy, we risk
-misrepresenting cost of climate change on mortality. In this paper, we
-convert excess mortality estimates due to increaes in extreme weather
-from climate change into potential reductions of life expectancy at
-birth and report the cost of climate change on human longevity. We find
-climate change extremes could become the third largest life expectancy
-reducer behind heart disease and cancer in some European countries.
-Thus, the cost of inaction on climate change could approach one year of
-life in some countries and possibly exceed that.
+\noindent Climate change-related excess mortality estimates clearly demonstrate a
+dramatic impact on public health and human mortality. However, life
+expectancy at birth is more easily communicated and understood by the
+public. By properly situating climate change mortality within the
+contexts of life expectancy, we better represent the cost of climate
+change on longevity. In this paper, we convert excess mortality
+estimates due to increases in extreme weather from climate change into
+potential reductions in life expectancy at birth in twenty-eight
+European countries. We project climate change extremes to reduce life
+expectancy at birth by 0.24 years for the average European country with
+differences in excess of 1.0 years in some countries by 2100. We only
+estimate the impact of mortality directly related to climate extremes,
+making our estimates conservative. Thus, the cost of inaction on climate
+change could approach, and likely to exceed, one year of life in some
+European countries.
 
@@
   Corresponding author.
-  \href{mailto:hauer@uga.edu}{\nolinkurl{hauer@uga.edu}}. p:
-  706-542-9369.
+  \href{mailto:mehauer@fsu.edu}{\nolinkurl{mehauer@fsu.edu}} p:
+  850-644-7103.
 \end{itemize}
@@
 \textsuperscript{1} Department of Sociology, Florida State University.
-201 N. Milledge Ave. Athens, GA USA 30602.
+113 Colleggiate Way, Bellamy Building, Tallahassee FL 30602.
 
-\textsuperscript{2} Department of Sociology and Criminology,
+\textsuperscript{2} Center for Demography and Population Health, Florida
+State University.
+
+\textsuperscript{3} Department of Sociology and Criminology,
 Pennsylvania State University. 901 Oswald Tower, University Park, PA USA
@@
 Climate change's implications on humanity go far beyond estimates of
-economic damages \citep{hsiang2017estimating}, estimates of displacement
-\citep{rigaud2018groundswell}, or human conflict
-\citep{barnett2007climate} but have the potential to lead to the loss of
-human life \citep{forzieri2017increasing, pachauri2014climate}. As
-impact quantification studies move further from the physical sciences of
-climate change into the social sciences, properly quantifying and
+economic damages (Hsiang et al. 2017), estimates of displacement (Rigaud
+et al. 2018), or human conflict (Barnett and Adger 2007) but have the
+potential to contribute to the loss of human life (Forzieri et al. 2017;
+Pachauri et al. 2014). As impact quantification studies move further
+from the physical sciences of climate change, properly quantifying and
 conveying the impact of climate change on public health is of increasing
-importance \citep{melillo2014climate, cloyd2016engagement}.
+importance (Cloyd et al. 2016; Melillo, Richmond, and Yohe 2014).
 
 Scholars have long estimated the mortality risks associated with climate
-change and typically use excess or extra mortality
-\citep{forzieri2017increasing, wilson2017climate, mcmichael2006climate, zanobetti2012summer}.
-Although such estimates are useful, the excess mortality estimates are
-rather sterile -- one death is a tragedy, a million deaths a statistic
--- and difficult to relate to on an individual level. Life expectancy at
-birth (\(e_0\)), on the other hand, provides potent comparisons of
-mortality vectors and converts excess mortality into an intuitively
-understood metric, relatable to everyone. To our knowledge, no studies
-exist that comprehensively examine the potential reductions of life
-expectancy due to climate change. Without properly situating the
-potential loss of life within the contexts of metrics such as life
-expectancy, we risk underestimating the impact of climate change on
-human mortality.
+change and typically use excess or extra mortality (Forzieri et al.
+2017; McMichael, Woodruff, and Hales 2006; Wilson et al. 2017; Zanobetti
+et al. 2012). For example, Forzieri et al. (2017) projected climate
+change extremes to increase European mortality by 150,000+ persons per
+year by the end of the century. Although such estimates are useful, the
+excess mortality estimates are rather sterile -- one death is a tragedy,
+a million deaths a statistic -- and difficult to relate to on a personal
+level. Life expectancy at birth (\(e_0\)), on the other hand, provides
+potent comparisons of mortality vectors and converts excess mortality
+into an intuitively understood metric, relatable to everyone. To our
+knowledge, no studies exist that comprehensively examine the potential
+reductions of life expectancy due to climate change. Without properly
+situating the potential loss of life within the contexts of metrics such
+as life expectancy, we risk underestimating the impact of climate change
+on human mortality.
 
-Using published data on excess mortality \citep{forzieri2017increasing},
-we connect climate change excess mortality to life expectancy in a
+Using published data on excess mortality (Forzieri et al. 2017), we
+connect climate change excess mortality to life expectancy in a
 mortality model. By estimating the increase in age-specific mortality
@@
 approach allows us to quantify the impact of climate change on human
-longevity and answer the following questions: What is the cost of
-inaction on climate change on human longevity? and How does climate
-change related mortality compare to other mortality vectors? Our results
-situate climate change mortality within the broader context of human
-mortality and can be used to inform public health interventions to
-prevent such futures.
+longevity and answer the following question: What is the cost of
+inaction on climate change on human longevity? Our results situate
+climate change mortality within the broader context of human mortality
+and can be used to inform public health interventions to prevent such
+futures.
 
-Our results suggest that climate change could emerge as a potent new
-mortality vector in the 20th century. Mitigation strategies (ie
-reductions in greenhouse gas emisions) and adaptation strategies (ie
-outreach efforts, retrofitting public buildings, etc.) would help
-prevent this new public health concern.
-
-\hypertarget{background}{%
-\section{Background}\label{background}}
-
-\hypertarget{climate-change-and-human-mortality}{%
-\subsection{Climate change and human
-mortality}\label{climate-change-and-human-mortality}}
+Our results suggest that climate change could emerge as a potent public
+health threat in the 20th century. Mitigation strategies (ie reductions
+in greenhouse gas emisions) and adaptation strategies (ie outreach
+efforts, retrofitting public buildings, etc.) would help prevent this
+new public health concern.
 
@@
 
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
 We estimate changes in life expectancy by combining three primary
 datasets: excess mortality data due to meterological hazards related to
-climate change from Forzieri et al. \citep{forzieri2017increasing},
-cause-of-death distributions from the Global Burden of Disease study
-(GBD) \citep{GBD, wang2012age}, and life tables from the Human Mortality
-Database (HMD) \citep{HMD}. Forzieri et al.
-\citep{forzieri2017increasing} focused on the projected excess mortality
-of heatwaves, coldwaves, wildfires, droughts, river and coastal floods,
-and windstorms under the business-as-usual greenhouse gas emission
-scenario (SRES A1B). They modelled long-term, small-area demographic
-changes in European countries that correspond to a middle-of-the-road
-socioeconomic scenario (SSP2 with medium fertility, medium mortality,
-medium migration, and the Global Education Trend education scenario) and
-based their projected mortality on extrapolations of an exhaustive
-examination of contemporary disaster databases within spatially
-rectified, high-resolution gridded datasets. While the disaster data are
+climate change from Forzieri et al. (2017), cause-of-death distributions
+from the Global Burden of Disease study (GBD) (Disease Study 2015 2017;
+Wang et al. 2012), and life tables from the Human Mortality Database
+(HMD) (Database 2017).
+
+Forzieri et al. (2017) combined data from the International Disaster
+Database (EMCAT) and the Natural Catastrophe Statistics tool
+(NatCatSERVICE) to create exposure and fatality statistics related to
+six weather-related hazards -- heatwaves, cold waves, droughts,
+wildfires, river and coastal floods, and windstorms -- and estimated
+excess mortality associated with those hazards during a baseline period
+of 1981-2010 in 28 European countries. They only considered fatalities
+\emph{directly attributable} to the event and \emph{exclude} increased
+deaths from common causes that were observed to rise during the event,
+such as cardiovascular or respiratory deaths\footnote{``The fatality
+  data from the two databases considered is likely to not include
+  increased deaths from common causes that were observed to rise at the
+  population level but for which individual deaths could not be
+  attributed to the event. For example,\ldots{} increased risk of
+  cardiovascular and respiratory deaths \ldots{} may be severly
+  underreported in the EMDAT and NatCatSERVICE.'' (see Forzieri et al.
+  2017:9, \emph{Supplementary Materials}).}. While the disaster data are
 not standardized across country, Forzeri et al.~imputed data in
@@
 spatial variability at the sub-national level, but here we use their
-country-aggregated results. Their results represent business-as-usual
-climate change and human development without incorporating potential
-adaptation. They found a potential 150,000+ climate change related
+country-aggregated results. They downscaled these exposures rates to 1km
+grid cells and integrated them into small-area demographic projections
+for Europe out to 2100 that correspond to a middle-of-the-road
+socioeconomic scenario (SSP2 with medium fertility, medium mortality,
+medium migration, and the Global Education Trend education scenario
+(Jiang 2014; O'Neill et al. 2017; Samir and Lutz 2017)) and based their
+projected mortality on extrapolations of their calculated exposures.
+Their results represent business-as-usual climate change and human
+development without incorporating potential adaptations to climate
+change. They found a potential 150,000+ climate change related
 fatalities per year by the mid 2080s with climate change contributing to
-90\% of mortality rise as opposed to population changes. These data
-provide the anticipated fatalities but not the effect these deaths might
-have on life expectancy.
+90\% of mortality rise as opposed to population changes. The number of
+future anticipated deaths are available in their Supplementary Table S8
+(Forzieri et al. 2017) and provide the magnitude of deaths for our
+study.
 
-To convert excess mortality to life expectancy, we use data gathered
-from the HMD \citep{HMD} for corresponding European countries in
-five-year life tables in conjunction with cause-specific mortality data
-from the GBD \citep{GBD, wang2012age}. The HMD comprises only complete,
-official vital event statistics and represents the most accurate and
-complete compilation of mortality data available. We allocate excess
-multi-hazard mortality for each country from supplementary table S8 in
-Forzieri et al. \citep{forzieri2017increasing} based on the observed
-mortality schedule for environmental heat and cold exposure deaths from
-the GBD for each country in the previous decade.
+\begin{figure}
+\centering
+\includegraphics{MS-cclifeexpec_files/figure-latex/suppfig-1.pdf}
+\caption{\textbf{Mortality distribution of Heat-related mortality in 28
+European countries}. The thick black line is the average\label{suppfig}}
+\end{figure}
+
+The Forzieri et al. (2017) data contain only mortality totals and report
+no age detail, precluding a direct conversion from excess mortality to
+life expectancy. To convert excess mortality to life expectancy, we
+allocate excess multi-hazard mortality for each country based on the
+observed age-specific mortality schedule for environmental heat and cold
+exposure deaths from the GBD (Disease Study 2015 2017; Wang et al.
+2012). Forzieri et al.~found that extreme heat (as opposed to other five
+weather-related hazards) account for 99\% of the anticipated excess
+mortality by the end of the century. The GBD is the most comprehensive
+worldwide dataset of epidemiological data produced and provides
+cause-specific mortality by age/sex/geography/year for 249 causes of
+death in 195 countries and territories. We gathered data on age-specific
+deaths and mortality rates (\(m_x\)) for mortality from environmental
+heat and cold exposure (cause C.2.9) for the period 2006-2015 for each
+country in the study. These data provide a mortality schedule to fit the
+projected climate change excess mortality from Fozieri et al (Forzieri
+et al. 2017). Thus, we assume that the age-specific mortality schedules
+observed between 2006 and 2015 due to environmental heat and cold
+exposure in the GBD (see \autoref{suppfig}) are likely to remain
+unchanged in the future. We only use the GBD to distribute the Forzieri
+et al.~deaths by age.
+
+With the Forzier et al data and the GBD we estimate the anticipated
+age-specific mortality schedules due to climate change in twenty-eight
+European countries but we cannot compare life expectancy differentials
+in the absence of Climate Change using those two datasets alone. To
+create baseline mortality schedules and life expectancy, we gather data
+from the Human Mortality Database (Database 2017) for the corresponding
+European countries. The HMD comprises only complete, official vital
+event statistics and represents the most accurate and complete
+compilation of mortality data available. We downloaded the most recent
+5x1 (age by year) life table data for the twenty-eight European
+countries (see \autoref{Table2} for a list of countries and most recent
+mortality data).
 
@@
 
-We then recalculated age-specific mortality rates in five-year life
-tables by adding the climate change mortality onto the underlying
-mortality data and create four subsequent life tables (\(baseline\),
-\(low\), \(mid\), \(high\)) using standard life table techniques
-\citep{wunsch2013life} based on the published confidence intervals from
-\citep{forzieri2017increasing}. We used a variant of the cause-deleted
-life tables approach
-\citep{brand2005approximations, beltran2008integrated} to measure the
-impact of excess mortality on life expectancy at birth or \(e_0\).
-Changes in life expectancy are reported based on differentials from
-\(e_0\) between the 2080s and the most recent complete life table in the
-HMD (see \textbf{Table 1} for the years used). Complete life tables for
-all countries are available in the \emph{Supplementary Dataset} and all
-computer code to replicate our results are available online\footnote{Replication
-  files are located here:
-  \url{https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1}}.
-
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
-
-To create life expectancy differentials, we used three sources of data.
-The first is the baseline life table data. Baseline life tables comes
-from the Human Mortality Database (HMD) \citep{HMD}. These datasets
-comprise only official vital event registration and are considered the
-most complete and accurate source of mortality data available. We
-downloaded the most recent 5x1 (by age and year) life table data for
-twenty-eight European countries. We used the following data elements
-from the five-year age group life table data in the HMD:
+We used the following data elements from the five-year age group life
+table data in the HMD:
 
@@
 \end{enumerate}
-
-The second dataset comes from published projections of climate-related
-mortality in Europe \citep{forzieri2017increasing}. Forzieri et al.
-\citep{forzieri2017increasing} combined data from the International
-Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool
-(NatCatSERVICE) to create exposure and fatality statistics related to
-six weather-related hazards -- heatwaves, cold waves, droughts,
-wildfires, river and coastal floods, and windstorms. They downscaled
-these exposures to 1km grid scales and integrated them into small-area
-demographic projections for Europe out to 2100. Heatwaves and coldwaves
-account for the vast majority of projected fatalities. The number of
-future anticipated deaths are available in their Supplementary Table S8
-\citep{forzieri2017increasing} and provide the magnitude of deaths for
-our study.
-
-The third dataset comes from the GBD \citep{GBD, wang2012age}. The GBD
-is the most comprehensive worldwide dataset of epidemiological data
-produced and provides cause-specific mortality by age/sex/geography/year
-for 249 causes of death in 195 countries and territories. We gathered
-data on age-specific deaths and mortality rates (\(m_x\)) for mortality
-from environmental heat and cold exposure (cause C.2.9) for the period
-2006-2015 for each country in the study. These data provide a mortality
-schedule to fit the projected climate change excess mortality from
-Fozieri et al \citep{forzieri2017increasing}. Thus, we assume that the
-age-specific mortality schedules observed between 2006 and 2015 due to
-environmental heat and cold exposure in the GBD (see \autoref{suppfig})
-are likely to remain unchanged in the future.
-
-\begin{figure}
-\centering
-\includegraphics{MS-cclifeexpec_files/figure-latex/suppfig-1.pdf}
-\caption{\textbf{Mortality distribution of Heat-related mortality in 28
-European countries}. The thick black line is the average\label{suppfig}}
-\end{figure}
 
@@
 each country \(i\) for each scenario \(s\) (\(BASE\), \(LOW\), \(MID\),
-\(HIGH\)) from Forzieri et al \citep{forzieri2017increasing}.
+\(HIGH\)) from Forzieri et al (Forzieri et al. 2017).
 
@@
 from the HMD, \(\hat{D_{i,s}}\) is the number of deaths from Forzieri et
-al. \citep{forzieri2017increasing} under scenario s, and \(t_{x,i}\) is
-the proportion of mortality experienced in each age group \(x\) from the
+al. (Forzieri et al. 2017) under scenario s, and \(t_{x,i}\) is the
+proportion of mortality experienced in each age group \(x\) from the
 GBD. Thus, the anticipated additional mortality due to climate change is
@@
 \(MID\), and \(HIGH\)) from \(e_{x,i,BASE}\) as described by
-\citep{beltran2008integrated}.
+(Beltrán-Sánchez, Preston, and Canudas-Romo 2008).Complete life tables
+for all countries are available in the \emph{Supplementary Dataset} and
+all computer code to replicate our results are available
+online\footnote{Replication files are located here:
+  \url{https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1}}.
 
@@
 (-2.5 months (-0.1 to -0.39 years)) in the average European country
-(\textbf{\autoref{figure1}}). This reduction is comparable to mortality
-due to influenza and pneumonia \citep{arias2013united} in the United
-States for the year 2000. Although the average European country could
-see a change of -0.23 years, several countries are likely to experience
-considerably greater reductions in life expectancy. Luxembourg could see
-a reduction of up to 2.2 years and the medium-variant of climate hazards
-could cost the average Spaniard 1.03 years of life.
+(\textbf{\autoref{figure1}}) by 2100. This reduction is comparable to
+mortality due to influenza and pneumonia (Arias, Heron, and Tejada-Vera
+2013) in the United States for the year 2000. Although the average
+European country could see a change of -0.23 years, several countries
+are likely to experience considerably greater reductions in life
+expectancy. Luxembourg could see a reduction of up to 2.2 years and the
+medium-variant of climate hazards could cost the average Spaniard 1.03
+years of life.
 
@@
 neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or
-all accident related mortality (-0.84 years) \citep{arias2013united} by
-the end of the century.
+all accident related mortality (-0.84 years) (Arias et al. 2013) by the
+end of the century.
 
@@
 
-\textbf{\autoref{Table1}} reports the rankings of Age Standardised Death
-Rates (ASDR) for the leading causes of death in these European countries
-(ASDR's for other CODs come from Eurostat (online data code:
-hlth\_cd\_asdr2)) with our estimates of climate change included. In many
-European countries, climate change could emerge as a top five killer.
-Only combined CODs, such as all circulatory diseases or all cancers,
-exhibit higher ASDRs if climatic hazards are more severe than
-anticipated for Spain, Luxembourg, and France.
-
-While Forzieri et al's \citep{forzieri2017increasing} results suggest
-that the greatest climate change related mortality will unfold along a
+While Forzieri et al's (Forzieri et al. 2017) results suggest that the
+greatest climate change related mortality will unfold along a
 north-south gradient, we find the greatest reduction in life expectancy
@@
 from climate change. Emergent research in Atmospheric Rivers suggest
-Western Europe could be more susceptible to these types of events
-\citep{ramos2015daily}. Additionally, these results point to the
-importance of converting excess mortality into life expectancy to
-properly quantify the effects of climate change mortality.
+Western Europe could be more susceptible to these types of events (Ramos
+et al. 2015). Additionally, these results point to the importance of
+converting excess mortality into life expectancy to properly quantify
+the effects of climate change mortality.
 
@@
 In this article, we demonstrate the impact climate change could have on
-life expectancy at birth in twenty-eight European countries. Previous
-studies on climate change and excess mortality potentially
-miscommunicate the impact climate change could have on human mortality.
-Contrary to excess mortality estimates, life expectancy is routinely
-used as a primary metric for communicating overall health outcomes and
-enjoys widespread use by major international organizations
-\citep{world2015world, marmot2012building, salomon2012healthy}. Life
-expectancy and its derivatives are the recommended metrics for
-population health \citep{parrish2010peer}. Additionally, it connects
+life expectancy at birth due to extreme weather events in twenty-eight
+European countries. Previous studies on climate change and excess
+mortality potentially miscommunicate the impact climate change could
+have on human mortality. Contrary to excess mortality estimates, life
+expectancy is routinely used as a primary metric for communicating
+overall health outcomes and enjoys widespread use by major international
+organizations (Marmot et al. 2012; Organization 2015; Salomon et al.
+2012). Life expectancy and its derivatives are the recommended metrics
+for population health (Parrish 2010). Additionally, it connects
 mortality estimates into intuitively understandable metrics, translating
-global estimates of mortality into individual outcomes. Our work reveals
-the extent to which climate change could reduce the average person's
-longevity by the end of the century, expanding our understanding of
-climate change and public health; thus linking two of the major areas in
-current developmental and sustainability discussion at both national and
-international levels \citep{abel2016meeting}.
-
-\begin{sidewaystable}
-
-\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1}
-\rowcolors{2}{gray!25}{white}
-\small
-\begin{tabularx}{\textwidth}{l|XXXXXXXXX}
-  \hline
-   \rowcolor{white}
- Country & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\ 
-  \hline
-Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05\\\cellcolor{blue!25}\begin{tiny}(63.26 - 101.6)\end{tiny}} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis.
-of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\
-Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93\\\cellcolor{blue!25}\begin{tiny}(2.36 - 160.9)\end{tiny}} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis.
-of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\
-Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87\\\cellcolor{blue!25}\begin{tiny}(28.15 - 68.2)\end{tiny}} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis.
-of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\
-France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis.
-of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61\\\cellcolor{blue!25}\begin{tiny}(18.24 - 93.5)\end{tiny}} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\
-Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95\\\cellcolor{blue!25}\begin{tiny}(29.93 - 49.4)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis.
-of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\
-Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis.
-of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65\\\cellcolor{blue!25}\begin{tiny}(13.57 - 119.6)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\
-Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02\\\cellcolor{blue!25}\begin{tiny}(11.19 - 37.8)\end{tiny}} & \makecell{Dis.
-of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\
-Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44\\\cellcolor{blue!25}\begin{tiny}(13.32 - 38.6)\end{tiny}} & \makecell{Dis.
-of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\
-Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis.
-of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88\\\cellcolor{blue!25}\begin{tiny}(5.74 - 37.5)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\
-Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis.
-of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23\\\cellcolor{blue!25}\begin{tiny}(0.68 - 37.4)\end{tiny}} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\
-   \hline
-\end{tabularx}
-
-\end{sidewaystable}
+global estimates of mortality into personably relatable outcomes. Our
+work reveals the extent to which climate change could reduce the average
+person's longevity by the end of the century, expanding our
+understanding of climate change and public health; thus linking two of
+the major areas in current developmental and sustainability discussion
+at both national and international levels (Abel et al. 2016).
 
@@
 public health threat for some European countries by the end of the
-century, echoing previous findings
-\citep{forzieri2017increasing, patz2005impact}. Life expectancy has
-steadily risen across the world for the last century
-\citep{gerland2014world} and our results suggest that climate change
-alone could spur a sharp reversal in these trends in some countries. We
-expect two European countries to see life expectancy reductions more
-than one year under the middle scenario (Luxembourg and Spain), but if
-climate change has a greater impact on mortality than anticipated, five
-European countries (Luxembourg, Spain, Italy, France, Ireland) could see
-life expectancy reductions more than one year with Luxembourg
-experiencing a reduction of more than two. These findings highlight the
-hyperlocalized impacts of climate change
-\citep{kendon2014heavier, rosenzweig2010cities, forzieri2017increasing}
+century, echoing previous findings (Forzieri et al. 2017; Patz et al.
+2005). Life expectancy has steadily risen across the world for the last
+century (Gerland et al. 2014) and our results suggest that climate
+change alone could spur a sharp reversal in these trends in some
+countries. We expect two European countries to see life expectancy
+reductions of more than one year under the middle scenario (Luxembourg
+and Spain), but if climate change has a greater impact on mortality than
+anticipated, five European countries (Luxembourg, Spain, Italy, France,
+Ireland) could see life expectancy reductions of more than one year with
+Luxembourg experiencing a reduction of more than two. These findings
+highlight the hyperlocalized impacts of climate change (Forzieri et al.
+2017; Kendon et al. 2014; Rosenzweig et al. 2010).
 
@@
 century and will be in the age groups (aged 65+) most threatened by the
-biggest mortality risk associated with climate change
-\citep{keatinge2000heat} -- extreme heat. If climate change unfolds as a
-more aggressive mortality vector, only all circulatory diseases combined
-or all cancers combined would contribute more to mortality rates in
-numerous European countries. This would make climate change one of the
-most aggressive new mortality vectors to emerge over the last
-quarter-century, representing a major threat to public health in many
-parts of Europe.
+biggest mortality risk associated with climate change (Keatinge et al.
+2000) -- extreme heat. If climate change unfolds as a more aggressive
+mortality vector, only all circulatory diseases combined or all cancers
+combined would contribute more to mortality rates in numerous European
+countries. This would make climate change one of the most aggressive new
+mortality vectors to emerge over the last quarter-century, representing
+a major threat to public health in many parts of Europe.
 
@@
 increase in the use of medications that reduce the ability to perspire
-and sweat, etc. \citep{patz2005impact} The International Classifications
-of Diseases and Related Health Problems (ICD) does not contain ``climate
+and sweat, etc. (Patz et al. 2005) The International Classifications of
+Diseases and Related Health Problems (ICD) does not contain ``climate
 change" as an official cause of death, so we can only speculate that the
@@
 
-We also share the concerns of Lee et al. \citep{lee2017comprehensive}
-concerning the business-as-usual climatic assumptions. It is likely that
-many countries and communities will deploy a wide variety of adaptation
-measures \citep{haines2006climate, kovats2003methods, ebi2006approach}.
-These adaptation measures rely on accurate information about the
+We also share the concerns of Lee et al. (Lee and Kim 2017) concerning
+the business-as-usual climatic assumptions. It is likely that many
+countries and communities will deploy a wide variety of adaptation
+measures (Ebi, Kovats, and Menne 2006; Haines et al. 2006; Kovats et al.
+2003). These adaptation measures rely on accurate information about the
 potential mortality vectors. Our models and those produced by Forzieri
-et al. \citep{forzieri2017increasing} present plausible scenarios on the
+et al. (Forzieri et al. 2017) present plausible scenarios on the
 potential impact of climate change on human mortality and provide
@@
 the broader impact of climate change on human longevity. The disaster
-databases that Forzieri et al \citep{forzieri2017increasing} use in
-generating their excess mortality estimates probably, though the
-disaster databases are unclear, only account for the deaths certified as
-\emph{directly} caused by these hazards and are unlikely to capture the
-overall numbers associated with these deaths. If the certification of
-deaths due to weather extremes is similar to the certication in the
-present, than our results and those of Forzieri et al reflect the exess
-mortality directly attributable to climatic extremes. Despite this
-limitation, the impact of these extremes on \(e_0\) is considerable,
-even if conservative.
+databases that Forzieri et al (Forzieri et al. 2017) use in generating
+their excess mortality estimates probably, though the disaster databases
+are unclear, only account for the deaths certified as \emph{directly}
+caused by these hazards and are unlikely to capture the overall numbers
+associated with these deaths. If the certification of deaths due to
+weather extremes is similar to the certication in the present, than our
+results and those of Forzieri et al reflect the exess mortality directly
+attributable to climatic extremes. Despite this limitation, the impact
+of these extremes on \(e_0\) is considerable, even if conservative.
 
@@
 increasing costs for health care systems in these countries. The health
-care structures are being taxed by population aging
-\citep{rechel2009can} and rising health care costs, yet it remains
-unclear how climate change will exacerbate these pressures.
+care structures are being taxed by population aging (Rechel et al. 2009)
+and rising health care costs, yet it remains unclear how climate change
+will exacerbate these pressures.
 
 \newpage
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-abel2016meeting}{}%
+Abel, Guy J., Bilal Barakat, KC Samir, and Wolfgang Lutz. 2016.
+``Meeting the Sustainable Development Goals Leads to Lower World
+Population Growth.'' \emph{Proceedings of the National Academy of
+Sciences} 113(50):14294--9.
+
+\leavevmode\hypertarget{ref-arias2013united}{}%
+Arias, Elizabeth, Melonie Heron, and Betzaida Tejada-Vera. 2013.
+``United States Life Tables Eliminating Certain Causes of Death,
+1999-2001.'' \emph{National Vital Statistics Reports: From the Centers
+for Disease Control and Prevention, National Center for Health
+Statistics, National Vital Statistics System} 61(9):1--128.
+
+\leavevmode\hypertarget{ref-barnett2007climate}{}%
+Barnett, Jon and W. Neil Adger. 2007. ``Climate Change, Human Security
+and Violent Conflict.'' \emph{Political Geography} 26(6):639--55.
+
+\leavevmode\hypertarget{ref-beltran2008integrated}{}%
+Beltrán-Sánchez, Hiram, Samuel H. Preston, and Vladimir Canudas-Romo.
+2008. ``An Integrated Approach to Cause-of-Death Analysis: Cause-Deleted
+Life Tables and Decompositions of Life Expectancy.'' \emph{Demographic
+Research} 19:1323.
+
+\leavevmode\hypertarget{ref-cloyd2016engagement}{}%
+Cloyd, Emily, Susanne C. Moser, Edward Maibach, Julie Maldonado, and
+Tinqiao Chen. 2016. ``Engagement in the Third Us National Climate
+Assessment: Commitment, Capacity, and Communication for Impact.''
+\emph{Climatic Change} 135(1):39--54.
+
+\leavevmode\hypertarget{ref-HMD}{}%
+Database, Human Mortality. 2017. ``University of California, Berkeley
+(Usa), and Max Planck Institute for Demographic Research (Germany).''
+Available at www.mortality.org or www.humanmortality.de, (datadownloaded
+on 1 August 2017).
+
+\leavevmode\hypertarget{ref-GBD}{}%
+Disease Study 2015, Global Burden of. 2017. ``Lobal Burden of Disease
+Study 2015 (Gbd 2015) Results.'' Seattle, UnitedStates: Institute for
+Health Metrics and Evaluation (IHME), 2016.Available from
+http://ghdx.healthdata.org/gbd--results--tool.For terms and conditions
+of use, pleasevisit
+http://www.healthdata.org/about/terms--and--conditions.
+
+\leavevmode\hypertarget{ref-ebi2006approach}{}%
+Ebi, Kristie L., R. Sari Kovats, and Bettina Menne. 2006. ``An Approach
+for Assessing Human Health Vulnerability and Public Health Interventions
+to Adapt to Climate Change.'' \emph{Environmental Health Perspectives}
+114(12):1930.
+
+\leavevmode\hypertarget{ref-forzieri2017increasing}{}%
+Forzieri, Giovanni, Alessandro Cescatti, Filipe Batista e Silva, and Luc
+Feyen. 2017. ``Increasing Risk over Time of Weather-Related Hazards to
+the European Population: A Data-Driven Prognostic Study.'' \emph{The
+Lancet Planetary Health} 1(5):e200--e208.
+
+\leavevmode\hypertarget{ref-gerland2014world}{}%
+Gerland, Patrick, Adrian E. Raftery, Hana Ševčíková, Nan Li, Danan Gu,
+Thomas Spoorenberg, Leontine Alkema, Bailey K. Fosdick, Jennifer Chunn,
+Nevena Lalic, and others. 2014. ``World Population Stabilization
+Unlikely This Century.'' \emph{Science} 346(6206):234--37.
+
+\leavevmode\hypertarget{ref-haines2006climate}{}%
+Haines, Andy, R. Sari Kovats, Diarmid Campbell-Lendrum, and Carlos
+Corvalán. 2006. ``Climate Change and Human Health: Impacts,
+Vulnerability and Public Health.'' \emph{Public Health} 120(7):585--96.
+
+\leavevmode\hypertarget{ref-hsiang2017estimating}{}%
+Hsiang, Solomon, Robert Kopp, Amir Jina, James Rising, Michael Delgado,
+Shashank Mohan, DJ Rasmussen, Robert Muir-Wood, Paul Wilson, Michael
+Oppenheimer, and others. 2017. ``Estimating Economic Damage from Climate
+Change in the United States.'' \emph{Science} 356(6345):1362--9.
+
+\leavevmode\hypertarget{ref-jiang2014internal}{}%
+Jiang, Leiwen. 2014. ``Internal Consistency of Demographic Assumptions
+in the Shared Socioeconomic Pathways.'' \emph{Population and
+Environment} 35(3):261--85.
+
+\leavevmode\hypertarget{ref-keatinge2000heat}{}%
+Keatinge, WR, GC Donaldson, Elvira Cordioli, M. Martinelli, AE Kunst, JP
+Mackenbach, S. Nayha, and I. Vuori. 2000. ``Heat Related Mortality in
+Warm and Cold Regions of Europe: Observational Study.'' \emph{Bmj}
+321(7262):670--73.
+
+\leavevmode\hypertarget{ref-kendon2014heavier}{}%
+Kendon, Elizabeth J., Nigel M. Roberts, Hayley J. Fowler, Malcolm J.
+Roberts, Steven C. Chan, and Catherine A. Senior. 2014. ``Heavier Summer
+Downpours with Climate Change Revealed by Weather Forecast Resolution
+Model.'' \emph{Nature Climate Change} 4(7):570.
+
+\leavevmode\hypertarget{ref-kovats2003methods}{}%
+Kovats, RS, K. Ebi, Bettina Menne, Diarmid Campbell-Lendrum, OF
+Canziani, A. Githeko, K. Kuhn, D. Le Sueur, P. Martens, AJ McMichael,
+and others. 2003. \emph{Methods of Assessing Human Health Vulnerability
+and Public Health Adaptation to Climate Change}. WHOHealth
+CanadaUNEPWMO.
+
+\leavevmode\hypertarget{ref-lee2017comprehensive}{}%
+Lee, Jae Young and Ho Kim. 2017. ``Comprehensive Assessment of Climate
+Change Risks.'' \emph{The Lancet Planetary Health} 1(5):e166--e167.
+
+\leavevmode\hypertarget{ref-marmot2012building}{}%
+Marmot, Michael, Jessica Allen, Ruth Bell, and Peter Goldblatt. 2012.
+``Building of the Global Movement for Health Equity: From Santiago to
+Rio and Beyond.'' \emph{The Lancet} 379(9811):181--88.
+
+\leavevmode\hypertarget{ref-mcmichael2006climate}{}%
+McMichael, Anthony J., Rosalie E. Woodruff, and Simon Hales. 2006.
+``Climate Change and Human Health: Present and Future Risks.'' \emph{The
+Lancet} 367(9513):859--69.
+
+\leavevmode\hypertarget{ref-melillo2014climate}{}%
+Melillo, Jerry M., TT Richmond, and G. Yohe. 2014. ``Climate Change
+Impacts in the United States.'' \emph{Third National Climate
+Assessment}.
+
+\leavevmode\hypertarget{ref-o2017roads}{}%
+O'Neill, Brian C., Elmar Kriegler, Kristie L. Ebi, Eric Kemp-Benedict,
+Keywan Riahi, Dale S. Rothman, Bas J. van Ruijven, Detlef P. van Vuuren,
+Joern Birkmann, Kasper Kok, and others. 2017. ``The Roads Ahead:
+Narratives for Shared Socioeconomic Pathways Describing World Futures in
+the 21st Century.'' \emph{Global Environmental Change} 42:169--80.
+
+\leavevmode\hypertarget{ref-world2015world}{}%
+Organization, World Health. 2015. \emph{World Health Statistics 2015}.
+World Health Organization.
+
+\leavevmode\hypertarget{ref-pachauri2014climate}{}%
+Pachauri, Rajendra K., Myles R. Allen, Vicente R. Barros, John Broome,
+Wolfgang Cramer, Renate Christ, John A. Church, Leon Clarke, Qin Dahe,
+Purnamita Dasgupta, and others. 2014. \emph{Climate Change 2014:
+Synthesis Report. Contribution of Working Groups I, Ii and Iii to the
+Fifth Assessment Report of the Intergovernmental Panel on Climate
+Change}. IPCC.
+
+\leavevmode\hypertarget{ref-parrish2010peer}{}%
+Parrish, R. Gibson. 2010. ``Peer Reviewed: Measuring Population Health
+Outcomes.'' \emph{Preventing Chronic Disease} 7(4).
+
+\leavevmode\hypertarget{ref-patz2005impact}{}%
+Patz, Jonathan A., Diarmid Campbell-Lendrum, Tracey Holloway, and
+Jonathan A. Foley. 2005. ``Impact of Regional Climate Change on Human
+Health.'' \emph{Nature} 438(7066):310.
+
+\leavevmode\hypertarget{ref-ramos2015daily}{}%
+Ramos, Alexandre M., Ricardo M. Trigo, Margarida LR Liberato, and
+Ricardo Tomé. 2015. ``Daily Precipitation Extreme Events in the Iberian
+Peninsula and Its Association with Atmospheric Rivers.'' \emph{Journal
+of Hydrometeorology} 16(2):579--97.
+
+\leavevmode\hypertarget{ref-rechel2009can}{}%
+Rechel, Berndl, Yvonne Doyle, Emily Grundy, Martin McKee, World Health
+Organization, and others. 2009. ``How Can Health Systems Respond to
+Population Ageing.''
+
+\leavevmode\hypertarget{ref-rigaud2018groundswell}{}%
+Rigaud, Kanta Kumari, Alexander M. De Sherbinin, Bryan Jones, Jonas
+Bergmann, Viviane Clement, Kayly Ober, Jacob Schewe, Susana Beatriz
+Adamo, Brent McCusker, Silke Heuser, and others. 2018. ``Groundswell:
+Preparing for Internal Climate Migration.''
+
+\leavevmode\hypertarget{ref-rosenzweig2010cities}{}%
+Rosenzweig, Cynthia, William Solecki, Stephen A. Hammer, and Shagun
+Mehrotra. 2010. ``Cities Lead the Way in Climate--Change Action.''
+\emph{Nature} 467(7318):909.
+
+\leavevmode\hypertarget{ref-salomon2012healthy}{}%
+Salomon, Joshua A., Haidong Wang, Michael K. Freeman, Theo Vos, Abraham
+D. Flaxman, Alan D. Lopez, and Christopher JL Murray. 2012. ``Healthy
+Life Expectancy for 187 Countries, 1990--2010: A Systematic Analysis for
+the Global Burden Disease Study 2010.'' \emph{The Lancet}
+380(9859):2144--62.
+
+\leavevmode\hypertarget{ref-samir2017human}{}%
+Samir, KC and Wolfgang Lutz. 2017. ``The Human Core of the Shared
+Socioeconomic Pathways: Population Scenarios by Age, Sex and Level of
+Education for All Countries to 2100.'' \emph{Global Environmental
+Change} 42:181--92.
+
+\leavevmode\hypertarget{ref-wang2012age}{}%
+Wang, Haidong, Laura Dwyer-Lindgren, Katherine T. Lofgren, Julie Knoll
+Rajaratnam, Jacob R. Marcus, Alison Levin-Rector, Carly E. Levitz, Alan
+D. Lopez, and Christopher JL Murray. 2012. ``Age-Specific and
+Sex-Specific Mortality in 187 Countries, 1970--2010: A Systematic
+Analysis for the Global Burden of Disease Study 2010.'' \emph{The
+Lancet} 380(9859):2071--94.
+
+\leavevmode\hypertarget{ref-wilson2017climate}{}%
+Wilson, Ander, Brian J. Reich, Christopher G. Nolte, Tanya L. Spero,
+Bryan Hubbell, and Ana G. Rappold. 2017. ``Climate Change Impacts on
+Projections of Excess Mortality at 2030 Using Spatially Varying
+Ozone--Temperature Risk Surfaces.'' \emph{Journal of Exposure Science
+and Environmental Epidemiology} 27(1):118--24.
+
+\leavevmode\hypertarget{ref-zanobetti2012summer}{}%
+Zanobetti, Antonella, Marie S. O'neill, Carina J. Gronlund, and Joel D.
+Schwartz. 2012. ``Summer Temperature Variability and Long-Term Survival
+Among Elderly People with Chronic Disease.'' \emph{Proceedings of the
+National Academy of Sciences} 109(17):6608--13.
 \newpage
 \singlespacing 
-\bibliography{LATEX/ccmort.bib}
 \end{document}

### 2018-10-23 supplementary_rcode.Rmd
minor edits.

@@
 ---
-title: "Climate Change Projected to Reduce European Life Expectancy"
+title: Inaction on climate change projected to cost one year of life in some European
+  countries
 output:
+  html_document:
+    df_print: paged
+  pdf_document: default
   word_document: default
-  pdf_document: default
-date: "April 5, 2018"
+date: "October 19, 2018"
 ---
@@
 
-To successfully run this code, several R pacakgse must be installed and these
+To successfully run this code, several R packages must be installed and these
 packages are located just below this text section but are tidyverse, HMHFDplus,
